### 2015-11-13 projects/project_1/article/article.tex
cleaning up

\documentclass[10pt,a4paper]{amsart}

\usepackage[]{graphicx} 
\usepackage[]{url} 
\usepackage[]{program} 
\usepackage[]{caption} 
\usepackage[]{subcaption} 
\usepackage[]{hyperref} 
\usepackage[]{cleveref}

\usepackage[]{libertine} 
\usepackage[T1]{fontenc}

\title[Solving the Poisson-equation in one dimension]{
  Solving the Poisson-equation in one dimension. \\
  \hrulefill\fbox{FYS3150}\hrulefill
}
\author[Ivar Stangeby]{
  Ivar Stangeby \\
  \href{https://github.com/qTipTip/FYS3150/projects/project_1/}{\texttt{github.com/qTipTip/FYS3150} }
}
\begin{document}

  \begin{titlepage}
\begin{abstract}
  In this project we examine and solve the Poisson equation in one dimension.
  We consider various numerical methods for solving the discretized
  differential equation and compare their numerical efficiency.
\end{abstract}
  \maketitle

  \begin{figure}[h!]
    \centering
    \includegraphics[angle=270, width=0.9\linewidth]{c_h.pdf}
  \end{figure}
  \tableofcontents
  \end{titlepage}

  \section{Introduction}
  \label{sec:introduction}

  In this project we are tasked with solving the one-dimensional Poisson
  equation.  One can look at the project as consisting of three parts.
  Analytical examination of the problem, numerical solution to the problem, and
  lastly, comparing our numerical results to our expectations in the first part.

  We start by deriving the three point approximation to the second
  derivative of a function $f$.  We then use this approximation to show that we
  can rewrite the set of equations as a matrix equation. Our task at hand is
  then to solve this matrix equation using some algorithm, which in this case
  is going to be the tridiagonal matrix algorithm (TDMA). We will analyze this
  algorithm to figure out what kind of results we can expect and we will
  compare this with the standard Gaussian elimination and LU decomposition
  methods.

  We then implement our devised algorithm and solve our system for various
  grid-resolutions and compare our results with the closed form solution of our
  given problem. We will do this visually by plotting the analytical solution
  against the numerical solution.

  Lastly, we want to examine the relative error that occurs for various step
  sizes and extract the largest relative error over the grid for different
  resolutions. The motivation behind this is to figure out which resolution
  (step size) gives the most optimal results.  We then compare this to the
  expectations from part 1.

  \section{Formulating the problem}
  \label{seq:formulating}

  \subsection{Physical background}
  \label{sub:physical_background}

  The poisson-equation in three dimensions read
  \begin{equation}
    \notag
    \nabla^2\Phi = -4\pi\rho(\mathbf r) 
  \end{equation}
  where $\nabla^2$ is the Laplacian-operator. In three dimensions it is
  expressed using spherical coordinates. In this project however, we are going
  to assume that $\Phi$ and $\rho$ are spherically symmetric which essentially
  mean that the whole equation reduces to a one-dimensional problem in the
  radius $r$. Writing out the laplacian for a spherically symmetric system we
  get
  \begin{equation}
    \notag
    \nabla^2 = \frac{1}{r^2}\frac{d}{d2}\left( r^2 \frac{d}{dr} \right).
  \end{equation}
  If we substitute $\Phi(r) = \phi(r)/r$ our poisson-equation then reads
  \begin{equation}
    \notag
    \frac{d^2\phi}{dr^2} = -4\pi r \rho(r).
  \end{equation}
  The motivation behind this substitution is to ensure that we achieve the
  Dirichlet boundary conditions we want.

  We finally let $\phi \rightarrow u$ and $r \rightarrow x$, and the equation
  now reads
  \begin{equation}
    \label{eq:diff_eq}
    -u''(x) = f(x).
  \end{equation}
  The function $f$ is called the \textit{source term}, and in this project we
  are going to assume that $f(x) = 100e^{-10x}$. We can then compare our
  results with the closed form solution $u(x) = 1 - (1-e^{-10})x -e^{-10x}$.
  \Cref{eq:diff_eq} naturally gives rise to the need of a numerical
  approximation to the second derivative of a function $u$.

  \subsection{Approximating the second derivative}
  \label{sub:approximating_the_second_derivative}

  Writing out the Taylor-expansion of a function $f : \mathbb{R} \rightarrow
  \mathbb{R} $ we get
  \begin{equation}
    \notag
    f(x) = \sum^{\infty}_{i=0} \frac{f^{(n)}(a)}{n!}(x - a)^n.
  \end{equation}

  If we examine that partial sums $T_n$ of this series we are truncating the
  exact expansion at some $n$. Note that $f(x) = \lim_{n \rightarrow \infty}
  T_n$.
  \begin{equation}
    \notag
    T_n(x) = \sum^{n}_{i=0} \frac{f^{(n)}(a)}{n!}(x-a)^n.
  \end{equation}
  By doing this, we are essentially approximating our function $f$ at a point
  $x$, and we are introducing a truncation error. We evaluate $T_n$
  around $x = a$ at the point $x \pm h$ and achieve
  \begin{equation}
    \notag
    T_n(x\pm h) = \sum^{n}_{i=0} \frac{f^{(n)}(x)}{n!}h^n.
  \end{equation}

  Setting $n = 2$ we obtain a term containing the second derivative of $f$,
  \begin{equation}
    \notag
    T_2(x \pm h) = f(x) \pm f'(x)h + \frac{f''(x)}{2}h^2,
  \end{equation}
  and we now write out the sum of $T_2(x+h)$ and $T_2(x-h)$. This yields
  \begin{align*}
    \notag
    f(x + h) + f(x-h) &= T_2(x + h) + T_2(x - h) + \mathcal{O}(h^4) \\
    &= 2f(x) + f''(x)h^2 + \mathcal{O}(h^4)
  \end{align*}
  which when solved for $f''(x)$ is equal to
  \begin{equation}
    \label{eq:diff_app}  
    f''(x) = \frac{f(x+h) - 2f(x) + f(x-h)}{h^2}  + \mathcal{O}(h^2).
  \end{equation}

  This equation is the one we will employ during the solution of the linear
  system. Note that $\mathcal{O}(h^2)$ is just shorthand for the truncation
  error introduced by setting $n = 2$. It  is an error-function that runs like
  $h^2$, however we will take the approximation, so ignoring this error-term.
  We will later compare the results for various values of $h$.

  \subsection{System of linear equations}
  \label{sub:system_of_linear_equations}

  We want to solve \cref{eq:diff_eq} over a discrete set of $n$ distinct points
  $\left\{ x_i \right\}_{i=1}^n$ where $x_i = ih$ over the interval $[0, 1]$.
  We define the step length $h = 1 / (n+1)$ and it then follows that $x_0 =
  0$ and $x_{n+1} = 1$.

  If we now write out \cref{eq:diff_eq} using our approximation to the second
  derivative, namely \cref{eq:diff_app}, and denote the discretized
  approximation to $u$ at a point $x_i$ as $v_i$ then we achieve the following
  for the discrete point $x_i$:
  \begin{equation}
    \notag
    -\frac{v_{i+1} - 2v_i + v_{i-1}}{h^2} = f_i \text{ for all } i = 1, \ldots, n,
  \end{equation} 
  where $f(x)$ is the source term. Note that we write $f(x_i) = f_i$.  In
  preparation for the next step we multiply through by $h^2$.
  \begin{equation}
    \label{eq:set_eq}
    -v_{i+1} + 2v_i - v_{i-1} = h^2f_i \text{ for all } i = 1, \ldots, n,
  \end{equation} 
  We now observe that \Cref{eq:set_eq} along with the set of $i$'s define a set
  of $n$ linear equations. Writing them out, we notice a pattern:
  \begin{gather*}
    \label{eq:system}
    -v_2 + 2v_1 - v_0 = h^2f_1 \\
    -v_3 + 2v_2 - v_1 = h^2f_2 \\
                      \vdots \\
    -v_{n} + 2v_{n-1} - v_{n-2} = h^2f_{n-1} \\
    -v_{n+1}+ 2v_n - v_{n-1} = h^2f_n \\
  \end{gather*}
  Remember that we only have $n$ equations and $n+2$ unknowns, hence the
  Dirichlet boundary conditions are key if we are to solve this system.

  This system can be rewritten in terms of a coefficient matrix $A$, a vector
  $\mathbf v$ consisting of the variables $v_i$, and a vector
  $\mathbf{\tilde{b}}$ that consists of $b_i = h^2f_i$. Our system is now
  reduced to a matrix equation
  \begin{equation}
    \label{eq:matrix}
    A \mathbf v = \mathbf{\tilde b}.
  \end{equation}
  Note that $A$ is an $n\times n$ matrix, which does not cover the end cases.
  Also note that $A$ is a tridiagonal matrix, because only the middle three
  diagonals are non-zero.

  \section{Solving the problem}
  \label{sec:solving_the_problem}

  In this section we examine the implementation specific details of this
  project.  We are now going to outline the algorithm known as the
  \textit{Thomas Algorithm}, or \textit{Tridiagonal Matrix
  Algorithm}.\cite{thomasalgo} The Tridiagonal Matrix Algorithm is a special case of LU-decomposition tailored to tridiagonal systems. Due to problems with getting the armadillo \textit{solve()} method to work, I will also implement the standard Gaussian algorithm for comparison.

  \subsection{Devising an algorithm}
  \label{sub:devising_an_algorithm}

  In order to solve this matrix equation involving a tridiagonal matrix there
  are several shortcuts we can take. First note that the matrix is sparse,
  i.e., that the number of non-zero elements scale as $\mathcal{O}(n)$. In our
  case, since the elements are all along the three main diagonals we can make
  significant cuts in the memory by only storing the three diagonals as
  vectors. This way we only have to store $3n$ of the elements, assuming we
  only work with vectors of length $n$.

  If we let $\mathbf a$, $\mathbf b$, $\mathbf c$ denote the three main
  diagonals with elements $a_i, b_i, c_i$ for $i = 1, \ldots, n$ respectively
  then \cref{eq:set_eq} can be written as
  \begin{equation}
    \notag
    a_iv_{i+1} + b_iv + cv_{i-1} = h^2 f_i = \tilde b_i.
  \end{equation}

  Our goal now is to use the first equation to eliminate $v_1$ from the second
  equation, then use the new modified second equation to eliminate $v_2$ from
  the third equation and so on. After doing this for all $n$ equations, the
  final equation is an equation in one variable, namely $v_{n}$, which can be
  used to solve the previous $n-1$ equations.  This is an algorithm in two
  steps, a forward sweep where we forward substitute, and then a backward sweep
  where we substitute back in to solve the equations.  This algorithm scales
  linearly in the number of grid points. We require $2n$ substitutions, so the
  algorithm runs in $\mathcal{O}(n)$.

  We now wish to decompose $A$ into two triangular matrices. We denote these by
  $L$ and $U$ respectively.  We can now write \cref{eq:matrix} in terms of
  these new matrices:
  \begin{equation}
    \label{eq:lu}
    L U \mathbf v = \mathbf{\tilde{b}},
  \end{equation}
  with 
  \begin{equation}
    \notag
    A = \begin{bmatrix}
      \lambda_1 & \ldots & \ldots & \ldots & \ldots & \ldots\\
      a_1 & \lambda_2 & \ldots & \ldots & \ldots & \ldots\\
      \ldots & \ldots & \ldots & \ldots & \ldots & \ldots\\ 
      \ldots & \ldots & \ldots & a_{n-1} & \lambda_{n-1} & \ldots\\
      \ldots & \ldots & \ldots & \ldots & a_{n} & \lambda_{n} \\
    \end{bmatrix}\cdot
    \begin{bmatrix}
      1 & \gamma_1 & \ldots & \ldots & \ldots & \ldots\\
      \ldots & 1 & \gamma_2 & \ldots & \ldots & \ldots\\
      \ldots & \ldots & \ldots & \ldots & \ldots & \ldots\\ 
      \ldots & \ldots & \ldots & \ldots & 1 & \gamma_{n-1} \\
      \ldots & \ldots & \ldots & \ldots & \ldots & 1 \\
    \end{bmatrix}
  \end{equation}

  We start by first solving $L\mathbf y = \mathbf{\tilde b}$. This gives us the
  equations
  \begin{align*}
    \label{eq:y}
    y_i = \begin{cases}
      \frac{\tilde b_i}{\lambda_i}, & i = 1 \\
      \frac{\tilde b_i - a_iy_{i-1}}{\lambda_i}, & i = 2, \ldots, n.
    \end{cases}
  \end{align*}
  And solving $U\mathbf v = \mathbf y$ yields:
  \begin{equation}
    \notag
    v_i = \begin{cases}
      y_i - \gamma_iv_{i+1}, & i = 1,\ldots, n-1 \\
      y_i, & i = n.
    \end{cases}
  \end{equation}
  All we need now is an expression for $\lambda_i$ and $\gamma_i$ for $i = 1,
  \ldots, n$. This follows directly from the fact that $A = LU$.
  \begin{align*}
    \lambda_i &= \begin{cases}
      b_i, & i = 1 \\
      b_i - a_i\gamma_{i-1}, & i = 2, \ldots, n.
    \end{cases} \\
    \gamma_i &= \frac{c_i}{\lambda_i}, \quad i = 1, \ldots, n.
  \end{align*}
  We now have everything we need for solving \cref{eq:lu}.

  The algorithm we will implement is given by \cref{fig:tdma}. It is interesting to note
  that when implementing this, we can use the same vector for $\mathbf v$ and
  $\mathbf y$ as they do not interfere. This saves us some memory.
  \begin{figure}[h]
    \centering
    \caption{Tridiagonal matrix algorithm}
    \begin{program}
      \COMMENT{Forward sweep}
      \lambda_1 = b_1
      \gamma_1 = \frac{c_1}{\lambda_1}
      y_1 = \frac{\tilde{b_1}}{\lambda_1}
      \FOR i = 2, \ldots,n \DO
      \lambda_i = b_i - a_i\gamma_{i-1}
      \gamma_i = \frac{c_i}{\lambda_i}
      y_i = \frac{\tilde{b_i} - a_iy_{i-1}}{\lambda_i}
      \OD
      \COMMENT{Backward sweep}
      v_n = y_n
      \FOR i = n-1, \ldots 1 \DO
        v_i = y_i - \gamma_i v_{i+1}
      \OD
    \end{program}
    \label{fig:tdma}
  \end{figure}

  We are now interested in the number of floating point operations (FLOPS) this
  algorithm constitutes. For the forward sweep, there are 2 FLOPS for the
  initialization, and then 6 for each iteration in the for loop for a total of
  $6(n-1) + 2$ FLOPS. For the backward sweep, we have $2(n-1)$ FLOPS.  For the
  whole algorithm this is then $8n - 6$ FLOPS. For large $n$ the $-6$ is
  negligible and we say that the algorithm spends $8n$ FLOPS. This again
  verifies the complexity of the algorithm as $\mathcal O(n)$.

  If we compare this to the standard Gaussian method which require $2n^3/3$
  FLOPS \cite{gauss}, which constitutes a complexity of $\mathcal O(n^3)$, our algorithm
  runs a whole lot faster for large $n$.

  \clearpage
  \subsection{General solution to LU-decomposed matrix}
  \label{sub:standard_gaussian_elimination}

  Using armadillo's LU-decomposition we achieve an upper and lower triangular matrix. We again wish to solve the equations
  \begin{equation}
    L\mathbf{y} = \tilde{\mathbf{b}} \text{ and } U\mathbf{v} = \mathbf{y}.
  \end{equation}
  The reason we wish to use armadillo's LU-decomposition is we want a more general solution algorithm to our equation in order to compare the run time.
  Using the same procedure as in the previous section we achieve the more general algorithm outlined in
  \cref{f:gauss}.
  \begin{figure}[h]
  \caption{Gaussian elimination}
  \label{f:gauss}
  \centering
  \begin{program}
    \COMMENT{Forward sweep}     
    \FOR i = 1, \ldots, n \DO
      y_i = \tilde{b_i}
      \FOR j = 1,\ldots, i-1 \DO
        y_i = y_i - L_{i,j}y_j
        \OD
      \OD
    \COMMENT{Backward sweep}     
    \FOR i = n, \ldots, 1 \DO
      v_i = y_i
      \FOR j = i + 1,\ldots n \DO
        v_i = v_i - U_{i, j}v_j
        \OD
        v_i = \frac{v_i}{U_{i,i}}
        \OD
  \end{program}
\end{figure}
\subsection{Algorithm performance}
\label{sub:algorithm_performance}
Logging the average run-time for the various values of $n$ gives us an estimate of how the two methods compare, our \emph{Tridiagonal Matrix Algorithm} and the more general
\emph{forward-backward substitution}. The results are given in \cref{T:solver_times}.
\begin{table}[h]
\caption{Elapsed time for various n}
\input{times.tex}
\label{T:solver_times}
\end{table}

We see that the time required for the LU-decomposition is quite significant compared to the Tridiagonal Matrix Algorithm. It is interesting then to see the differences in the numerical approximation. The results are given in figure 3. As we can see, the numerical approximation is ''exact'' for $n = 100$, so it converges very quickly. For our case, the LU-method and the Tridiagonal Matrix Algorithm are identical, hence the only difference in performance is the computation needed to decompose $A$ into lower and upper triangular matrices. Therefore, for all intents and purposes, our algorithm is strictly better than the LU-decomposition.

I could not get Armadillo's \texttt{solve()} method to work. The results would probably have differed a bit more had the solution to the LU-decomposed method not been implemented by hand.

\begin{figure}[htb]
  \label{fig:results}
  \caption{Numerical approximation for various $n$}
  \centering
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=\textwidth]{../plots/tdma_10} 
  \end{subfigure}
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=\textwidth]{../plots/lu_10} 
  \end{subfigure}
  \\
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=\textwidth]{../plots/tdma_100} 
  \end{subfigure}
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=\textwidth]{../plots/lu_100} 
  \end{subfigure}
  \\
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=\textwidth]{../plots/tdma_100} 
  \end{subfigure}
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=\textwidth]{../plots/lu_100} 
  \end{subfigure}
\end{figure}

Of interest is also the relative error in our data set using the TDMA for various $n$. This has been computed and is presented in \cref{f:relerr}.

\begin{figure}[htb]
  \centering
  \caption{Relative error in dataset}
  \includegraphics[width=0.8\linewidth]{../plots/relative_error}
  \label{f:relerr}
\end{figure}

The maximum relative error for each time-step is given in \cref{t:relerr_max}.
We see that at $n = 10^5$ the error start increasing again due to the error in
numerical precision becomes large as $h$ becomes small, as we can see in
\cref{eq:diff_app}.
\begin{table}[h]
\caption{Maximum relative error in data set for various $n$}
\input{relative_error_table.tex}
\label{t:relerr_max}
\end{table}

\section{Conclusion}
\label{sec:conclusion}

Our implementation of the \emph{Tridiagonal Matrix Algorithm} improves vastly
on the standard \emph{Gaussian elimination method} and \emph{LU-decomposition}
if speed is to be considered.  However, our algorithm is just an optimization
of the latter methods for use with tridiagonal matrix problems. The main
difference between our TDMA and the LU-method we used for comparison is that
our TDMA is tailored to this specific problem, while the LU-decomposition
method in Armadillo is for a general case.  Hence the latter method is much
much slower. Other than that, the results are similar and no loss in numerical
precision occurs using the one over the other.

In addition, our TDMA is memorypreserving in the sense that it does not have to store each element of a sparse matrix, but just the diagonal vectors. We could even improve on this further, by observing that our diagonal vectors are constant.

\subsection{Things to consider}
\label{sub:things_to_consider}

During this project I learned the importance of planning my implementation of a
numerical algorithm before starting to implement it. Knowing how to count, from
1 to $n$ is key. I struggled a bit with layout out the algorithm for the
boundary conditions as well as knowing what vector length to use. Having the algorithms
written down before starting to implement, as in Figure 1 and 2 helped a lot
and I`ll keep this in mind for the next projects.

I also spent significant time debugging my \verb Python -implementation when the error was in the writing to file in my \verb C++ -program.

\begin{thebibliography}{10}
  \bibitem{thomasalgo}{\textit{https://en.wikipedia.org/wiki/Tridiagonal\_matrix\_algorithm}}
  \bibitem{candh}{\textit{http://www.gocomics.com/calvinandhobbes/1986/11/26}}
  \bibitem{gauss}{\textit{https://en.wikipedia.org/wiki/Gaussian\_elimination\#Computational\_efficiency}}
  \bibitem{fys3150}{ \textit{Computational Physics - Lecture Notes Fall 2015 - Morten Hjorth-Jensen}}
\end{thebibliography}

\end{document}

### 2015-11-13 projects/project_2/article/article.tex
cleaning up

\documentclass[a4paper, 10pt]{amsart}

\usepackage[]{enumerate}
\usepackage[]{setspace}
\usepackage[]{libertine}
\usepackage[]{minted}
\usepackage[]{algorithm}
\usepackage[]{graphicx}
\usepackage[]{subcaption}

\usepackage[]{microtype}
\usepackage[]{inputenc}
\usepackage[]{fontenc}
\usepackage[]{caption}

\usepackage[]{hyperref}
\usepackage[]{cleveref}
\setdisplayskipstretch{3} % vertical spacing between text and math environment
\setminted{fontsize=\footnotesize}

\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
\newcommand{\party}[1]{\frac{d}{d #1}}
\newcommand{\mline}[1]{\mintinline{cpp}{#1}}
\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{github}}}
\renewcommand{\v}[1]{\mathbf{#1}}

\title[Schr\"odinger equation for two electrons]{Schr\"odinger equation for
  two electrons\\
FYS3150}
\author{Ivar Haugal\o kken Stangeby\\}

\date{\today}

\begin{document}

\begin{abstract}
  In this project we wish to solve the Schr\"odinger equation for two electrons
  in a three-dimensional harmonic oscillator well. In approaching this problem
  we first wish to solve the equation for one electron, and then later expand to
  two electrons. In fact, it turns out that the only term differing in the two
  cases is the potential, hence we can develop an algorithm that runs for any
  given potential. \\
  \noindent
  All source code can be found on my GitHub page: \\

  \centering{\href{https://github.com/qTipTip/FYS3150}{\githublogo \, \color{gray!50}\textit{github.com/qTipTip}}}

\end{abstract}

\maketitle

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\linewidth]{ch.pdf}
  \caption*{\href{http://www.gocomics.com/calvinandhobbes/1988/01/06}{\color{gray!50} \footnotesize\textit{Shamelessly stolen from \cite{ch}}}}
\end{figure}

\tableofcontents

\section{Physical background}\footnote{This first bit is in order for me to try
to understand what I am doing, might not be as relevant for this specific
project.}
\label{sec:physical_background}
In classical mechanics, if we know the position and momentum of a particle the
system is completely determined. The analogue in quantum mechanics is the
\emph{wave-function} \( \mathbf{\Psi}(\mathbf{r}, t)\) which describes the
state of a system. If we know the wave-function we cannot obtain, and need not
obtain any more information about the system. We will later examine the
norm-squared of this wave-function, which represents the probability of finding
a particle at a given position $\v{r}$, or if you please the \emph{probability
density} of our wave function.  Denoted
\begin{equation}
  \notag
  |\Psi(\v{r})|^2 = \mathbb{P}(\v{r}).
\end{equation}
We are however typically interested in the probability of finding a particle in
some neighborhood of $\v{r}$, which is represented by $|\Psi(\v{r})|^2 dx$.  If
we have two wave functions $\Psi_1$ and $\Psi_2$ that represent two different
configurations of the same system, then we also have a set of
\textit{superpositions}, linear combinations of $\Psi_1$ and $\Psi_2$ that
represent some configuration of the system.

We typically require that the wave function is normalized, and we will later
enforce this in our implementation. This essentially means that the
integral\footnote{I have often seen the notation $\int dx |\psi(x)|^2$ used.
Why is this done, isn't this ambiguous?} over all possible positions is equal
to one:
\begin{equation}
  \notag
  \int |\psi(x)|^2 \, dx = 1.
\end{equation}
In other words, this tells us that the probability of finding a particle
somewhere is $100\%$.

\emph{Heisenberg's uncertainty principle} tells us that if we have great
certainty in the position of a particle, then we have uncertainty in the
momentum of a particle.  Any wave function $\Psi$ can be represented as a
superposition of wave functions with a well defined observable quantity, namely
wave-length or momentum. This is essentially what de Broglies` relations tells
us:
\begin{align*}
  \psi(x) &= \int \psi(x_0) \delta(x - x_0) \, dx_0 \quad \text{definite position given by the delta-function, }\\
  \psi(x) &= \frac{1}{\sqrt{2\pi}}\int \tilde{\psi}(k) e^{ikx} \, dk \quad \text{definite wave-length given by plane-waves}.
\end{align*}

\subsection{Deriving the eigenvalue problem}
\label{sub:deriving_the_eigenvalue_problem}

The Schr\"odinger equation is a partial differential equation that tells us how
a physical system evolves with time.  The general \emph{time-independent}
Schr\"odinger equation reads
\begin{equation}
  \label{eq:time_independent_eq}
  E\Psi(\v{r}) = \hat{H}\Psi(\v{r}),
\end{equation}
which we see is an eigenvalue-equation. If we set up the Hamiltonian operator
\( \hat{H} \) for our specific system, in this case one electron, this equation
then reads
\begin{equation}
  \notag
  - \frac{\hbar}{2m}\left( \frac{1}{r^2} \party{r} r^2 \party{r} - \frac{l(l+1)}{r^2}\right) R(r) + V(r)R(r) = ER(r).
\end{equation}
We are interested in the radial part \(R(r)\) of this equation.  A series of
algebraic manipulations converts this to a form we can work with, namely
\begin{equation}
  \notag
  - \frac{d^2}{d\rho^2}u(\rho) + \rho^2 u(\rho) = \lambda u(\rho).
\end{equation}
This equation can be discretized using finite-differences which leads to
\begin{equation}
  \label{eq:discrete_eq}
  -\frac{u_{i+1} - 2u_{i} + u_{i-1}}{h^2} + V_i u_i = \lambda u_i.
\end{equation}
This defines a system of equations which can be written as a single matrix
equation with a tridiagonal coefficient matrix. We can then rewrite
\cref{eq:discrete_eq} as a matrix eigenvalue problem:
\begin{equation}
  \begin{bmatrix}
    d_1 & e_1 & 0 & 0 & \ldots & 0 & 0 \\
    e_1 & d_2 & e_2 & 0 & \ldots & 0 & 0 \\
    0 & e_2 & d_3 & e_3 & 0 & \ldots & 0 \\
    \ldots & \ldots & \ldots & \ldots & \ldots & \ldots & \ldots \\
    0 & \ldots & \ldots & \ldots & \ldots & d_{n-2} & e_{n-1} \\
    0 & \ldots & \ldots & \ldots & \ldots & e_{n-1} & d_{n-1}
  \end{bmatrix}
  \begin{bmatrix}
    u_1 \\
    u_2 \\
    \vdots \\
    u_{n-2} \\
    u_{n-1}
  \end{bmatrix}
  = \lambda
  \begin{bmatrix}
    u_1 \\
    u_2 \\
    \vdots \\
    u_{n-2} \\
    u_{n-1}
  \end{bmatrix}
\end{equation}
We observe that each non-diagonal matrix element \( e_i = -1 / h^2 \) is
constant. This can now be solved using the Jacobi method which will be outlined
in following sections.

\section{Outlining the algorithm}
\label{sec:outlining_the_algorithm}

The \emph{Jacobi eigenvalue algorithm} is an algorithm for calculating the
eigenvalues and eigenvectors of a real symmetric matrix. By performing a series
of similarity transforms the matrix is transformed into an almost diagonal
matrix with the eigenvalues along the diagonal. The eigenvectors are stored
along the way.

\subsection{Mathematical derivation}
\label{sub:mathematical_derivation}

We start by first considering a two-dimensional symmetric matrix
\begin{equation}
  A = \begin{bmatrix}
    \alpha & \gamma \\
    \gamma & \beta
  \end{bmatrix}.
\end{equation}
We can then form the orthogonal matrix $R$ given by
\begin{equation}
  \notag
  R = \begin{bmatrix}
    \cos \theta && -\sin\theta \\
    \sin\theta && \cos\theta
  \end{bmatrix},
\end{equation}
which when applied to two-dimensional vector $\v{x}$ rotates it an angle
$\theta$. Letting $c = \cos\theta$ and $s = \sin\theta$ we can examine the
similarity transform $R^{-1}AR$. This transform diagonalizes $A$ given that the
off diagonals are zero. We see that
\begin{equation}
  \label{eq:givens_two}
  \notag
  R^{-1}AR = \begin{bmatrix}
    \alpha c^2 + 2\gamma cs + \beta s^2 & (\beta - \alpha)cs + \gamma(c^2 - s^2)\, \\
    \, (\beta - \alpha)cs + \gamma(c^2 - s^2) & as^2 - 2\gamma cs + \beta c^2
  \end{bmatrix},
\end{equation}
hence we need to solve $(\beta - \alpha)cs + \gamma(c^2 - s^2) = 0$ which
gives:
\begin{equation}
  \notag
  \tau = \cot 2\theta = \frac{\alpha - \beta}{2\gamma}.
\end{equation}

We can now express all the trigonometric functions in terms of $\tau$. If we let
$t = \tan \theta$, $c = \cos \theta = 1 / \sqrt{t^2 + 1}$ and $s = ct$. This in
terms yields $\cot 2\theta = 1/2 (\cot \theta - \tan \theta) = 1/2 (t^{-1} -
t)$; which gives us the quadratic equation
\begin{equation}
  \notag
  t^2 + 2\tau t - 1 = 0.
\end{equation}
Solving this equation gives us two roots. We wish to chose the \emph{smaller}
of the two roots, in order to ensure maximum numerical stability. It is better
to rotate as small an angle as possible each iteration.

We can now express the eigenvalues of $A$ in terms of $t$. The diagonalized
matrix looks like
\begin{equation}
  \notag
  R^{-1}AR = \begin{bmatrix}
    \alpha + t\gamma & 0 \\
    0 & \beta - t\gamma
  \end{bmatrix}.
\end{equation}
Solving the characteristic equation gives us the eigenvalues. For the
eigenvectors we simply perform the inverse rotation on the identity matrix $I$.

The algorithm outlined above for a $(2 \times 2)$-matrix can very easily be
extended to an $(n\times n)$-matrix case. We introduce the \emph{Givens rotation
  matrix} given by
\begin{equation}
  \label{eq:givens}
  G(i, j, \theta) = \begin{bmatrix}
    1 & \hdots & 0 &\hdots & 0 & \hdots & 0 \\
    \vdots & \ddots & \vdots & {} & \vdots & {} & \vdots \\
    0 & \hdots & c &\hdots & -s & \hdots & 0 \\
    \vdots & {} & \vdots & \ddots & \vdots & {} & \vdots \\
    0 & \hdots & s &\hdots & c & \hdots & 0 \\
    \vdots & {} & \vdots & {} & \vdots & \ddots & \vdots \\
    0 & \hdots & 0 &\hdots & 0 & \hdots & 1 \\
  \end{bmatrix},
\end{equation}
where $c = \cos(\theta)$ and $s = \sin(\theta)$. The linear map this matrix
represents is a rotation in the $(i, j)$-plane of $\theta$ radians.  The reason
for introducing this matrix is that is works in general as the matrix $R$ in
\cref{eq:givens_two} does for two dimensional cases.

Multiplying a matrix $A$ by $G^{-1}(k, l, \theta)$ from the left and by $G(k,
l, \theta)$ from the right and solving for element $a_{kl}$ and $a_{lk}$ equal
to zero yields the following set of equations:
\begin{align}
  \label{eq:difference}
  \begin{split}
  a^{\prime}_{hk} &= a^{\prime}_{kh} = ca_{hk} - sa_{hl} \\
  a^{\prime}_{hl} &= a^{\prime}_{lh} = ca_{hl} - sa_{hk} \\
  a^{\prime}_{kl} &= a^{\prime}_{lk} = (c^2 - s^2)a_{kl} + sc(a_{kk} - a_{ll}) = 0\\
  a^{\prime}_{kk} &= c^2a^{\prime}_{kk} + s^2a_{ll} - 2sca_{kl}\\
  a^{\prime}_{ll} &= s^2a^{\prime}_{kk} + c^2a_{ll} + 2sca_{kl}
  \end{split}
\end{align}

We know that this rotation preserves the eigenvalues, however --- we also wish
to keep track of the corresponding eigenvectors. However, these get changed
under a rotation. A way to remedy this is to do the corresponding rotations to
the identity matrix $I$. The columns of the transformed $I^{\prime}$ will then
correspond to the original eigenvectors of our matrix $A$.

After applying the Jacobi method to a tridiagonal matrix $A$, we are left with
the eigenvalues $\left\{ \lambda_i \right\}_{i=0}^{n}$ along the diagonal of the
transformed matrix $A'$, and we are left with the corresponding eigenvectors as
columns in our matrix $I'$.

When plotting the wave-functions, we want to examine the norm-squared of the
functions, however we require that $\int |\psi(x)|^2 dx = 1$. We therefore
normalize our eigenvectors as following:
\begin{equation}
  \notag
  \v{v}_{\lambda_i} =
  \frac{\v{v}_{\lambda_i}}{\sqrt{\int_{\rho_{\textrm{min}}}^{\rho_{\textrm{max}}}
  |\psi(x)|^2 dx}} \quad \text{ for } i = 1, \ldots, n.
\end{equation}

\subsection{Implementation}
\label{sub:implementation}

We now discuss some of the implementation specific details related to this
project. The implementation of this follow closely the implementation details
outlined in \cite[pages~217--220]{t1}, however tailored to work with the
\mline{armadillo} library.

\subsubsection{Jacobi Method}
\label{ssub:jacobi_method}

In order to perform a complete diagonalization of a matrix $A$ using the Jacobi
method we implement the  auxiliary functions \mline{rotate},
\mline{max_off_diag} and \mline{solve}.

The function \mline{rotate} performs one single rotation in the $(k, l)$-plane
in order to zero out the elements specified. The function \mline{max_off_diag}
returns the absolute value of the maximal off-diagonal element in our matrix
and returns by pointer the position $(k, l)$ of this given element.  Finally,
the function \mline{solve} performs the Jacobi rotation until we reach our max
number of iterations or the maximum off-diagonal element is sufficiently small.

\subsubsection{Unit tests}
\label{ssub:unittest}

In order to make sure that our jacobi-method does what we want it to do I
implemented a set of test-functions using the UnitTesting framework
\emph{Catch}. I wanted to ensure that my Jacobi implementation satisfied some
basic requirements.

\begin{enumerate}[i)]
  \item One rotation needed to diagonalize a $(2\times2)$-matrix
  \item The \mline{max_off_diag} method returned the correct indices as well as
    the magnitude of correct value
  \item Repeated iteration of the Jacobi method preserves orthogonality.
  \item Application to a physical system: Solving the Schr\"odinger equation for one electron and
    returning the three lowest eigenvalues exactly within some tolerance.
\end{enumerate}

\subsection{Performance}
\label{sec:performance}

There is no way of telling a-priori how many rotations the algorithm requires
in order to finish.  This is due to the fact that off-diagonal elements that
have been zeroed out earlier may get returned to a non-zero value by a
rotation. We therefore allow our program to run until all elements are
\emph{essentially zero} given some tolerance $\varepsilon$, and at most we want our program
to run for $n^3$ iterations where $n$ is our number of time steps.

The method requires an order of $(n^2 - n)/2$ comparisons in a for-loop, with a
quadratic convergence rate.

The Jacobi method --- being a very slowly converging method --- is however well
suited for parallelization due to the fact that one can implement the method
using block matrices. A numerical performance analysis of the \emph{Block
Jacobi Method} for symmetric eigenvalue problems can be found in
\cite{blockjaco}.

\section{Two electrons in a harmonic oscillator well}
\label{sec:electrons_in_a_harmonic_oscillator_well}

We now wish to examine the Schr\"odinger equation given by a two-electron wave
function $u(r_1, r_2)$.  Introducing the \emph{relative coordinate} $\v{r} =
\v{r_1} - \v{r_2}$ and the \emph{center-of-mass coordinate} $\v{R} = 1/2
(\v{r_1} + \v{r_2})$ we can --- using similar manipulations as in
\cref{sec:physical_background} --- rewrite the Schr\"odinger equation as
\begin{equation}
  \notag
  -\frac{d^2}{d\rho^2}\psi(\rho) + \omega_r^2\rho^2\psi(p) + \frac{1}{\rho} = \lambda\psi(\rho).
\end{equation}

The quantity $\omega_r$ reflects the strength of the oscillator potential, so
intuitively the stronger the oscillator potential, the less broad should the
wave-function $\psi$ look like. We see based on our results in
\cref{fig:vary_omega} that this is indeed the case. We see that the
higher $\omega_r$ we chose, the smaller $\rho_{ \textrm{max} }$ we need.

In \cref{fig:varying_frequency} we see that the quantification of

If we plot the eigenstates for the same parameters with and without
coloumb-interaction we see that the coloumb interaction constitutes a
``widening'' of the norm-squared. Ideally I should have had some plots to
display this, but time ran out.
\begin{figure}[h!]
  \centering
  \includegraphics[width=0.6\linewidth]{./plots/schrodinger_200_10_5.pdf}
  \caption{Plotting the lowest lying eigenstate (ground state) of two electrons for various
    oscillator-frequencies $\omega_r$. We assume Coloumb-interaction. We see
    that increasing $\omega_r$ results in a more certain position closer to
    $\rho_{\textrm{min}}$ which is reasonable since $\omega_r$ represents the
    strength of the harmonic oscillator. This essentially means that the
    stronger the potential the smaller distance between the two electrons.
    Here we plot for $n = 200$ and $\rho_{\textrm{max}} = 10$.}
  \label{fig:vary_omega}
\end{figure}
\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{./plots/schrodinger_200_100_001.pdf}
  \end{subfigure}
  ~
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{./plots/schrodinger_200_10_05.pdf}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{./plots/schrodinger_200_7_1.pdf}
  \end{subfigure}
  ~
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{./plots/schrodinger_200_5_5.pdf}
  \end{subfigure}
  \caption{We here consider the three lowest eigenstates for each oscillator
    frequency, with $n=200$ and an appropriate $\rho_{\textrm{max}}$. We see
    that for the various oscillator frequencies the position of the electrons
    are restricted to ``bands'' of various radii from the origin. We also notice
    a small boundary-error in my program.}
  \label{fig:varying_frecquency}
\end{figure}

\clearpage
\section{Conclusion}

In this project we have examined the behaviour of two electrons with
Coloumb-interaction in a harmonic oscillator well. We reformulated the physical
problem as a discrete eigenvalue problem and solved this using the \emph{Jacobi
  Method} for diagonalizing symmetric matrices. We examine our results for
various oscillator frequencies, and the results seemed to coincide with the
predictions I made with the limited knowledge I have for quantum-physics.

A test-suite was implemented in order to verify and test the methods employed,
and I wish that I implemented the test for the non-interacting case earlier as a
test as I for a while had a bug in my potential calculatins.

No performance analysis of the algorithms were done, but I will hopefully later
have some time to compare this to the \emph{Householder's method} and the
routines included in the \mline{armadillo}-library. Since, in our case we were
working with a tridiagonal symmetric matrix, we could have written a specialized
tridiagonal solver instead which would have improved the efficiency
significantly.

\begin{thebibliography}{9}
  \bibitem{t1}
    Morten Hjort-Jensen,
    \emph{Computational Physics - Lecture Notes Fall 2015},
    Department of Physics, University of Oslo,
    August 2015
  \bibitem{t2}
  M. Taut, \emph{Two electrons in an external oscillator potential: Particular
  analytic solutions of a Coloumb correlation problem}, Laboratory of Atomic
  and Solid State Physics, Cornell University, New York, November 1993
  \bibitem{blockjaco}
    Yuusuke Takahashi, Yuusuke Hirota, Yusaku Yamamoto, \emph{Performance of
    the block Jacobi method for the symmetric eigenvalue problem on a modern
  massively parallel computer}, Algoritmy 2012
  \bibitem{ch}
    \emph{http://www.gocomics.com/calvinandhobbes/1988/01/06}, Retrieved
    October 2015
\end{thebibliography}

\end{document}

### 2015-11-13 projects/project_3/article/article.tex
cleaning up

\documentclass[intlimits]{amsart}
\usepackage[]{xcolor} 
\usepackage[]{graphicx} 
\usepackage[]{hyperref}
\usepackage[]{cleveref}

\renewcommand{\v}[1]{\mathbf{#1}}
\newcommand{\intlimits}{\int_{0}^\infty\int_0^\infty\int_0^{\pi}\int_0^\pi\int_0^{2\pi}\int_0^{2\pi}}
\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{github}}}

\setcounter{tocdepth}{1}
\title{Electrons in a helium atom}
\author{Ivar Haugal\o kken Stangeby}
\begin{document}
\begin{abstract}
  In this project we wish to compute the expected value of the ground state
  correlation energy between two electrons in a helium atom. In doing so, we also
  take the opportunity to discuss various methods of numerical integration.
  Namely Gaussian quadrature with both Laguerre and Legendre polynomials; brute
  force Monte Carlo integration, and using importance sampling.\\
  \noindent
  All source code can be found on my GitHub page: \\

  \centering{\href{https://github.com/qTipTip/FYS3150}{\githublogo \, \color{gray!50}\textit{github.com/qTipTip}}}
\end{abstract}

\maketitle

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\linewidth]{ch.pdf}
\end{figure}

\tableofcontents
\section{Introduction}
\label{sec:introduction}
We define the position of electron $i$ as $\v{r}_i = x_i\v{e}_x + y_i\v{e}_y +
z_i\v{e}_z$, and define $r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$.  The \emph{ground
state correlation energy} of two electrons --- and more specifically its
expectation value is given by the integral
\begin{equation}
  \label{eq:cart}
  \langle \frac{1}{\left| \v{r}_1 - \v{r}_2 \right|} \rangle = \int_{-\infty}^{\infty} d\v{r}_1d\v{r}_2 e^{-2\alpha(r_1, r_2)} \frac{1}{\left| \v{r}_1 - \v{r}_2 \right|}.
\end{equation}
In this project we fix $\alpha=2$ which corresponds to the charge of the helium
atom. It can be shown that this specific integral has a closed form solution of
$5\pi^2/16^2$ which we will use to assess the numerical stability and
effectiveness of our numerical methods.

We will later, when we apply the Gauss-Laguerre quadrature, need our integral
rewritten in spherical coordinates. The differentials are then given by
\begin{equation}
  \label{eq:diff}
  d\v{r}_1d\v{r}_2 = r_1^2dr_1r_2^2dr_2\sin\theta_1d\theta_1\sin\theta_2d\theta_2d\phi_1d\phi_2,
\end{equation}
with a scale factor of
\begin{equation}
  \label{eq:scale}
  \gamma(\beta) = \frac{1}{r_{12}} = \left( r_1^2 + r_2^2 - 2r_1r_2\cos\beta \right)^{-1/2}
\end{equation}
where $$\cos\beta = \cos\theta_1\cos\theta_2 +
\sin\theta_1\sin\theta_2\cos(\phi_1 - \phi_2).$$

\section{Gaussian Quadrature}
\label{sec:gaussian_quadrature}

In Gaussian Quadrature, the main idea is to rewrite an integral as a weighted
sum of function evaluations,
\begin{equation}
  \notag
  I = \int_a^b f(x) \, dx = \sum_{i=1}^{N} \omega_i f(x_i)
\end{equation}
however, we can do use various orthogonal polynomials which arise from
different differential equations in the natural sciences. We will make use of
both Legendre and Laguerre polynomials. The only downfall of these polynomials
are that their orthogonality relation is only satisfied on certain intervals,
hence we need to pick our polynomials carefully.

\subsection{Legendre}
\label{sub:legendre_polynomials}

The Legendre polynomials are defined on the interval $[-1, 1]$ with a weight
function $W(x) = 1$.  This would lead us to believe that we can only perform
integrals over the interval $[-1, 1]$, but a simple change of variables lets us
use any arbitrary interval. However, one must note that this can significantly
degrade the numerical accuracy. The Gauss-Legendre quadrature gives rise to the following
relation:
\begin{equation}
  \notag
  I = \int_{-1}^{1} f(x) \, dx = \int_{-1}^{1}W(x)g(x)\, dx = \int_{-1}^{1}g(x)
  \, dx = \sum^{N}_{i=1} \omega_ig(x_i).
\end{equation}
It turns out that the associated integration points $x_i$ are the roots of the
Legendre polynomials. We can rewrite \cref{eq:cart} in the Legendre form using a simple change of variables.

\subsection{Laguerre}
\label{sub:laguerre_polynomials}

The Laguerre polynomials are orthogonal on $[0, \infty)$ and have an associated
weight function $W(x) = x^\alpha e^{-x}$. Again, the orthogonality property gives us the relation
\begin{equation}
  \notag I = \int_0^\infty f(x) \, dx = \int_0^\infty W(x)g(x) \, dx =
  \int_0^\infty x^\alpha e^{-x} g(x) \, dx = \sum^{N}_{i=1} \omega_ig(x_i).
\end{equation}

In order to use the Gauss-Laguerre quadrature, we need to rewrite
\cref{eq:cart} to spherical coordinates. We do this using \cref{eq:diff} and
\cref{eq:scale}. This gives us
\begin{align*}
  I &= \int_{-\infty}^{\infty} d\v{r}_1d\v{r}_2 e^{-4(r_1, r_2)} \frac{1}{\left| \v{r}_1 - \v{r}_2 \right|}.\\
    &= \intlimits r_1^2r_2^2e^{-4r_1} e^{-4r_2} \gamma(\beta)\sin\theta_1\sin\theta_2 dr_1dr_2d\theta_1d\theta_2d\phi_1d\phi_2\\
  \intertext{Substituting $u_i = 4r_i$, $i = 1, 2$ yields}
  &= \frac{1}{1024}\intlimits u_1^2u_2^2e^{-u_1}e^{-u_2} \frac{\sin\theta_1\sin\theta_2}{\sqrt{u_1^2 + u_2^2 - 2u_1r_2\cos\beta}} du_1du_2d\theta_1d\theta_2d\phi_1d\phi_2\\
    &=\frac{1}{1024} \int_{0}^\infty\int_0^\infty\int_0^{\pi}\int_0^\pi\int_0^{2\pi}\int_0^{2\pi}W(u_1)W(u_2)g(u, \theta, \phi)\,du_1du_2d\theta_1d\theta_2d\phi_1d\phi_2.
\end{align*}

Based on these integral limits it can be wise to apply Gauss-Laguerre only for
the first two integrals, and use Gauss-Legendre for the other four.

\section{Monte Carlo Integration}
\label{sec:monte_carlo_integration}

In Monte Carlo integration we use a \emph{non-deterministic} approach to
computing a definite integral. We sample integration points at random. The
naive approach is to just pick numbers at random in the interval $[0, 1]$,
which is what we will do in the brute force Monte Carlo method however, as we
will see, another method is to use so-called \emph{importance-sampling} to find
a probability distribution function (PDF) that closely resembles the function
we wish to integrate. Since these methods are non-deterministic, there is
always an error associated with the random aspect of the computations. We
therefore, in addition to compute the integral itself, compute the standard
deviation to get a grip on the uncertainty in our results. One of the effects
of importance sampling is to reduce the standard deviation.

\subsection{Brute force Monte Carlo}
\label{sub:brute_force_monte_carlo}

The probability distribution function associated with the brute force Monte Carlo method is
the \emph{uniform distribution} given by
\begin{equation}
  \notag
  p(x) = \begin{cases}
    1, & x \in [0, 1] \\
    0, & x \in [0, 1]^c.
  \end{cases}
\end{equation}
Since $p(x)$ satisfy $\int_{-\infty}^\infty p(x) \, dx = 1$ we have that
\begin{align*}
  \notag
  \langle f \rangle = \int_{-\infty}^\infty p(x) f(x) \, dx = \int_0^1 p(x)f(x) \, dx &\approx \sum^{N}_{i=1} f(x_i).
\end{align*}
In order to apply this for more general integration limits $a, b$ we simply
note that by conservation of probability $p(y)dy = dx$, hence standard change
of variables will suffice: $y(x) = a + (b - a)x$.

\subsection{Importance Sampling Monte Carlo}
\label{sub:importance_sampling_monte_carlo}

We now wish to find a probability distribution function $p(y)$ that closely
mimics the behavior of $f(x)$ over our chosen integration interval.
For our integral, it is suitable to use the exponential distribution $p(y) = e^{-y}$, thus
$p(r) = e^{-4r}$. However, in order to apply this, we must have the normalization criteria satisfied, hence
\begin{equation}
  \notag
  p(r) = Ae^{-4r}
\end{equation} 
is a suitable probability distribution function. Note that we
wish to express our integral in spherical coordinates in order to apply this
specific probability distribution.

In order to make this work for arbitrary intervals, we introduce the change of variable
\begin{equation}
  \notag
  y(x) = -\frac{\ln(1 - x)}{2\alpha}
\end{equation}
which will give us values $y \in [0, \infty)$.

\section{Implementation}
\label{sec:implementation}

In this project we employ a lot of \emph{black-box} functions, so there aren't
that many implementation details to talk about. However, I have included some
thoughts on selected methods.

\subsection{Gauss-Legendre}
\label{sub:gauss_legendre}

In order to evaluate the integral using Gaussian Quadrature with Legendre
polynomials, we need to find suitable integration limits. Clearly, we can't
integrate from $-\infty$ to $\infty$ numerically. We observe that the function
$e^{-4r_i}$ is \emph{essentially zero} for $r_i \approx 3$. This can be
verified by the following calculations. Say we want $e^{-4r_i} = 10^{-5}$:
\begin{equation}
  \notag
  e^{-4r_i} = 10^{-5} \Longrightarrow r_i = -\ln(10^{-5}) / 4
\end{equation}
which gives us that $r_i \approx 2.88$. We can therefore chose to neglect any
errors that arise from incorrect integration limits by integrating from $-3$ to
$3$ for instance.

\subsection{Gauss-Laguerre}
\label{sub:gauss_laguerre}

For the integral in spherical coordinates we chose to use Laguerre polynomials
only for the two radial integrals, and leave the integrals with respect to
$\theta$ and $\phi$ for Legendre polynomials which is more suited for such
narrow integration limits.

\section{Results}
\label{sec:results}

\subsection{Gauss-Legendre}
In \cref{tab:legendre} we see that the convergence rate of the Gauss-Legendre
method is quite poor and the numerical method can be seen as quite unstable.
Increasing the number of integration steps does not necessarily translate
directly into a more accurate approximation, as can be seen for the $N=25$ to
$N=30$ jump. We have in these simulations used the integration interval $[-3,
3]$.
\begin{table}[p]
  \centering
  \caption{In this table we present the computed integral, the absolute error
    in the calculations as well as the time spent for $N$ integration steps for the Gauss-Legendre
    quadrature. The time complexity of the integral itself is $\mathcal{O}(N^6)$
which is pretty immense, and can clearly be seen based on the increase in time
elapsed for increasing $N$. }
  \label{tab:legendre}
  \begin{tabular}{cccc}
    $N$ & Result & Absolute error & Time [sec]\\
    \hline
    \hline
    5& 0.2642& 0.0714& 0.0004\\
    10& 0.0719& 0.1207& 0.0267\\
    15& 0.2390& 0.0463& 0.3184\\
    20& 0.1561& 0.0366& 1.7411\\
    25& 0.1958& 0.0030& 6.6549\\
    30& 0.1772& 0.0154& 19.6983\\
    35& 0.1899& 0.0028&55.4347\\
    40&0.1844&0.0083&117.377\\
    \hline
  \end{tabular}
\end{table}

\subsection{Gauss-Laguerre}
\label{sub:gauss_laguerre}
As seen in \cref{tab:laguerre}, converting to spherical coordinates and
applying Gauss-Laguerre for the two integrals with limits $0$ and $\infty$
dramatically increases the numerical approximation. A strong convergence
towards the exact value is observed. A slight increase in elapsed time is seen,
however only $N = 15$ integration steps is required for a result that matches
the best achieved using only Gauss-Legendre.
\begin{table}[p]
  \centering
  \caption{Presented is the computed integral, the absolute error in calculations
  as well as time elapsed for $N$ integration steps. The time complexity of the
  integral itself is again $\mathcal{O}(N^6)$ however the numerical method is
  converging properly as opposed to the Legendre quadrature.}
  \label{tab:laguerre}
  \begin{tabular}{cccc}
    $N$ & Result & Absolute error & Time [sec]\\
    \hline
    \hline
    5& 0.1734& 0.0193& 0.0011\\
    10& 0.1864& 0.0063& 0.0675\\
    15& 0.1897& 0.0030& 0.8190\\
    20& 0.1910& 0.0016& 4.3892\\
    25& 0.1917& 0.0010& 16.8339\\
    30& 0.1921& 0.0006& 50.1823\\
    35& 0.1923& 0.0004& 128.2480\\
    40& 0.1924&0.002&291.6470\\
    \hline
  \end{tabular}
\end{table}

\subsection{Brute Force Monte Carlo}
\label{sub:brute_force_monte_carlo}

In \cref{tab:MC_brute} we see immediately that the Monte Carlo methods require
a very large number of integration steps. We also see that the results converge
quickly, and the standard deviation -- although fluctuating a fair bit -- also
decreases with increasing $N$. The time elapsed for this brute force method
mimics the time elapsed for the brute force Gauss-Legendre method.

\begin{table}[p]
  \centering
  \caption{Using brute force Monte Carlo integration we see that the results
  converge to the exact fairly quickly. The computed standard deviation however
fluctuates a fair bit. None the less, the brute force Monte Carlo method is
have similar performance to the brute force Gauss-Legendre integral.}
  \label{tab:MC_brute}
  \begin{tabular}{ccccc}
    N & Result & Absolute error & Standard Deviation & Time [sec]\\
    \hline\hline
    $10^3$ &0.0235 &0.1691 &0.0089 &0.00007\\
    $10^4$ &0.0473 &0.1453 &0.0200 &0.00073\\
    $10^5$ &0.1732 &0.0195 &0.0497 &0.00632\\
    $10^6$ &0.1759 &0.0168 &0.0159 &0.06715\\
    $10^7$ &0.1829 &0.0097 &0.0080 &0.64801\\
    $10^8$ &0.1981 &0.0053 &0.0030 &6.46146\\
    $10^9$ &0.1938 &0.0010 &0.0010 &64.5177\\
    \hline
  \end{tabular}
\end{table}

\end{document}

### 2015-11-13 projects/project_4/article/appendix_A_2x2Ising.tex
cleaning up

\section{Analytical derivations}
\label{sec:analytical_derivations}

In this section we derive the closed form analytical expressions for the
physical quantities for a $2\times2$ Ising model. The mean energy and specific
heat of a thermodynamical system can be expressed in terms of the partition
function.
\begin{align*}
  \langle E \rangle = - \frac{\partial}{\partial \beta} \ln Z; &&  \langle C_V \rangle = \frac{1}{kT^2}\frac{\partial^2}{\partial\beta^2}\ln Z.
\end{align*}
We recall that the partition function for a $2\times2$ model is given by ${Z = 4\cosh(8\beta J) + 12}$. The mean energy then becomes
\begin{align*}
  \langle E \rangle = -\frac{\partial}{\partial\beta} \ln \left( 4\cosh(8\beta J) + 12 \right) = -8J\frac{\sinh(8\beta J)}{\cosh(8\beta J+3)}, 
\end{align*} 
and the mean specific heat becomes
\begin{align*}
  \langle C_V \rangle  &= -\frac{1}{kT^2}\frac{\partial}{\partial\beta}\left(-8J\frac{\sinh(8\beta J)}{\cosh(8\beta J+3)}\right)\\
                       &= \frac{1}{kT^2}\frac{64J^2}{\cosh(8\beta J)+3} \left( \cosh(8\beta J) - \frac{\sinh^2(8\beta J)}{\cosh(8\beta J)+3} \right).
\end{align*}

The magnetization of a configuration is given by the sum of all spins. In order
to examine the mean magnetization of a system of a given lattice size, we
simply sum over all possible magnetizations multiplied by the probability of
the system being in each such configuration. Mathematically, this translates into
\begin{equation}
  \notag
  \langle \mathcal{M} \rangle = \frac{1}{Z} \sum^{M}_{i=1} M_i e^{-\beta E_i}.
\end{equation}
For a simple $2\times2$ lattice this simply reduces to counting all the
possibilities and this gives us that the mean magnetization is
\begin{equation}
  \notag
  \langle \mathcal{M} \rangle = \frac{1}{Z} \left( -4e^{8\beta J} - 8e^0 + 8e^0 + 4e^{8\beta J} \right) = 0.
\end{equation}
If we were to consider the mean of the absolute magnetization $|\mathcal{M}|$, then this simply becomes
\begin{equation}
  \notag
  \langle |\mathcal{M}| \rangle = \frac{1}{Z} \left( 4e^{8\beta J} + 8e^0 + 8e^0 + 4e^{8\beta J} \right) = \frac{4 + 2e^{8\beta J}}{\cosh(8\beta J)+3}.
\end{equation}

The susceptibility $\chi$ of a thermodynamical system can be calculated if we
know the variance $\sigma_\mathcal{M}^2$ of the magnetization and is given by 
\begin{equation}
  \notag
  \chi = \frac{1}{kT}\sigma_\mathcal{M}^2.
\end{equation}
We first start by computing the variance of the magnetization:
\begin{align*}
  \sigma_\mathcal{M}^2 = \langle \mathcal{M}^2 \rangle - \langle \mathcal{M} \rangle^2 = \frac{32}{Z}\left( e^{8\beta J} + 1 \right) - 0 = \frac{8 \left( e^{8\beta J} + 1 \right)}{\cosh(8\beta J) + 3}.
\end{align*}
The susceptibility then reads
\begin{equation}
  \notag
  \chi = \frac{8 \left( e^{8\beta J} + 1 \right)}{kT(\cosh(8\beta J) + 3)}.
\end{equation}

### 2015-11-13 projects/project_4/article/base.tex
cleaning up

\documentclass{article}

\usepackage[flushmargin, ragged, side]{footmisc}

\usepackage[]{amsmath} 
\usepackage[]{xcolor} 
\usepackage[]{amsthm} 
\usepackage[]{amssymb} 
\usepackage[]{enumerate} 
\usepackage[]{graphicx} 
\usepackage[]{subcaption} 
\usepackage[]{hyperref} 
\usepackage[]{cleveref}

\newcommand{\ua}{\uparrow}
\newcommand{\da}{\downarrow}
\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{github}}}

\title{%
  \textsc{The Ising model in two dimensions}\\
  \small{FYS3150}
  }
\author{Ivar Haugal\o kken Stangeby}
\begin{document}
  \maketitle
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\linewidth]{new_york_skyline.pdf}
    \caption*{A Metropolis skyline as a function of temperature for 100000 MC-cycles.}

    \label{fig:new_york_skyline}
  \end{figure}
  \clearpage

  \begin{abstract}
    In this project we look at the Ising model in two dimensions. There are a
    set of physical quantities related to magnetic systems that are interesting
    to examine. We implement the Metropolis algorithm and compute various
    expectation-values for these quantities. For a $2\times2$ Ising model we
    derive the analytical expressions for these values and compare them with
    the numerically computed ones. We also examine what happen when we reach
    the thermodynamical limit, as the lattice size of the model increases.
    \noindent
    All source code can be found on my GitHub page: \\

    \centering{\href{https://github.com/qTipTip/FYS3150}{\githublogo \, \color{gray!50}\textit{github.com/qTipTip}}}
  \end{abstract}
  \tableofcontents
  % Explain the physical system we want to model
  % Introduce the Ising Model
  \input{ising_model.tex} 
  % - 1D model
  % - 2D model
  %     - Square lattice Ising Model
  %       - interacting vs noninteracting
  %       - with our without external magnetic field
  %       - we wish to lower energy, explain parallel vs anti-parallel spin
  % Introduce the Metropolis Algorithm and why we want to use it
  \input{metropolis_algorithm.tex}
  % implementation
  \input{implementations.tex}
  % simulations
  \input{simulations.tex}
  % Calculations of 2x2 analytical quantities
  \clearpage
  \input{conclusion.tex}
  \appendix
  \input{appendix_A_2x2Ising.tex}
\end{document}

### 2015-11-13 projects/project_4/article/implementations.tex
cleaning up

\section{Implementation}
\label{sec:implementation}

In this section we take a closer look at some of the implementation specific
details in this project. The Ising model was implemented in \textsc{C++} and
the class \texttt{Ising.cpp} contains all the methods needed for simulating the
model.

\subsection{The Ising class}
\label{sub:the_ising_class}
I chose an object-oriented approach to this implementation because I wanted the
possibility of having class instances that represented a physical system at any
given time. By wrapping all the methods in a class I could easily rerun a
similar simulation for a different temperature by re-initializing the system
and keeping any values from the previous temperature.\footnote{If I managed to
pull any of this through by any stretch of the imagination can be discussed.} I
also had a boolean flag for checking whether the system had been previously
thermalized for a previous temperature - in that case, one could skip the
thermalization.

I did not manage to figure out a way of outputting data that would satisfy the
demands of all the text-problems in the project at once, so I did have to do
some monkey-patching for some of them.

The computationally heavy methods are the \texttt{simulate} and
\texttt{metropolis}-methods, where \texttt{simulate} calls \texttt{metropolis}
once for each Monte-Carlo cycle. If we let $L$ denote the grid dimension and
$M$ the number of Monte-Carlo cycles, then each call of the \texttt{metropolis}
method constitutes a complexity of $\mathcal{O}(L^2)$. The \texttt{simulate}
method calls \texttt{metropolis} $M$ times, hence the total complexity for one
simulation for one temperature is $\mathcal{O}(ML^2)$.

\subsection{Main method}
\label{sub:main_method}
The main method is fairly simple, given a temperature range $[T_1, T_2]$ it
assigns a subinterval based on the number of nodes running in parallel to each
node. Each node then proceeds to simulate the system for its assigned
temperature range, and outputs its data the same data-file. The sorting of the
data is done by \texttt{Python}. The parallelization is done using OpenMPI.

\subsection{Errors}
\label{sub:errors}

My program gives correct values for certain systems, and very wonky values for
other systems. I have not been able to determine whether this is an artifact of
running too few simulations, whether my temperature resolution should have been
higher, or whether it is an artifact introduced by the stochastic nature of the
algorithms or (most probable) there is a bug somewhere in my program.

Based on the plots produced, all the weird behavior occurs in the quantities
directly dependent on the magnetization, which might indicate that that is
where the problem lies.

### 2015-11-13 projects/project_4/article/ising_model.tex
cleaning up

\section{The Ising model}

We first start by discussing the Ising Model in its most general form. We later
specialize the model to our specific needs.

\subsection{The general Ising model}
The \emph{Ising Model} is a model in statistical physics that describe phase
transitions in ferromagnetism\footnote{Ferromagnetism is the basic mechanism by
which certain materials form permanent magnets, or are attracted to magnets.
Ferromagnetism is the strongest type of magnetism.}. In this model we represent
magnetic dipole moments of atomic spins that can be in one of two states,
$\uparrow$ or $\downarrow$. The spins are arranged in a \emph{lattice}, which
enables us to relate one spin to its neighbors.

We first introduce some notation. Let $\Lambda$ be a set of lattice positions,
where each element in $\Lambda$ is assigned a set of neighboring sites. The
lattice then forms a graph. We assign for each $k \in \Lambda$ a variable
$\sigma_k$ which takes the value 1 or -1, representing the site's spin,
$\uparrow$ or $\downarrow$, respectively. We can now create a \emph{spin
configuration} $\sigma$ which is a specific configuration of this lattice.  For
two sites $i, j \in \Lambda$ one can talk about the \emph{interaction}
$J_{ij}$. Generally, for each $i \in \Lambda$ one also has an \textit{external
magnetic field} $h_i$ interacting with $i$.

We can talk of the \emph{energy} of a specific configuration, which is given by
the Hamiltonian
\begin{equation}
  \label{eq:general ising model}
  E = H(\sigma) = - \sum^{}_{\langle i k\rangle} J_{ij}\sigma_i\sigma_j - \mu \sum^{}_{j} h_j\sigma_j
\end{equation}
where the notation $\langle i j \rangle$ just means that we sum over adjacent
spins only.

In order to calculate relevant physical quantities for such a system, we need
an appropriate probability that describe the probability of finding the system
in a certain configuration. The \emph{configuration probability} is given by
the Boltzmann distribution
\begin{equation}
  \label{eq:configuration probability}
  P_\beta(\sigma) = \frac{e^{-\beta H(\sigma)}}{Z_\beta}
\end{equation}
where $Z_\beta$ is a normalization constant. We call $Z_\beta$ the
\emph{partition function} for the \emph{canonical ensemble}. It is interesting
to note that when increasing the temperature $T$, the probability
$P_\beta(\sigma)$ of finding the system in a specific configuration $\sigma$
\emph{decreases}. This is essentially due to the fact that when we increase the
temperature the probability of finding states in an unfavorable configuration
become probabilistically feasible in comparison with the most common
configurations.

When modeling a magnetic physical system there are certain physical quantities
that we are interested in. Firstly, the \emph{mean energy} given by
\begin{equation}
  \label{eq:mean energy}
  \langle E \rangle = \sum^{M}_{i=1} E_i P_\beta(\sigma) = \frac{1}{Z} \sum^{M}_{i=1} E_i e^{-\beta E_i}.
\end{equation}
The \emph{mean magnetization} is given by
\begin{equation}
  \label{eq:mean magnetization}
  \langle \mathcal{M} \rangle = \sum^{M}_{i=1} \mathcal{M}_i P_\beta(\sigma) =\frac{1}{Z} \sum^{M}_{i=1} \mathcal{M}_i e^{\beta E_i},
\end{equation}
where $\mathcal{M}_i = \sum^{}_{j \in \Lambda}\sigma_j$ for all configurations
$\sigma$ and $M$ denotes the number of possible configurations.  We are also
interested in the \emph{magnetic susceptibility} $\chi$ which tells us how much
an extensive parameter changes when an intensive parameter increases. It is
given by
\begin{equation}
  \label{eq:magnetic susceptibility}
  \chi = \frac{1}{k_BT}\left( \langle \mathcal{M}^2 \rangle - \langle \mathcal{M} \rangle^2 \right).
\end{equation}
The \emph{specific heat} $C_V$, which tells us how much the energy changes by
increasing the temperature, is given by
\begin{equation}
  \label{eq:specific heat}
  C_V = \frac{1}{k_BT^2} \left( \langle E^2 \rangle - \langle E \rangle ^2 \right).
\end{equation}

If we can make some assumptions about the interaction $J_{ij}$ we can classify
the Ising model.  If for all $i, j \in \Lambda$ we have
\begin{enumerate}[i)]
  \item $J_{ij} > 0$, we call the interaction \emph{ferromagnetic};
  \item $J_{ij} < 0$, we call the interaction \emph{antiferromagnetic};
  \item $J_{ij} = 0$, the spins are non-interacting.
\end{enumerate}

\subsection{Ising model specialized to our project}

We can make some simplifications to the general Ising model in order to tailor
it to our specific needs. First of all, we wish to examine the system with no
external magnetic field\footnote{This specific system was solved analytically
by Lars Onsager in 1944.}. We can therefore, due to $h_j = 0$ for all $j \in
\Lambda$, rewrite
\cref{eq:general ising model} as
\begin{equation}
  \label{eq:no magnetic ising}
  E = H(\sigma) = - \sum^{}_{\langle i k \rangle} J_{ij}\sigma_i\sigma_j.
\end{equation}
We also assume that the coupling constant $J_{ij}$ that describe the
interaction between neighboring spins as constant $J$ for all $i, j \in
\Lambda$. \Cref{eq:no magnetic ising} can then be read as
\begin{equation}
  \label{eq:constant interaction ising}
  E = H(\sigma) = - J\sum^{}_{\langle i k \rangle} \sigma_i\sigma_j
\end{equation}

In this project we wish to examine a ferromagnetic system, $J > 0$. This in
turn, means that it is favorable for neighboring spins to be parallel, as this
leads to a lower energy. This can be seen from the fact that $\sigma_i \sigma_j
= 1$ whenever spin $i$ and $j$ have the same sign.

\subsection{A $2\times2$ Ising model}
\label{sub:a_2times2_ising_model}

In order to get a feel for what these systems look like we examine a two
dimensional Ising model with lattice dimensions $L = 2$ and periodic boundary
conditions. Our model has $M = 2^4$ different configurations $\sigma$. The
energy for any arbitrary configuration is given by
\begin{align*}
  \label{eq:abitrary energy}
  E_i = -J \sum^{4}\nolimits_{\langle kl \rangle} \sigma_k\sigma_l.
\end{align*}

Based on this expression we see that whenever all spins are parallel we have a
configuration energy of $E_i = -8J$. In the cases where one single spin is
anti-parallel to the other three we have a configuration energy of $E_i = 0$.
If we have two parallel and two anti-parallel, the configuration energy is $E_i
= 0$ or $E_i = 8J$. Each energy state $E_i$ has a \emph{degeneracy}
$\Omega(E_i)$ which corresponds to the number of configurations with the same
energy state. The ground state for this system has degeneracy 2, so we can
either start with all spin up or all spin down and still be in the lowest
energy state possible. For a small $2\times2$ system such as this one, it is
not unlikely that we can transition from a configuration with all spin up to a
configuration with all spin down (although keeping the same energy level),
however this is unlikely for large lattices.

We are now interested in examining the physical quantities of interest
introduced above.  For the $2\times2$ Ising model these quantities have closed
form expressions. The work needed to show these relations are given in Appendix
A. The partition function for this particular system is given by
\begin{equation}
  \notag
  Z = \sum^{16}_{i=1} e^{-\beta E_i} = 4 \cosh (8\beta J) + 12.
\end{equation}
We can now compute the mean energy of the system. This gives
\begin{equation}
  \notag
  \langle E \rangle = 8J \frac{\sinh(8\beta J)}{\cosh (8\beta J)+3}
\end{equation}
Differentiating once more gives us an expression for the specific heat $C_V$
\begin{equation}
  \notag
  C_V = -\frac{64J^2 }{k_BT^2(\cosh(8\beta J) + 3)} \left( \cosh(8\beta J) - \frac{\sinh^2(8\beta J)}{\cosh(8\beta J)+3} \right)
\end{equation}
Similarly, we have a closed form expression for the susceptibility:
\begin{equation}
  \notag
  \chi = \frac{8 \left( e^{8\beta J} + 1 \right)}{\cosh(8\beta J) + 3} \frac{1}{k_BT}.
\end{equation}
We will later refer back to these specific values for the $2\times2$ Ising
model when we compare the analytical results to the numerically computed
values.

### 2015-11-13 projects/project_4/article/metropolis_algorithm.tex
cleaning up

\section{The Metropolis-Hastings Algorithm}

In this project we need random samples from a probability distribution
$P_\beta(\sigma)$ associated with the physical system we wish to model.
However, direct sampling can be difficult due to the fact that we need to
compute the partition function in order to express the probability distribution
in its entirety.

The Metropolis-Hastings algorithm is very well suited for this type of problem
because in order to sample from a specific distribution $P_\beta(\sigma)$ all
we need is a function $f$ proportional to the distribution density.

\subsection{The Metropolis Algorithm}
\label{sub:the_metropolis_algorithm}

Under the assumption that the proposed probability distribution function is
symmetric, we can make some simplifications to the Metropolis-Hastings
algorithm. This is what we call the Metropolis Algorithm.  The general
procedure can be described as follows:

\begin{enumerate}
  \item Initialization: We pick an initial sample.
  \item For every iteration: \\
    \begin{enumerate}
      \item We generate a new candidate for the new sample by picking a random element in the sample. 
      \item We calculate the acceptance ratio, which decides whether to reject
        or accept the new candidate. (This is why the Metropolis algorithm is
        good for this sort of problem. When computing the acceptance ratio, we
        divide out the normalization factor $Z$, which is the computationally
        heavy bit.)
      \item If the acceptance ratio is greater than one, we automatically
        accept the candidate. Otherwise, we accept the candidate with a
        probability equal to the acceptance ratio. In the case of a rejection
        the new sample is set to the previous one.
    \end{enumerate} 
\end{enumerate}

### 2015-11-13 projects/project_4/article/simulations.tex
cleaning up

\section{Simulations}
\label{sec:simulations}
In this section we simulate the Ising model with various properties and examine
the properties of interest as functions of temperature $T$ and as functions of
lattice dimension $L$.
\subsection{$2\times2$ Ising with $T = 1$}
\label{sub:_2x2_ising_with_t_1_}
We assume we have a grid size of $2\times2$ as in
\cref{sub:a_2times2_ising_model}. We wish to consider these for a fixed
temperature $T' = 1.0$ in units of $Tk_B / J$.  We can therefore make the
following algebraic manipulations: Let $E'$ be the energy scaled by the
coupling constant $J$, that is $E' = E / J$. Similarly let $C_V' = C_V / k$ and
$\chi' = \chi J$. Let $N$ denote the number of Monte-Carlo cycles.

We can now easily compute the quantities using the closed form solutions given
in \cref{sub:a_2times2_ising_model}. The results are given in Table
\ref{tab:numerical comparison to analytical 2x2}. If we look at the error in
the numerical approximation, by means of the ratio between numerical and
analytical results we should expect a convergence towards 1. However, for the
specific heat, this is not the case, as the numerical results converge fairly
slowly. In \cref{fig:relative_error_1} we see that the magnetization and energy
converge quickly, after only two thousand cycles, the susceptibility follows
suit after some 10000 cycles while the specific heat never seem to converge,
even after a million cycles.
\begin{table}
  \centering
  \caption{Comparing numerical results to closed form analytical solutions for a $2\times2$ Ising model.}
  \label{tab:numerical comparison to analytical 2x2}
  \begin{tabular}{lcccc}
    \hline
    $N$ & $\langle E \rangle$ & $\langle \mathcal{M} \rangle$ & $C_V$ & $\chi$\\
    \hline
    100 & -2& 1 & 0& 0\\
    1000 & -2& 1& 0& 0 \\
    10000 & -1.9964& 0.67935& 0.0287482& 2.14783 \\
    100000 & -1.99654& 0.0276321& 0.0276321& 3.98085 \\
    1000000 & -1.99611& 0.05832& 0.0310754& 3.97991 \\
    \hline
    Analytical & -1.99598 & 0 & 0.03208 & 3.99330
  \end{tabular}
\end{table}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_b_million.pdf}
  \caption{This plot shows the convergence rate of the four quantities of
  interest. The magnetization and the energy converges after 2000 cycles, while
the susceptibility is slightly slower at 10000 cycles. The specific heat fails
to converge, even for a million cycles.}
  \label{fig:relative_error_1}
\end{figure}
\subsection{$20\times20$ Ising}
\label{sub:_20times20_ising}

We now consider a 20 by 20 lattice. We are interested in examining the number
of Monte-Carlo cycles needed to establish an equilibrium situation where it is
feasible to start calculating the expectation values. We start by plotting the
various expectation values as functions of the number of Monte-Carlo cycles.
We consider both an initial random configuration and an initial ordered
configuration for both the temperature $T = 1$ and $T = 2.4$.

While we're at it, we also keep track of the number of accepted configurations
and plot this as a function of temperature.
We see that for both random and ordered initial conditions, the system with a
temperature of $T = 1$ stabilizes fairly rapidly as can be seen in
\cref{fig:1_mc_ordered} and \cref{fig:1_mc_random} With a temperature of $T =
2.4$ however, the expectation values are a bit jittery at first, and seem to
somehow converge for upwards of 4000 cycles. This can be seen in
\cref{fig:2_mc_ordered} and \cref{fig:2_mc_random}.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{1_mc_ordered.pdf}
  \caption{Plotting the expected energy and expected absolute magnetization as
  functions of the number of Monte-Carlo cycles. This system has a temperature
of $T = 1$ and an ordered initial configuration. In this situation the values
seem to stabilize around $2000$ cycles.}
  \label{fig:1_mc_ordered}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{1_mc_random.pdf}
  \caption{Plotting expected energy and expected absolute magnetization as
  functions of the numbers of the number of Monte-Carlo cycles. System has a
temperature of $T = 1$ with a random initial configuration. The system
stabilizes around $2000$ cycles.}
  \label{fig:1_mc_random}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{2-4_mc_ordered.pdf}
  \caption{Plotting expected energy and expected absolute magnetization as
  functions of the numbers of the number of Monte-Carlo cycles. System has a
temperature of $T = 2.4$ with a random initial configuration. The system
stabilizes around $4000$ cycles.}
  \label{fig:2_mc_ordered}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{2-4_mc_random.pdf}
  \caption{Plotting expected energy and expected absolute magnetization as
  functions of number of Monte-Carlo cycles. System has a temperature of $T =
2.4$ with a random initial configuration. The values seem to stabilize around
4000 cycles.}
  \label{fig:2_mc_random}
\end{figure}

When we consider the number of accepted states as a function of temperature, we
see that it increases drastically as the temperature increases. This is of
course in accordance with what we would assume if we were to just look at the
probability of any given state. As the temperature increases, the probability
of any arbitrary state becomes close to the probability of the most likely
state. You can view this as ''unlocking'' more possible states as the
temperature increases. The results are shown in \cref{fig:accepted_configs}.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{accepted_configs.pdf}
  \caption{Number of accepted configurations as a function of temperature. The
  results are in accordance with the mathematical expression for the
probability of any given state. This number is probably more interesting when
divided by total number of configurations looked at. We then get a graph
converging to some number $t \in \left[ 0, 1 \right]$ which corresponds to the
probability of a configuration being accepted.}
  \label{fig:accepted_configs}
\end{figure}

We also want to look at the probability $P(E)$ for this same system for both
temperatures of $T=1$ and $T = 2.4$. These are given in \cref{fig:probability
1} and \cref{fig:probability 24}. If we compare these results to the computed
energy variance $\sigma_E^2$ which for $T = 1$ is evaluated to $\sigma_E^2
\approx 9.34 $ and for $T = 2.4$ is evaluated to $\sigma_E^2 \approx  3229$ we
see that the plots seem reasonable.\footnote{These are not per spin as the
other quantities discussed.}
For $T = 1$ the standard deviation $\sigma_E \approx 3.06$ and for $T = 2.4$ it
is $\sigma_E \approx 56.8$. The physical interpretation of this is that when
increasing the temperature we accept more possible states as previously
discussed. We therefore see more energy states appearing further from the most
likely state.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{probability_1.pdf}
  \caption{The probabilities of various energy levels in a Ising model with a
    temperature of $T = 1.0$. The energy states with a high probability are
    squashed all the way to the left which seems reasonable, since we start
    with an initially ordered configuration already in its ground state. Under
  low temperatures, the chance of escaping this ground state is very slim.}
  \label{fig:probability 1}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{probability_24.pdf}
  \caption{The probabilities of various energy levels in a Ising model with a
  temperature of $T = 2.4$. Here the energy states of interest are distributed
more evenly with the distribution centered at about $-500J$.}
  \label{fig:probability 24}
\end{figure}

\subsection{The critical temperature $T_C$ and the thermodynamical limit.}

In this section we examine the properties of the Ising model for various grid
sizes, and we fix our attention specifically to the temperature interval
$\left[ 2.0, 2.4 \right]$ as we know that the system undergoes a phase
transition for $T \approx 2.3$.

If we look at the plots given in \cref{fig:task_f_20}, \cref{fig:task_f_40},
\cref{fig:task_f_60} and \cref{fig:task_f_80} we see that increasing the grid
size changes the behavior of the system close to the critical temperature.
Firstly, we see that the susceptibility $\chi$ dramatically increase. As far as
I have understood, this means that near critical temperature, any change in
intensive parameters that does not depend on the number of particles in the
system constitutes a large change in magnetization. We also see that as the
absolute magnetization converges to zero for increasing grid size near critical
temperature which definitely indicates that the system is undergoing a phase
transition. I used python to visualize this phase transition for a
$200\times200$-model. A \texttt{GIF} can be found in the
\texttt{animations}-folder.

Near the critical temperature it turns out we can look at the behavior of certain physical quantities by means of a power law.
For $\beta = 1/8$, $\alpha = 0$ and $\gamma = 7/4$ we see that the mean magnetization, specific heat and susceptibility as functions of temperature $T$ closely follows the following relations:
\begin{align*}
  \langle \mathcal{M}(T) \rangle &\sim \left( T - T_C \right)^\beta \\
  C_V(T) &\sim \left| T_C - T \right|^\alpha \\
  \chi(T) &\sim \left|T_C - T\right|^\gamma
\end{align*}
We can also look at the relation between a quantity computed for a finite
lattice size and the corresponding quantity in the thermodynamical limit. We
now wish to see whether we can approximate the thermodynamical limit using our
simulations of finite grids. The critical temperature as a function of lattice
size satisfies the following relation:
\begin{equation}
  \notag
  T_C(L) - T_C(L = \infty) = aL^{-1/v},
\end{equation}
where for $v = 1$ the exact result for critical temperature is ${kT_C/J = 2 /
\ln(1+\sqrt{2})}$ which evaluates to approximately 2.269. Rearranging the equation gives us
\begin{equation}
  T_C(L = \infty) = T_C(L) - aL^{-1/v}.
\end{equation}

We can now look at our simulations and see if we can approximate the critical
temperature in the thermodynamical limit. My simulations are not sufficiently
good for me to be able to point out the differences in critical temperature for
the various grid sizes, so all I know is that it is located between $T=2.25$
and $T=2.30$. That being said, the only reasonable assumption based on the
results my simulations produced is that $T_C$ is 2.275 for all lattice sizes.
However, this renders us unable to calculate the factor $a$ in the following relations:
\begin{align*}
  T_C(L = \infty) &= T_C(20) - a20^{-1} = 2.275 - \frac{a}{20}\\
  T_C(L = \infty) &= T_C(40) - a40^{-1} = 2.275 - \frac{a}{40}\\
  T_C(L = \infty) &= T_C(60) - a60^{-1} = 2.275 - \frac{a}{60}\\
  T_C(L = \infty) &= T_C(80) - a80^{-1} = 2.275 - \frac{a}{80}
\end{align*}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_f_20.pdf}
  \caption{Physical quantities near the critical temperature $T_C$ for a grid size of 20.}
  \label{fig:task_f_20}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_f_40.pdf}
  \caption{Physical quantities near the critical temperature $T_C$ for a grid size of 40.}
  \label{fig:task_f_40}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_f_60.pdf}
  \caption{Physical quantities near the critical temperature $T_C$ for a grid size of 60.}
  \label{fig:task_f_60}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_f_80.pdf}
  \caption{Physical quantities near the critical temperature $T_C$ for a grid size of 80.}
  \label{fig:task_f_80}
\end{figure}

### 2015-11-13 projects/project_1/src/framework.py

@@ -1,0 +1,153 @@
+""" This python program works as a framework for this project. It handles the
+running of the C++ script for various grid resolutions n.
+It runs both the tdma and the LU-decomposition algorithms for each time step.
+It produces a plot containing the closed form solution of the Poisson equation
+as well as the two numerical methods for each n. The file name will be on the form
+step_size.pdf
+It also extracts the largest relative error in our numerical approximation for
+each n and plots these versus the step length h. File name will be on the form
+relative_error.pdf
+The C++ program spits out file in the format solver_timesteps.dat
+Example: tdma_1000.dat
+It is a three column data file with
+grid_point numerical exact
+"""
+import os
+import numpy as np
+from matplotlib.pyplot import plot, savefig, legend, title, show, close, ylim, xlim, xscale, xlabel, ylabel
+
+def plot_rel_err():
+    """
+    Plots the relative error in the TDMA for various n
+    Generates a table of values in tex format
+    """
+    error_values = []
+    n_values = []
+    with open("../data/relative_error.dat", 'r') as errors:
+        for line in errors:
+            line = line.split()
+            n_values.append(int(line[0]))
+            error_values.append(float(line[1]))
+    n_values = np.array(n_values)
+    error_values = np.array(error_values)
+    plot(n_values, error_values)
+    xscale('log')
+    legend(['relative error'])
+    title('Relative error in data set')
+    savefig('../plots/relative_error.pdf')
+    print 'writing relative error to ../plots/relative_error.pdf'
+
+    with open("../article/relative_error_table.tex", 'w') as out:
+        output = "\\begin{tabular}{lc} \\\\\n"
+        output += "\\hline\n"
+        output += "$n$ & $\\varepsilon_{\\text{max}}$ \\\\\n"
+        output += "\\hline\n"
+        for n, eps in zip(n_values, error_values):
+            output += "%d & %f" % (n, eps)
+            output += "\\\\\n"
+        output += "\\hline\n"
+        output += "\\end{tabular}"
+        out.write(output) 
+
+def print_time(tdma, lu):
+    with open('../data/timer_tdma.dat', 'r') as tdma_times, open('../data/timer_lu.dat') as lu_times:
+        for line in tdma_times:
+            line = line.split()
+            n = int(line[0])
[…]

### 2015-11-13 projects/project_1/src/library.cpp

@@ -1,0 +1,105 @@
+#include "library.h"
+
+#include <chrono>
+#include <cmath>
+#include <iostream>
+
+using namespace arma;
+
+double source_term(double x_i)
+{
+    return 100 * exp(-10 * x_i);
+}
+
+void tdma(vec &a, vec &b, vec &c, vec &v, vec &b_tilde, int n)
+{
+    vec gamma;
+    vec lambda;
+
+    gamma.set_size(n);
+    lambda.set_size(n);
+
+    v[0] = 0;
+    v[n-1] = 0;
+
+    lambda[1] = b[1];
+    gamma[1] = c[1] / lambda[1];
+    v[1] = b_tilde[1] / lambda[1];
+
+    /* Forward sweep */
+    for(int i = 1; i < n-1; ++i)
+    {
+        lambda[i] = b[i] - a[i]*gamma[i-1];
+        gamma[i] = c[i] / lambda[i];
+        v[i] = (b_tilde[i] - a[i]*v[i-1]) / lambda[i];
+    }
+
+    /* Backward sweep */
+
+    for(int i = n-2; i > 0; --i)
+    {
+        v[i] -= gamma[i] * v[i+1];
+    }
+
+    return;
+}
+
+void lu_decomp(vec &a, vec &b, vec &c, vec &v, vec &b_tilde, int n)
+{
+
+    mat A = zeros(n, n);
+    mat L, U, P;
+    vec y = zeros(n);
+
+    // Filling the diagonals of A
+    A.diag(-1) += -1;
+    A.diag(0) += 2;
+    A.diag(1) += -1;
+
+    lu(L, U, P, A);
+
+    // Cannot get the solve() method to work, so
+    // implementing solution by hand.
+    // Forward sweep
+    for (int i = 1; i < n-1; ++i) {
+      y(i) = b_tilde(i);
+      for (int j = 1; j < i; ++j) {
+        y(i) = y(i) - L(i-1, j-1) * y(j);
+      }
+    }
+
+    // Backward sweep
+    for (int i = n-2; i > 0; --i) {
+      v(i) = y(i);
+      for (int j = i+1; j < n-1; ++j) {
+        v(i) = v(i) - U(i-1, j-1)*v(j);
+      }
+      v(i) = v(i) / U(i-1, i-1);
+    }
+}
+
+void data_to_file(vec v, vec v_exact, vec grid_points, std::string file_name, int n)
+{
[…]

### 2015-11-13 projects/project_1/src/library.h

@@ -1,0 +1,11 @@
+# pragma once
+
+#include <armadillo>
+#include <string>
+
+
+double source_term(double);
+void tdma(arma::vec& a, arma::vec& b, arma::vec& c, arma::vec& v, arma::vec& b_tilde, int n);
+void lu_decomp(arma::vec& a, arma::vec& b, arma::vec& c, arma::vec& v, arma::vec& b_tilde, int n);
+void data_to_file(arma::vec v, arma::vec v_exact, arma::vec grid_points, std::string file_name, int n);
+void max_relative_error(arma::vec& v, arma::vec& exact, int n);

### 2015-11-13 projects/project_1/src/main.cpp

@@ -1,0 +1,98 @@
+#include <iostream>
+#include <cmath>
+#include <armadillo>
+#include <string>
+#include <chrono>
+
+#include "library.h"
+
+using namespace arma;
+
+
+int main(int argc, char* argv[])
+{
+    vec a; // Lower diagonal
+    vec b; // Main diagonal
+    vec c; // Upper diagonal
+    vec v; // Unknown variables
+
+    vec v_exact; // The closed form solution
+    vec b_tilde; // The right hand side
+    vec grid_points; // The X_i's
+
+    int n;
+    int solver;
+
+    if(argc > 2){
+        n = atoi(argv[1]);
+        solver = atoi(argv[2]);
+    } else {
+        std::cout << "Usage: " << argv[0] << " n (time steps) solver (1: tdma/2: lu)"<< std::endl;
+        exit(1);
+    }
+
+    double step_length = 1.0 / (n-1);
+
+    std::ostringstream stream;
+    std::string file_name;
+    std::string timer_file_name;
+
+    /* Initialization */
+    a.set_size(n);
+    b.set_size(n);
+    c.set_size(n);
+    v.set_size(n);
+
+    b_tilde.set_size(n);
+    grid_points.set_size(n);
+    v_exact.set_size(n);
+
+    a.fill(-1.0);
+    b.fill(2.0);
+    c.fill(-1.0);
+
+    /* Populating v_exact, b_tilde and grid_points */
+    double x_i;
+    double exp_const = exp(-10);
+    for(int i = 0; i < n; ++i)
+    {
+        x_i = i * step_length;
+        grid_points[i] = x_i;
+        v_exact[i] = 1 - (1 - exp_const) * x_i - exp(-10*x_i);
+        b_tilde[i] = source_term(x_i) * step_length * step_length;
+    }
+
+    switch(solver) {
+        case 1: {
+            stream << "../data/tdma_" << n << ".dat";
+            auto start = std::chrono::high_resolution_clock::now();
+            tdma(a, b, c, v, b_tilde, n);
+            auto end = std::chrono::high_resolution_clock::now();
+            auto total = std::chrono::duration_cast<std::chrono::nanoseconds>(end - start).count();
+
+            std::ofstream file("../data/timer_tdma.dat", ios::app);
+            file << n << " " << total << "\n";
+            file.close();
+            max_relative_error(v, v_exact, n);
+            break;
+                }
+        case 2: {
+            stream << "../data/lu_" << n << ".dat";
[…]

### 2015-11-13 projects/project_2/src/jacobi.cc

@@ -1,0 +1,100 @@
+#include "jacobi.h"
+#include <chrono>
+
+void Jacobi::rotate(arma::mat &A, arma::mat &R, int k, int l, int n) {
+
+  /* Rotates around the (k, l) plane in order to zero out elements.
+   * Also updates the eigenvectors */
+
+  double c;
+  double s;
+  double t;
+  double tau;
+
+  if ( A(k, l) != 0.0 ) {
+    tau = ( A(l, l) - A(k, k) ) / ( 2*A(k, l) );
+    t = ((tau > 0) ? 1.0 / (tau + sqrt(1.0 + tau*tau)) : -1.0 / (-tau + sqrt(1.0 + tau*tau)));
+
+    c = 1.0 / sqrt(1 + t*t);
+    s = c*t;
+  } else {
+    c = 1.0;
+    s = 0.0;
+  }
+
+  double a_kk, a_ll, a_ik, a_il, r_ik, r_il;
+
+  a_kk = A(k, k);
+  a_ll = A(l, l);
+
+  A(k, k) = c*c*a_kk - 2.0*c*s*A(k, l) + s*s*a_ll;
+  A(l, l) = s*s*a_kk + 2.0*c*s*A(k, l) + c*c*a_ll;
+  A(k, l) = 0.0;
+  A(l, k) = 0.0;
+
+  for (int i = 0; i < n; ++i) {
+    if (i != k && i != l) {
+      a_ik = A(i, k);
+      a_il = A(i, l);
+      A(i, k) = c*a_ik - s*a_il;
+      A(k, i) = A(i, k);
+      A(i, l) = c*a_il + s*a_ik;
+      A(l, i) = A(i, l);
+    }
+
+    r_ik = R(i, k);
+    r_il = R(i, l);
+    R(i, k) = c*r_ik - s*r_il;
+    R(i, l) = c*r_il + s*r_ik;
+  }
+
+  return;
+}
+
+double Jacobi::max_off_diag(arma::mat A, int *k, int *l, int n) {
+  /*
+   * returns the absolute value of the greatest element in absolute value
+   * as well as updating the pointers k and l to point to the correct
+   * indices. Assumes the matrix to be symmetric along the diagonal.
+   */
+  double max = 0.0;
+
+  for (int i = 0; i < n; ++i) {
+    for (int j = i + 1; j < n; ++j) {
+      if (fabs(A(i, j)) > max) {
+        max = fabs(A(i, j));
[…]

### 2015-11-13 projects/project_2/src/jacobi.h

@@ -1,0 +1,13 @@
+# pragma once
+
+#include <armadillo>
+
+class Jacobi {
+
+  public:
+
+    void solve(arma::mat &A, arma::mat &R, int n);
+    void rotate(arma::mat &A, arma::mat &R, int k, int l, int n);
+
+    double max_off_diag(arma::mat A, int *k, int *l, int n);
+};

### 2015-11-13 projects/project_2/src/schrodinger.cc

@@ -1,0 +1,81 @@
+#include "jacobi.h"
+#include <armadillo>
+#include <iostream>
+#include <map>
+#include <string>
+
+using namespace arma;
+
+double potential(double omega_r, double rho);
+void write_to_file(int n, mat A, mat R, vec rho, double omega_r, std::string file_name);
+
+int main(int argc, char *argv[])
+{
+
+  if (argc < 3) {
+    std::cout << "Usage: " << argv[0] << " n rho_max omega_r" << std::endl;
+  }
+  int n = atoi(argv[1]);
+
+  double rho_min = 0.0;
+  double rho_max = atof(argv[2]);
+  double omega_r = atof(argv[3]);
+
+  double step_length = (rho_max - rho_min) / (n + 1);
+
+  double off_diag_const = -1.0 / (step_length * step_length);
+  double diag_const = -2 * off_diag_const;
+
+  mat A = zeros(n, n);
+  mat R = zeros(n, n);
+  vec V = zeros(n);
+  vec rho = zeros(n);
+
+  R.eye();
+  A.diag(-1).fill(off_diag_const);
+  A.diag(1).fill(off_diag_const);
+  for (int i = 0; i < n; ++i) {
+    rho(i) = rho_min + (i+1) * step_length;
+    V(i) = potential(omega_r, rho(i));
+    A.diag()(i) = V(i) + diag_const;
+  }
+
+  Jacobi solver;
+  solver.solve(A, R, n);
+
+  std::ostringstream stream;
+  std::string file_name;
+
+  stream << "../data/schrodinger_" << n << "_" << rho_max << "_" << omega_r << ".dat";
+
+  file_name = stream.str();
+
+  write_to_file(n, A, R, rho, omega_r, file_name);
+  return 0;
+}
+
+inline double potential(double omega_r, double rho) {
+  return omega_r*omega_r * rho*rho + (1.0 / rho);
+}
+
+void write_to_file(int n, mat A, mat R, vec rho, double omega_r, std::string file_name) {
+
+  std::map<double, int> lambdamap;
+
+  for (int i = 0; i < n; ++i) {
+    lambdamap[A.diag(0)(i)] = i;
+  }
+
+  vec eigenvalues = sort(A.diag());
+  vec eigenvector_one = R.col(lambdamap[eigenvalues(0)]);
+  vec eigenvector_two = R.col(lambdamap[eigenvalues(1)]);
[…]

### 2015-11-13 projects/project_2/tests/test.cc

@@ -1,0 +1,203 @@
+#define CATCH_CONFIG_MAIN
+
+#include "../../test_framework/catch.hpp"
+#include "../src/jacobi.h"
+
+#include <armadillo>
+
+TEST_CASE( "2x2 matrix rotated once zeroes out off-diagonals", "[arma::mat]" ) {
+
+  Jacobi test_jacobi;
+
+  arma::mat A;
+  arma::mat R;
+
+  A << 1 << 2 << arma::endr
+    << 2 << 1;
+
+  R = arma::zeros(2, 2);
+  R.eye();
+
+  test_jacobi.rotate(A, R, 0, 1, 2);
+
+  REQUIRE ( A(0, 1) == 0 );
+  REQUIRE ( A(1, 0) == 0 );
+  REQUIRE ( A(0, 0) != 0 );
+  REQUIRE ( A(1, 1) != 0 );
+
+}
+
+TEST_CASE ( "max_off_diag returns correct element and indices in a tridiagonal matrix" ) {
+
+  Jacobi test_jacobi;
+
+  arma::mat A;
+
+  A << 1 << -3.7 << arma::endr
+    << -3.7 << 1;
+
+  int n = 2;
+  int k;
+  int l;
+
+  double maxoff = test_jacobi.max_off_diag(A, &k, &l, n);
+
+  REQUIRE( maxoff == 3.7 );
+  REQUIRE( A(k, l) == -3.7 );
+
+}
+
+SCENARIO ( " Jacobi rotation preserves eigenvector orthogonality " ) {
+
+  GIVEN ( " A 4x4 matrix with random elements " ) {
+
+    Jacobi test_jacobi;
+
+    arma::mat A;
+    arma::mat R;
+
+    A << 1 << 2  << 3 << 4 << arma::endr
+      << 2 << 1 << 13 << 23 << arma::endr
+      << 7 << 2 << 11 << 4 << arma::endr
+      << 62 << 1 << 23 << 3;
+
+    R = arma::zeros(4, 4); R.eye();
+
+    WHEN ( "rotated at max-off-diag 100 times" ) {
+      int l;
+      int k;
+
+      for(int i = 0; i < 100; ++i) {
+        test_jacobi.max_off_diag(A, &k, &l, 4);
+        test_jacobi.rotate(A, R, k, l, 4);
+      }
+
+      THEN ( "eigenvalues in R are orthogonal" ) {
+
+        double tolerance = 1.0e-14;
+
[…]

### 2015-11-13 projects/project_2/tools/framework.py

@@ -1,0 +1,91 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import os
+
+class wavefunction:
+    """
+    A class containing the methods needed for
+    reading data and normalizing and plotting the wave-function.
+    Takes file_name of solutions as well as number of steps as input.
+    """
+
+    def __init__(self, file_name, n, rho_max, omega):
+        self.solution_file = file_name
+        self.n = n
+        self.rho_max = rho_max
+        self.omega = omega
+        self.h = rho_max / float(n-1)
+
+    def get_results(self):
+        """
+        extracts three lowest eigenvalues, eigenvectors found in solution_file as well as rho.
+        """
+        self.rho = np.zeros(self.n)
+        self.eigenvalues = np.zeros(self.n)
+        self.eigenvectors = np.zeros((3, self.n))
+        self.eigenvalues = np.zeros(self.n)
+
+        with open(self.solution_file, 'r') as infile:
+            eigenvalues = infile.readline().split()
+            self.eigenvalues[0] = float(eigenvalues[1])
+            self.eigenvalues[1] = float(eigenvalues[2])
+            self.eigenvalues[2] = float(eigenvalues[3])
+
+            for i, line in enumerate(infile):
+                line = line.split()
+                self.rho[i] = float(line[0])
+                self.eigenvectors[0][i] = float(line[1])
+                self.eigenvectors[1][i] = float(line[2])
+                self.eigenvectors[2][i] = float(line[3])
+
+    def normalize(self):
+        """
+        normalizes the wave functions
+        """
+        for eigenvector in self.eigenvectors:
+            eigenvector = eigenvector / np.sqrt(self._trapezoidal(eigenvector**2))
+
+    def _trapezoidal(self, vector):
+        """
+        Computes the area
+        """
+        total = 0
+        for i in range(self.n-1):
+            total += vector[i+1] + vector[i]
+            return (self.h / 2.0) * total
+
+    def plot_n_lowest_states(self, n):
+        """
+        Plots the n lowest eigenstates, in our case we only
+        store the information for the three lowest.
+        Saves the plot based on the string representation of the object
+        """
+        for i in range(n):
[…]

### 2015-11-13 projects/project_2/tools/plotter.py

@@ -1,0 +1,88 @@
+import matplotlib.pyplot as plt
+from numpy import vectorize, array, sqrt
+from numpy.linalg import norm
+import os
+
+def reader(filename):
+    """
+    extracts rho and u(x) given datafile
+    """
+
+    with open(filename, 'r') as infile:
+        rho = [0]
+        u_one = [0]
+        u_two = [0]
+        u_three = [0]
+        for line in infile:
+            line = line.split()
+            rho.append(float(line[0]))
+            u_one.append(float(line[1]))
+            u_two.append(float(line[2]))
+            u_three.append(float(line[3]))
+
+    return array(rho), array(u_one), array(u_two), array(u_three)
+
+def trapezoidal(step_length, u):
+
+    area = 0
+    for i in range(len(u)-1):
+        area += u[i+1] + u[i]
+    return step_length * 0.5 * area
+
+def normalize(step_length, u):
+    """
+    normalizes the vector u by dividing by dividing by the integral of u**2 computed by
+    the trapezoidal rule
+    """
+
+    u /= sqrt(trapezoidal(step_length, u ** 2))
+
+    return u
+
+
+if __name__ == '__main__':
+
+    """
+    Computes the lowest ground state for various omega_r
+    """
+    omega_r = [0.01, 0.5, 1, 5]
+    rho_max = 10
+    n = 200
+    h = rho_max / float(n-1)
+
+    # Running simulations
+    for omega in omega_r:
+        os.system("../build/schrodinger %d %d %f" % (n, rho_max, omega))
+
+    # Reading data and plotting lowest values for varying omega
+    for omega in omega_r:
+        file_name = "../data/schrodinger_%d_%d_" % (n, rho_max) + str(omega) + ".dat"
+        rho, u = reader(file_name)[0:2]
+        u = abs(normalize(h, u)) ** 2
+        plt.plot(rho, u, label='$\omega_r = %.2f$' % omega)
+
+    plt.legend()
+    plt.xlabel('$\\rho$')
+    plt.ylabel('$|\\psi(\\rho)|^2$')
+    plt.savefig('../plots/schrodinger_%d_%d_' % (n, rho_max) + str(omega) + '.pdf')
+
+    """
+    Computing the three lowest eigenstates for various omega_r and appropriate rho_max
+    rho_max_values = [50, 7, 5, 2]
+
[…]

### 2015-11-13 projects/project_3/aux_programs/exp_plotter.py

@@ -1,0 +1,12 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+def exp_alpha(r, alpha):
+    return np.exp(-alpha * r)
+
+x_values = np.linspace(-8, 8, 1000)
+y_values = exp_alpha(x_values, 2)
+
+for i, y in enumerate(y_values):
+    if y < 1.0e-6:
+        print x_values[i]

### 2015-11-13 projects/project_3/aux_programs/legendre_poly.cc

@@ -1,0 +1,27 @@
+#include <iostream>
+// Computes the Legendre polynomial of degree N
+double Legendre( int n, double x) {
+  double r, s, t;
+  int m;
+  
+  r = 0; 
+  s = 1;
+  // Using recurrence relation 
+  for (m = 0; m < n; ++m) {
+    t = r;
+    r = s;
+    s = (2 * m + 1)*x*r - m*t;
+    s = s / (m+1);
+  }
+  return s;
+}
+
+int main(int argc, char *argv[])
+{
+  int n = atoi(argv[1]);
+  double x = atof(argv[2]);
+
+  std::cout << "The Legendre polynomial of order " << n << " at "
+            << x << " is " << Legendre(n, x) << std::endl;  
+  return 0;
+}

### 2015-11-13 projects/project_3/aux_programs/poly_plotter.py

@@ -1,0 +1,57 @@
+"""
+Plots 5 Laguerrue polynomials
+and 5 Lagrange polynomials
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+@np.vectorize
+def legendre(n, x):
+    """
+    Returns the legendre polynomial of degree n evaluated at x
+    """
+    r = 0.
+    s = 1.
+    t = 0.
+    for i in range(n):
+        t = r
+        r = s
+        s = (2*i + 1)*x*r - i*t
+        s = s / (i + 1)
+    return s
+
+@np.vectorize
+def laguerre(n, x):
+    """
+    Returns the laguerre polynomial of degree n evaluated at x
+    """
+    L_prev = 1
+    L_curr = 1 - x
+    L_new = 0
+    for i in range(n):
+        L_new = ((2*i + 1 - x)*L_curr - i*L_prev) / float(i + 1)
+        L_prev = L_curr 
+        L_curr = L_new
+    return L_curr
+
+x_values = np.linspace(-1, 1, 1000)
+
+for n in range(5):
+    y_values = legendre(n, x_values)
+    plt.plot(x_values, y_values, label='$L_%d$' % n)
+plt.grid('on')
+plt.legend()
+plt.savefig('../plots/legendre_polynomials.pdf')
+plt.clf()
+
+x_values = np.linspace(-5, 15, 1000)
+
+for n in range(5):
+    y_values = laguerre(n, x_values)
+    plt.plot(x_values, y_values, label='$L_%d$' % n)
+plt.ylim((-10, 20))
+plt.grid('on')
+plt.legend()
+plt.savefig('../plots/laguerre_polynomials.pdf')
+

### 2015-11-13 projects/project_3/src/framework.py

@@ -1,0 +1,95 @@
+import os
+from numpy import pi
+def run_gauss_legendre():
+    n_values = range(5, 45, 5)
+    limit_values = [3]
+
+    exact = 5 * pi**2 / 16.0**2
+
+    for n in n_values:
+        for r in limit_values:
+            os.system('./a.out %d 1 %d %d' % (n, -r, r))
+
+def run_gauss_laguerre():
+    n_values = range(5, 45, 5)
+    limit_values = [3]
+
+    exact = 5 * pi**2 / 16.0**2
+
+    for n in n_values:
+        for r in limit_values:
+            os.system('./a.out %d 2 %d %d' % (n, -r, r))
+
+def run_bruteforce_MC():
+    n_values = range(10)
+    for n in n_values:
+        os.system('./a.out %d 3 -3 3' % 10**n)
+
+def read_gauss_legendre():
+    times = {}
+    values = {}
+     
+    with open('../data/gauss_legendre.dat', 'r') as infile:
+        for line in infile:
+            line = line.split()
+            n = int(line[0])
+            t = float(line[2])
+            result = float(line[1])
+            a = int(line[3])
+            b = int(line[4])
+
+            if n in times:
+                times[n] += t
+            else:
+                times[n] = t
+
+            if n in values:
+                values[n].append(result)
+            else:
+                values[n] = [result]
+
+        # Averaging time elapsed and changing to seconds
+        for key in times:
+            times[key] /= 5 * 1.0e9
+
+    return times, values
+
+def dict_to_tex(dictionary, filename):
+    output = '\\begin{tabular}{cc}\n'
+    output += '$n$ & average time elapsed [s]\\\\\n'
+    output += '\\hline\n'
+
+    for key in dictionary:
+        output += '%d & %0.4f \\\\\n' % (key, dictionary[key])
+    output+= '\\end{tabular}\n'
+
+    with open(filename, 'w') as outfile:
+        outfile.write(output)
+
+def matrix_to_tex(dictionary, filename):
+    exact = 5 * pi * pi / 16.0**2
+    output = '\\begin{tabular}{cccccc}\n'
+    for key in dictionary:
+        if key == 30:
+            output += str(key) + "\\\\\n"
+        else:
+            output += str(key) + " & " 
+    output += '\\hline\n'
+    for key in dictionary:
+        for val in dictionary[key][:-1]:
+            output +=  "%0.4f & " % val - exact
[…]

### 2015-11-13 projects/project_3/src/integration_methods.cpp

@@ -1,0 +1,247 @@
+#include <cmath>
+#include <iostream>
+#include <fstream>
+#include <iomanip>
+#include <cmath>
+#include <stdlib.h>
+#include <stdio.h>
+
+#define EPS 3.0e-14
+#define MAXIT 10
+#define ZERO 1.0E-10
+#define pi 3.14159265359
+
+using namespace std;
+
+/* Computes the integrand we want to integrate
+ * Skips the cases where the denominator
+ * is sufficiently small, could cause problems.
+ */
+double cartesian_function(double x1, double y1, double z1, double x2, double y2, double z2) {
+  double tolerance = 1.0e-6;
+  double denominator = sqrt((x2 - x1)*(x2 - x1) + (y2 - y1)*(y2 - y1) + (z2 - z1)*(z2 - z1));
+  if (denominator < tolerance ) {
+    return 0.0;
+  }
+
+  double alpha = 2;
+  double r1 = sqrt(x1*x1 + y1*y1 + z1*z1);
+  double r2 = sqrt(x2*x2 + y2*y2 + z2*z2);
+
+
+  double exp_one = -2 * alpha * r1;
+  double exp_two = -2 * alpha * r2;
+
+  return exp(exp_one + exp_two) / denominator;
+}
+
+/* Computes the integrand for spherical coordinates
+ */
+double spherical_function(double r1, double r2, double theta1, double theta2, double phi1, double phi2)
+{
+  double numerator = sin(theta1)*sin(theta2);
+  double cos_beta = (cos(theta1)*cos(theta2)) + (sin(theta1)*sin(theta2)*cos(phi1-phi2));
+  double discriminant = (r1*r1) + (r2*r2) - (2*r1*r2*cos_beta);
+  double denominator = sqrt(discriminant);
+
+  // Ignoring cases where our denominator is close to zero and
+  // when we have complex solutions
+  double tolerance = 1.0e-6;
+  if (denominator < tolerance || discriminant < 0) {
+    return 0.0;
+  }
+
+  return (numerator/denominator);
+}
+void legendre(double x1, double x2, double x[], double w[], int n)
+{
+   int         m,j,i;
+   double      z1,z,xm,xl,pp,p3,p2,p1;
+   double      *x_low, *x_high, *w_low, *w_high;
+
+   m  = (n + 1)/2;                             // roots are symmetric in the interval
+   xm = 0.5 * (x2 + x1);
[…]

### 2015-11-13 projects/project_3/src/integration_methods.h

@@ -1,0 +1,13 @@
+double cartesian_function(double x1, double y1, double z1, double x2, double y2, double z2);
+double spherical_function(double r1, double r2, double theta1, double theta2, double phi1, double phi2);
+
+void legendre(double x1, double x2, double x[], double w[], int n);
+void laguerre(double x[], double w[], int n, double alph);
+void bruteforce_MC(double lower_lim, double upper_lim, int N);
+void importance_sampl_MC(int N);
+
+double gammln(double xx);
+
+double legendre_integral(double *x, double *w, int N, double (*func)(double, double, double, double, double, double));
+double laguerre_integral(double *xr, double *wr, double *x_theta, double *w_theta, double *x_phi, double *w_phi, int N, double (*func)(double, double, double, double, double, double));
+

### 2015-11-13 projects/project_3/src/main.cpp

@@ -1,0 +1,85 @@
+#include <iostream>
+#include <fstream>
+#include <cmath>
+#include <chrono>
+#include "integration_methods.h"
+#define pi 3.14159265359
+// Function prototypes
+int main(int argc, char *argv[])
+{
+  int n;
+  int method_id;
+  double a;
+  double b;
+
+  // Retrieving parameters
+  if (argc < 5) {
+    std::cout << "Usage: " << argv[0] << " N method "<< std::endl;
+    std::cout << "Parameters\n" << "\t N - integration steps\n \t method - 1/2/3 { Legendre/Laguerre/MonteCarlo }\n" << "a, b - integration limits" << std::endl;
+    exit(1);
+  } else {
+    n = atoi(argv[1]);
+    method_id = atoi(argv[2]);
+    a = atof(argv[3]);
+    b = atof(argv[4]);
+  }
+
+  // Initializing empty arrays for weights and corresponding integration points
+  // as well as integration limits
+  double *x = new double[n];
+  double *w = new double[n];
+  double *x_theta = new double[n];
+  double *w_theta = new double[n];
+  double *x_phi = new double[n];
+  double *w_phi = new double[n];
+
+  switch (method_id) {
+    case 1:
+    {
+      // Using Gauss-Legendre quadrature
+      auto start = std::chrono::high_resolution_clock::now();
+
+      legendre(a, b, x, w, n);
+      double result = legendre_integral(x, w, n, &cartesian_function);
+      double exact = 5 * pi * pi / (16.0*16.0);
+      auto end = std::chrono::high_resolution_clock::now();
+      auto time = std::chrono::duration_cast<std::chrono::nanoseconds>(end - start).count();
+      std::cout << "Gauss Legendre for " << n << " steps: " << result << " absolute error: " << fabs(result - exact) << " time: " << time/1.0e9 << " s" << std::endl;
+      std::ofstream outfile;
+      outfile.open("../data/gauss_legendre.dat", std::ios::app);
+      outfile << n <<  " " << result << " " << time << " " << a << " " << b << "\n";
+      outfile.close();
+      break;
+    }
+    case 2:
+    {
+      auto start = std::chrono::high_resolution_clock::now();
+
+      double exact = 5 * pi * pi / (16.0*16.0);
+      // Using Gauss-Laguerre quadrature
+      laguerre(x, w, n, 2);
+      legendre(0, pi, x_theta, w_theta, n);
[…]

### 2015-11-13 projects/project_4/src/ising.cpp

@@ -1,0 +1,279 @@
+#include "ising.h"
+#include <iomanip>
+#include <time.h>
+
+Ising::Ising(int lattice_dimension, std::string file_name, int intermediate_expectations)
+{
+  this->lattice_dimension = lattice_dimension;
+  this->file_name = file_name;
+  this->file_name_expectations = "expectations.dat";
+  write_intermediate_expectation = intermediate_expectations == 1;
+
+  thermalized = false;
+  number_of_spins = lattice_dimension * lattice_dimension;
+
+  spin_matrix = arma::mat(lattice_dimension, lattice_dimension);
+  delta_energy = arma::vec(17);
+  average_values = arma::vec(5);
+  energy_state_counter = arma::vec(4 * number_of_spins + 1);
+
+  energy = 0;
+  magnetization = 0;
+
+  mean_energy = 0;
+  energy_variance = 0;
+  mean_magnetization = 0;
+  specific_heat = 0;
+  susceptibility = 0;
+  mean_absolute_magnetization = 0;
+
+};
+
+/*
+ * Initializes the system with a given temperature. All average values are
+ * reset to zero and the initial energy and magnetization is computed.  Based
+ * on the temperature the initial configuration is either ground state, or
+ * random. We also compute a lookup-table for the delta-energies, in order to
+ * save significant time when simulating.
+ *
+ * If the system is already thermalized, this indicates that we can use the
+ * previous grid configuration.
+ */
+void Ising::initialize_system(double temp)
+{
+  temperature = temp;
+  number_of_accepted_states = 0;
+
+  // energy lookup table
+  for (int i = -8; i <= 8; i++) delta_energy(i+8) = 0;
+  for (int i = -8; i <= 8; i += 4) delta_energy(i+8) = exp(-i/temp);
+ 
+  if (!thermalized) {
+    std::cout << "not thermalized!" << std::endl;
+    // reset averages
+    average_values.zeros();
+    energy_state_counter.zeros();
+    // initial grid_configuration and magnetization
+    energy = 0;
+    magnetization = 0;
+    number_of_accepted_states = 0;
+    for (int i = 0; i < lattice_dimension; i++)
+    {
+      for (int j = 0; j < lattice_dimension; ++j)
+      {
+        // if the temperature is sufficiently low, we might as well
+        // start in a ground state. Otherwise, start in an random configuration;
+        if (temperature < 1.5)
+        {
[…]

### 2015-11-13 projects/project_4/src/ising.h

@@ -1,0 +1,70 @@
+#pragma once
+
+#include <armadillo>
+#include <string>
+
+/* This class contains the methods required to simulate a magnetic system with
+ * a discrete square lattice of arbitrary size for a range of temperatures.
+ */
+class Ising
+{
+  private:
+    // FIELDS
+    // =======================================
+    // properties of the system and simulation
+    int lattice_dimension;
+    int number_of_spins;
+    int number_of_accepted_states;
+    int number_of_cycles;
+    int write_intermediate_expectation;
+
+    bool thermalized;
+
+    double temperature;
+
+    arma::mat spin_matrix;
+    arma::vec average_values;
+    arma::vec delta_energy;
+    arma::vec energy_state_counter;
+
+    std::string file_name;
+    std::string file_name_expectations;
+
+    // physical quantities of interest
+    double mean_energy;
+    double energy_variance;
+    double mean_magnetization;
+    double specific_heat;
+    double susceptibility;
+    double mean_absolute_magnetization;
+
+    // auxilliary variables for computation
+    double energy;
+    double magnetization;
+
+    // METHODS
+    // =================
+    // auxilliary methods
+    void write_probabilities(int cycles_burned);
+    void initialize_energy();
+    void metropolis();
+
+    int periodic(int x, int L, int offset); // returns periodic indices
+    int thermalize(int max_burn_cycles);
+
+    double get_energy(int x, int y);
+
+  public:
+    // METHODS
+    // constructor
+    Ising(int lattice_dimension, std::string file_name="simulation_results.dat", int write_intermediate_expectation=0);
+
+    // initialization and simulation
+    void simulate(int number_of_cycles);
+    void initialize_system(double temperature);
+
+    // results
+    void print_results();
+    void write_to_file();
+    void write_intermediate_expectations(int current_cycle_count);
+};

### 2015-11-13 projects/project_4/src/main.cpp

@@ -1,0 +1,45 @@
+#include "ising.h"
+#include <string>
+#include <mpi.h>
+
+int main(int argc, char *argv[])
+{
+   
+  int world_size; 
+  int world_rank;
+
+  MPI_Init(&argc, &argv);
+  MPI_Comm_size(MPI_COMM_WORLD, &world_size);
+  MPI_Comm_rank(MPI_COMM_WORLD, &world_rank);
+
+   
+  // system as a whole
+  int grid_size = atoi(argv[1]);
+  int mc_cycles = atoi(argv[2]);
+  
+  double temperature = atof(argv[3]);
+  double temperature_end = atof(argv[4]);
+  double temperature_step = atof(argv[5]);
+  std::string file_name = argv[6];
+  
+  double dT = (temperature_end - temperature) / (double) world_size;
+
+  // each node
+  double my_temp = temperature + world_rank * dT;
+  double my_temp_end = my_temp + dT - temperature_step;
+  
+  if (world_rank == world_size - 1) { my_temp_end += temperature_step; } // edge case
+
+  Ising solver = Ising(grid_size, file_name);
+
+  while(my_temp <= my_temp_end) 
+  {
+    solver.initialize_system(my_temp);
+    solver.simulate(mc_cycles);
+    solver.write_to_file();
+    my_temp += temperature_step;
+  }
+
+  MPI_Finalize();
+  return 0;
+}

### 2015-11-13 projects/project_4/tools/2d_ising.py

@@ -1,0 +1,85 @@
+import numpy as np
+from PIL import Image
+from random import randint, random
+import shutil, glob, os
+
+def create_image(configuration, dim_x, dim_y):
+    """
+    Given a matrix representing the current configuration of spins - assigns
+    the color blue to a pixel with positive sign and color white to pixel with
+    negative sign.
+    """
+    image = Image.new('RGB', (dim_x, dim_y), 'black')
+    pixels = image.load()
+
+    for a, i in enumerate(configuration):
+        for b, j in enumerate(i):
+            if j == 1:
+                pixels[a, b] = (255, 0, 0)
+            else:
+                pixels[a, b] = (0,0, 255)
+
+    return image
+
+def initialize(dim_x, dim_y):
+    """
+    given the dimensions of a grid
+    initializes it with random spin
+    """
+
+    matr = np.random.randint(2, size=(dim_x, dim_y))
+    for a, i in enumerate(matr):
+        for b, j in enumerate(i):
+            if j == 0:
+                matr[a, b] = -1
+    return matr
+def metropolis():
+    i, j = [randint(0, dim_x-1), randint(0, dim_y-1)]
+    E = get_energy(matr, i, j, dim_x, dim_y)
+    if E > 0:
+        matr[i, j] = -matr[i,j]
+    else:
+        r = random()
+        if r < np.exp(2 * E / T):
+            matr[i, j] = -matr[i,j]
+
+def get_energy(configuration, i, j, dim_x, dim_y):
+
+    this = configuration[i, j] 
+
+    up =  configuration[i, j-1] if (j > 0) else configuration[i, dim_y-1]
+    down =  configuration[i, j+1] if (j < dim_y-1) else configuration[i, 0]
+    left = configuration[i-1, j] if (i > 0) else configuration[dim_x-1, j]
+    right = configuration[i+1, j] if (i < dim_x-1) else configuration[0, j]
+
+    return -this * (up + down + left + right)
+
+
+if __name__ == '__main__':
+    subdir = 'animation'
+    if os.path.isdir(subdir): # if the directory is already present, remove it
+        shutil.rmtree(subdir)
+    os.mkdir(subdir) # create directory
+    os.chdir(subdir) # enter new directory
+    dim_x = 400
+    dim_y = 400
+    N = 30000000
+    T_values = np.linspace(1, 5, N)
+    matr = initialize(dim_x, dim_y)
[…]

### 2015-11-13 projects/project_4/tools/framework.py

@@ -1,0 +1,75 @@
+from numpy import linspace, sinh, cosh, exp
+from matplotlib.pyplot import plot, show, subplot, legend, xlabel, ylabel, title, suptitle, grid, savefig, tight_layout, subplots_adjust
+import os
+
+mc_cycles = 50000
+T_start = 2.0
+T_end = 2.4
+T_step = 0.01
+grid_dim = 80
+
+temperature = []
+mean_energy = []
+mean_magnetization = []
+specific_heat = []
+susceptibility = []
+absolute_magnetization = []
+file_name = '../data/output.dat'
+os.system('rm ../data/output.dat')
+os.system('../src/a.out %d %f %f %f %d %s' % (grid_dim, T_start, T_end, T_step, mc_cycles, file_name))
+for line in open('../data/output.dat', 'r').readlines():
+    data = line.split()
+    temperature.append(float(data[0]))
+    mean_energy.append(float(data[1]))
+    specific_heat.append(float(data[2]))
+    susceptibility.append(float(data[3]))
+    absolute_magnetization.append(float(data[4]))
+
+suptitle('$%d\\times%d$ Ising model with $%d$ MC-cycles' % (grid_dim, grid_dim, mc_cycles), y=1.00)
+subplots_adjust(top=0.5)
+subplot(221)
+grid('on')
+plot(temperature, mean_energy)
+xlabel('$k_BT$'); ylabel('$E / J$')
+subplot(222)
+grid('on')
+plot(temperature, specific_heat)
+xlabel('$k_BT$'); ylabel('$C_V / Jk_B$')
+subplot(223)
+grid('on')
+plot(temperature, susceptibility)
+xlabel('$k_BT$'); ylabel('$\\chi$')
+subplot(224)
+grid('on')
+plot(temperature, absolute_magnetization)
+xlabel('$k_BT$'); ylabel('$\\langle | M | \\rangle$')
+tight_layout()
+savefig('output.pdf')
+
+n = grid_dim**2 # number of spins
+partition_function = lambda temp : 4*cosh(8/temp) + 12 
[…]

### 2015-11-13 projects/project_4/tools/framework_c.py

@@ -1,0 +1,15 @@
+import matplotlib.pyplot as plt
+T = []
+accepts = []
+for line in open('../data/accepted_configs_as_function_of_temp.dat', 'r').readlines():
+    line = line.split()
+    T.append(float(line[0]) )
+    accepts.append(float(line[-1]))
+
+
+
+plt.plot(T, accepts)
+plt.xlabel('$k_BT$')
+plt.ylabel('Accepted configurations')
+plt.title('Number of accepted configurations as a function of temperature')
+plt.savefig('accepted_configs.pdf')

### 2015-11-13 projects/project_4/tools/framework_e.py

@@ -1,0 +1,77 @@
+import os
+import matplotlib.pyplot as plt
+temp = []
+energy = []
+magnet = []
+heat = []
+suscept = []
+
+grid_dims = [20, 40, 60, 80]
+for i, grid_dimension in enumerate(grid_dims):
+
+    t_start = 2.0
+    t_end = 2.4
+    t_step = 0.0005
+    cycles = 6000
+    file_name = '../data/e_%d.dat' % grid_dimension
+    os.system('rm ' + file_name)
+    command = '../src/a.out %d %f %f %f %d %s' % (grid_dimension, t_start, t_end, t_step, cycles,  file_name)
+
+    os.system(command)
+    
+    t = []
+    e = []
+    m = []
+    c = []
+    s = []
+
+    for line in open(file_name, 'r').readlines():
+        line = line.split()
+
+        t.append(float(line[0]))
+        e.append(float(line[1]))
+        m.append(float(line[2]))
+        c.append(float(line[3]))
+        s.append(float(line[4]))
+
+    temp.append(t)
+    energy.append(e)
+    magnet.append(m)
+    heat.append(c)
+    suscept.append(s)
+
+for i in range(len(temp)):
+    plt.plot(temp[i], energy[i], label='$L = %d$' % grid_dims[i])
+plt.legend(loc='best')
+plt.grid('on')
+plt.xlabel('$k_BT$')
+plt.ylabel('$E / J$')
+plt.savefig('energy_grid_size.pdf')
+plt.clf()
+
+for i in range(len(temp)):
+    plt.plot(temp[i], magnet[i], label='$L = %d$' % grid_dims[i])
+plt.legend(loc='best')
+plt.grid('on')
+plt.xlabel('$k_BT$')
+plt.ylabel('$\\langle |\\mathcal{M}| \\rangle$')
+plt.savefig('magnet_grid_size.pdf')
+plt.clf()
+
+for i in range(len(temp)):
+    plt.plot(temp[i], heat[i], label='$L = %d$' % grid_dims[i])
+plt.legend(loc='best')
+plt.grid('on')
+plt.xlabel('$k_BT$')
+plt.ylabel('$C_V$')
+plt.savefig('heat_grid_size.pdf')
+plt.clf()
+
+for i in range(len(temp)):
+    plt.plot(temp[i], suscept[i], label='$L = %d$' % grid_dims[i])
+plt.legend(loc='best')
[…]

### 2015-11-13 projects/project_4/tools/framework_mc_cycles.py

@@ -1,0 +1,36 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import os
+
+E_values = []
+M_values = []
+
+dimension = 20
+temp_start = 1.0
+temp_end = 1.0
+dtemp = 0.1
+
+mc_cycles = 10000
+file_name = 'throwaway.dat'
+
+os.system('rm ../data/expectations.dat')
+command = '../src/a.out %d %f %f %f %d %s' % (dimension, temp_start, temp_end, dtemp, mc_cycles, file_name)
+os.system(command)
+for line in open('../data/expectations.dat', 'r').readlines():
+    line = line.split()
+
+    E_values.append(float(line[1]))
+    M_values.append(float(line[2]))
+
+mc_cycles = [100*i for i in range(len(E_values))]
+plt.suptitle('$%d\\times%d$ Ising model with $T = %0.2f$' % (dimension,dimension, temp_start), y=1.00)
+plt.subplot(211)
+plt.plot(mc_cycles, E_values)
+plt.xlabel('Monte-Carlo cycles')
+plt.ylabel('$\\langle E \\rangle / J$')
+plt.subplot(212)
+plt.plot(mc_cycles, M_values)
+plt.xlabel('Monte-Carlo cycles')
+plt.ylabel('$\\langle |\\mathcal{M}| \\rangle / J$')
+plt.tight_layout()
+plt.savefig('../article/1_mc_random.pdf')

### 2015-11-13 projects/project_4/tools/probability_plotter.py

@@ -1,0 +1,17 @@
+import matplotlib.pyplot as plt
+import numpy as np
+
+
+energies = range(-800, 801)
+probabil = []
+
+for line in open('energy_prob_1.dat', 'r').readlines():
+    line = line.split()   
+    probabil.append(float(line[-1]))
+
+print sum(probabil)
+plt.plot(energies, probabil)
+plt.title('Probabilty of various energies for $T = 1$')
+plt.ylabel('$P(E)$')
+plt.xlabel('$E / J$')
+plt.savefig('probabilites.pdf')

### 2015-11-13 projects/project_4/tools/2d_ising.py
better animation

@@ -72,7 +72,7 @@ if __name__ == '__main__':
     framecounter = 0
     for T, n in zip(T_values, range(N)):
         metropolis()
-        if n % 10000 == 0:
+        if n % 50000 == 0:
             print 'ping', n
             create_image(matr, dim_x, dim_y).save('tmp_%5d.png' % framecounter)
             framecounter += 1

### 2015-11-17 projects/project_5/article.tex
started project 5, astro

\documentclass[a4paper, 11pt]{article}

\title{$N$-body simulation of an open galactic cluster}
\begin{document}
\maketitle
\begin{abstract}
    In this project we look at the Newtonian $N$-body problem using various
    numerical methods for solving ordinary differential equations. Specifically
    of interest is the fourt order Runge-Kutta method and the Velocity-Verlet
    method.
\end{abstract}

\section{$N$-body problem}
\label{sec:_n_body_problem}

The $N$-body problem considers $N$ point masses denoted $m_i$ with $i = 1, 2,
\ldots N$ in an inertial reference frame in three dimensional space
$\mathbb{R}^3$. To each point mass $m_i$ we can associate a position vector
$\mathbf{q}_i$.  According to newtons second law we have that the sum of the
forces on the mass is equal to the mass times the acceperation. We have from Newton's law of gravity that the gravitational force fel on $m_i$ by a single mass $m_j$ ($j \neq i$) is given by
\begin{equation}
    \notag
    \mathbf{F}_{ij} = \frac{Gm_im_j \left( \mathbf{q}_j - \mathbf{q}_i \right)}{\|g_j 0 q_i\|_{}^2}
\end{equation}

\end{document}

### 2015-11-17 projects/project_5/article.tex
Blindly copying N-body problem from wikipedia

@@
 \documentclass[a4paper, 11pt]{article}
+
+\usepackage[]{amsmath, amsthm, amssymb}
 
@@
 $\mathbf{q}_i$.  According to newtons second law we have that the sum of the
-forces on the mass is equal to the mass times the acceperation. We have from Newton's law of gravity that the gravitational force fel on $m_i$ by a single mass $m_j$ ($j \neq i$) is given by
+forces on the mass is equal to the mass times the acceperation. We have from
+Newton's law of gravity that the gravitational force felt on $m_i$ by a single
+mass $m_j$ ($j \neq i$) is given by
 \begin{equation}
     \notag
-    \mathbf{F}_{ij} = \frac{Gm_im_j \left( \mathbf{q}_j - \mathbf{q}_i \right)}{\|g_j 0 q_i\|_{}^2}
+    \mathbf{F}_{ij} = \frac{Gm_im_j \left( \mathbf{q}_j - \mathbf{q}_i \right)}{\|g_j - q_i\|_{}^3}
 \end{equation}
-
+with $G$ being the gravitational constant and $\|\mathbf{q}_j -
+\mathbf{q}_i\|_{}$ is the magnitude of the distance between $\mathbf{q}_i$ and
+$\mathbf{q}_j$.
+We can now sum over all masses, which gives us the $N$-body \emph{equations of
+motion}:
+\begin{equation}
+    \notag
+    m_i \frac{d^2\mathbf{q}_i}{dt^2} = \sum^{N}_{j=1, j\neq i} \frac{Gm_im_j \left( \mathbf{q}_j \right) - \mathbf{q}_i)}{\|\mathbf{q}_j - \mathbf{q}_i\|_{}^3} = \frac{\partial U}{\partial \mathbf{q}_i}
+\end{equation}
+where $U$ is the \emph{self-potential} energy
+\begin{equation}
+    \notag
+    U = \sum^{}_{1 \leq i < j \leq N} \frac{Gm_im_j}{\|\mathbf{q}_j-\mathbf{q}_i \|_{}}.
+\end{equation}
+We can now define the momentum associated to each mass $m_i$ to be $\mathbf{p}_i = m_id\mathbf{q}_i/d_t$, which gives us that \emph{Hamilton's equations of motion} for the $N$-body problem become
+\begin{align*}
+    \notag
+    \frac{d\mathbf{q}_i}{dt} = \frac{\partial H}{\partial \mathbf{p}_i} && \frac{d\mathbf{p}_i}{dt} = -\frac{\partial H}{\partial \mathbf{q}_i},
+\end{align*}
+where $H = T + U$ and $T$ is the kinetic energy given by
+\begin{equation}
+    \notag
+    T = \sum^{N}_{i=1} \frac{\|\mathbf{p}_i\|_{}^2}{2m_i}.
+\end{equation}
 \end{document}

### 2015-11-17 projects/project_5/article.tex
kept writing

@@
 \usepackage[]{amsmath, amsthm, amssymb} 
+\usepackage[]{algorithmic}
 
 \title{$N$-body simulation of an open galactic cluster}
+\author{Ivar Haugal{\o}kken Stangeby}
+
 \begin{document}
@@
 
+\tableofcontents
+
 \section{$N$-body problem}
 \label{sec:_n_body_problem}
+
+In this section we look at the general form of the $N$-body problem, as well as
+examine the special case of $N = 2$, the two-body problem.
+
+\subsection{General form}
+\label{sub:general_form}
 
@@
     \notag
-    \mathbf{F}_{ij} = \frac{Gm_im_j \left( \mathbf{q}_j - \mathbf{q}_i \right)}{\|g_j - q_i\|_{}^3}
+    \mathbf{F}_{ij} = \frac{Gm_im_j \left( \mathbf{q}_j - \mathbf{q}_i \right)}{\|\mathbf{q}_j - \mathbf{q}_i\|_{}^3}
 \end{equation}
@@
 \end{equation}
+
+\subsection{Two-body problem}
+\label{sub:two_body_problem}
+
+If we consider the motion of two bodies, for instance the Earth-Sun system, then:
+\begin{align*}
+    \mathbf{F}_{12} = \frac{Gm_1m_2 \left( \mathbf{q}_2 - \mathbf{q}_1 \right)}{\|\mathbf{q}_j - \mathbf{q}_i\|_{}^3} && \text{Earth Sun} \\
+    \mathbf{F}_{21} = \frac{Gm_2m_1 \left( \mathbf{q}_1 - \mathbf{q}_2 \right)}{\|\mathbf{q}_1 - \mathbf{q}_2\|_{}^3} && \text{Sun Earth}
+\end{align*}
+
+\section{Numerical Methods}
+\label{sec:numerical_methods}
+
+We examine the numerical methods used in this project, and look at various
+properties related to the performance of these methods in $N$-body simulations.
+We benchmark using a two-body problem.
+
+\subsection{4-th order Runge-Kutta}
+\label{sub:4_th_order_runge_kutta}
+
+In the fourth-order Runge-Kutta method, also known as \emph{the classical
+Runge-Kutta method} we determine the next values the approximations using the
+value before as well as a weighted average of four increments where each
+increment is a multiple of the interval and an estimated slope, given by an
+explicit formula for the first derivative.
+
+\subsubsection{Algorithm}
+\label{ssub:algorithm}
+
+Specify an initial value problem as
+\begin{equation}
+    \notag
+    \dot{y} = f(t, y), \quad y(t_0) = y_0
+\end{equation}
+where $f$, $t_0$ and $y_0$ is given. Then the fourth order Runge-Kutta method
+can be described as follows:
+\begin{algorithmic}
+    \FOR{$n = 0$ to $N-1$}
+    \STATE $y_{n+1} = y_n + \frac{h}{6}\left( k_1 + 2k_2 + 2k_3 + k_4 \right)$
+    \STATE $t_{n+1} = t_n + h$
+    \ENDFOR
+\end{algorithmic}
+where 
+\begin{flalign*}
+    k_1 &= f(t_n, y_n), & k_2 &= f(t_n + h/2, y_n + k_1 h/2), \\
+    k_3 &= f(t_n + h/2, y_n + k_2h/2), &  k_4 &= f(t_n + h, y_n + hk_3).
+\end{flalign*}
+\subsection{Velocity-Verlet}
+\label{sub:velocity_verlet}
 \end{document}

### 2015-11-17 projects/project_5/python_prototype/N_body_problem.py
Working on python thingy

@@ -1,0 +1,87 @@
+"""
+Solves the N-body problem using RungeKutta or Velocity-Verlet
+"""
+import ODESolver
+import numpy as np
+import random
+import matplotlib.pyplot as plt
+from  mpl_toolkits.mplot3d import Axes3D
+
+def pick_point_in_sphere(R):
+    """
+    Picks a uniformly random point in a sphere of radius R.  Pretends to be picking in a
+    cube, and discards anything outside the sphere range.
+    """
+    while True:
+        x = random.random()*R    
+        y = random.random()*R
+        z = random.random()*R
+        if np.sqrt(x**2 + y**2 + z**2) < R:
+            return np.array([x, y, z])
+    
+class Cluster:
+    """
+    Represents a system of N particles
+    """ 
+    G = 6.6740831E-11
+    point_of_origin = [0, 0, 0] 
+    cb_list = []
+    def __init__(self, N=2, n_steps=1000):
+        Cluster.cb_list = [CelestialBody(n=n_steps, M=1.0e30) for i in range(N)]
+        self.solvers = [ODESolver.RungeKutta4]
+    
+    def simulate(self, t_values):
+        for i, t in enumerate(t_values[:-1]):
+            dt = t_values[i+1] - t
+            for body in Cluster.cb_list:
+                body.step(dt, i)
+    
+    def visualize(self):
+        fig = plt.figure()
+        ax = fig.add_subplot(111, projection='3d')
+        for body in Cluster.cb_list:
+            print 'plotting for ', body
+            ax.plot(body.r[:,0], body.r[:,1], body.r[:,2])
+        plt.show() 
+            
+class CelestialBody:
+    """
+    Represents one celestial particle, either a star or a small number of
+    stars. With is is associated a size (mass), a position vector, a velocity
+    vector and an acceleration vector.
+    """
+ 
+    def __init__(self, radius=1, M=1.0e5, n=100):
+         
+        self.r = np.zeros([n+1, 3])
+        self.r[0] = pick_point_in_sphere(radius)
+        self.v = np.zeros([n+1, 3])
+        self.a = np.zeros([n+1, 3])
+        self.M = float(M)
+         
+    
+    def __call__(self, celestial_body_list, i):
+        """
+        Returns the total force exerted by all the other celestial bodies on
+        this body.
+        """    
+        total_force = 0 
+        for body in celestial_body_list:
+            if body == self:
[…]

### 2015-11-19 projects/project_5/python_prototype/CelestialBody.py
Started visualization

@@ -1,0 +1,61 @@
+import numpy as np
+class CelestialBody:
+    """
+    Represents one celestial particle. With it is associated a size, a position
+    vector, a velocity vector and an acceleration vector. 
+    """
+    G = 6.6740831E-11
+    def __init__(self, radius=20.0, M=1.0e5, n=10):
+        """
+        Initializes the celestial particle with vectors of given length, a mass
+        and a radius.
+        """
+        self.r = np.zeros([n+1, 3])
+        self.v = np.zeros([n+1, 3])
+        self.a = np.zeros([n+1, 3])
+        self.M = np.zeros([n+1, 3])
+        self.R = radius
+
+    def set_initial_conditions(self, r0, v0, a0):
+        """
+        Sets the initial position, velocity and acceleration of the celestial
+        body.
+        """
+        try:
+            self.r[0, :] = r0
+            self.v[0, :] = v0
+            self.a[0, :] = a0
+        except IndexError:
+            print 'Cannot assign to index - is the Celestial-Body initialized with a proper n?'
+        except ValueError:
+            print 'Invalid arguments - expects numpy arrays of length 3'
+    
+    def __call__ (self, celestial_body_list, i):
+        """
+        Returns the total forces exerted by all the other celestial bodies on
+        this body.
+        """
+        total_force = [0, 0, 0]
+        for body in celestial_body_list:
+            if body == self:
+                print 'Not calculating self'
+                continue
+            total_force += Cluster.G * self.M * body.M * (self - (body, i))/float(np.linalg.norm(self - (body, i)))** 3
+
+        print type(total_force)
+        return total_force
+
+    def __sub__(self, (other, i)):
+        """
+        Returns the distance vector between two celestial bodies at timestep i
+        """
+        return other.r[i,:] - self.r[i, :] 
+
+    def step(self, dt, i, cb_list):
+        """
+        Pushes the celestial body one time increment - Using EulerCromer for now
+        """
+        self.a[i] = self(cb_list, i) / self.M
+        self.v[i+1] = self.v[i] + self.a[i] * dt
+        self.r[i+1] = self.r[i] + self.v[i] * dt
+

### 2015-11-19 projects/project_5/python_prototype/Cluster.py

@@ -1,0 +1,38 @@
+import numpy as np
+import random
+import matplotlib.pyplot as plt
+from  mpl_toolkits.mplot3d import Axes3D
+
+from CelestialBody import CelestialBody
+class Cluster:
+    """
+    Represents a system of N particles
+    """
+    point_of_origin = [0, 0, 0] # might need this
+
+    def __init__(self, N=2, n_steps=1000):
+        self.cb_list = [CelestialBody(radius=random.randint(30, 500), M=1.0e5, n=n_steps) for i in range(N)]
+         
+    def visualize(self, i=0):
+        fig = plt.figure()
+        ax = fig.add_subplot(111, projection='3d')
+        for body in self.cb_list:
+            ax.scatter(body.r[i,0], body.r[i,1], body.r[i,2], s=body.R)
+        plt.show()
+        
+    def initialize(self, radius=20):
+        for body in self.cb_list:
+            body.set_initial_conditions(r0=Cluster.pick_point_in_sphere(radius), v0=[0, 0, 0], a0=[0, 0, 0])
+
+    @staticmethod
+    def pick_point_in_sphere(R):
+        """
+        Picks a uniformly random point in a sphere of radius R. Pretends to be
+        picking in a cube, then discards anything outside the sphere radius.
+        """
+        while True:
+            x = random.random()*R
+            y = random.random()*R
+            z = random.random()*R
+            if np.sqrt(x**2 + y**2 + z**2) < R:
+                return np.array([x, y, z])

### 2015-11-19 projects/project_5/python_prototype/N_body_problem.py

@@ -1,87 +1,10 @@
 """
 Solves the N-body problem using RungeKutta or Velocity-Verlet
 """
-import ODESolver
-import numpy as np
-import random
-import matplotlib.pyplot as plt
-from  mpl_toolkits.mplot3d import Axes3D
-
-def pick_point_in_sphere(R):
-    """
-    Picks a uniformly random point in a sphere of radius R.  Pretends to be picking in a
-    cube, and discards anything outside the sphere range.
-    """
-    while True:
-        x = random.random()*R    
-        y = random.random()*R
-        z = random.random()*R
-        if np.sqrt(x**2 + y**2 + z**2) < R:
-            return np.array([x, y, z])
-    
-class Cluster:
-    """
-    Represents a system of N particles
-    """ 
-    G = 6.6740831E-11
-    point_of_origin = [0, 0, 0] 
-    cb_list = []
-    def __init__(self, N=2, n_steps=1000):
-        Cluster.cb_list = [CelestialBody(n=n_steps, M=1.0e30) for i in range(N)]
-        self.solvers = [ODESolver.RungeKutta4]
-    
-    def simulate(self, t_values):
-        for i, t in enumerate(t_values[:-1]):
-            dt = t_values[i+1] - t
-            for body in Cluster.cb_list:
-                body.step(dt, i)
-    
-    def visualize(self):
-        fig = plt.figure()
-        ax = fig.add_subplot(111, projection='3d')
-        for body in Cluster.cb_list:
-            print 'plotting for ', body
-            ax.plot(body.r[:,0], body.r[:,1], body.r[:,2])
-        plt.show() 
-            
-class CelestialBody:
-    """
-    Represents one celestial particle, either a star or a small number of
-    stars. With is is associated a size (mass), a position vector, a velocity
-    vector and an acceleration vector.
-    """
- 
-    def __init__(self, radius=1, M=1.0e5, n=100):
-         
-        self.r = np.zeros([n+1, 3])
-        self.r[0] = pick_point_in_sphere(radius)
-        self.v = np.zeros([n+1, 3])
-        self.a = np.zeros([n+1, 3])
-        self.M = float(M)
-         
-    
-    def __call__(self, celestial_body_list, i):
-        """
-        Returns the total force exerted by all the other celestial bodies on
-        this body.
-        """    
-        total_force = 0 
-        for body in celestial_body_list:
-            if body == self:
[…]

### 2015-11-19 projects/project_5/python_prototype/CelestialBody.py
Fixed mass bug

@@ -13,7 +13,7 @@ class CelestialBody:
         self.r = np.zeros([n+1, 3])
         self.v = np.zeros([n+1, 3])
         self.a = np.zeros([n+1, 3])
-        self.M = np.zeros([n+1, 3])
+        self.M = M
         self.R = radius
 
     def set_initial_conditions(self, r0, v0, a0):
@@ -33,16 +33,16 @@ class CelestialBody:
     def __call__ (self, celestial_body_list, i):
         """
         Returns the total forces exerted by all the other celestial bodies on
-        this body.
+        this body at a timestep i.
         """
         total_force = [0, 0, 0]
         for body in celestial_body_list:
             if body == self:
                 print 'Not calculating self'
                 continue
-            total_force += Cluster.G * self.M * body.M * (self - (body, i))/float(np.linalg.norm(self - (body, i)))** 3
+            else: 
+                total_force += CelestialBody.G * self.M * body.M * (self - (body, i))/float(np.linalg.norm(self - (body, i)))**3
 
-        print type(total_force)
         return total_force
 
     def __sub__(self, (other, i)):
@@ -55,7 +55,7 @@ class CelestialBody:
         """
         Pushes the celestial body one time increment - Using EulerCromer for now
         """
-        self.a[i] = self(cb_list, i) / self.M
-        self.v[i+1] = self.v[i] + self.a[i] * dt
-        self.r[i+1] = self.r[i] + self.v[i] * dt
+        self.a[i,:] = self(cb_list, i) / self.M
+        self.v[i+1,:] = self.v[i,:] + self.a[i,:] * dt
+        self.r[i+1,:] = self.r[i,:] + self.v[i,:] * dt
 

### 2015-11-19 projects/project_5/python_prototype/Cluster.py

@@ -10,16 +10,33 @@ class Cluster:
     """
     point_of_origin = [0, 0, 0] # might need this
 
-    def __init__(self, N=2, n_steps=1000):
+    def __init__(self, N=2, n_steps=6):
+        self.N = N
+        self.n_steps = n_steps
         self.cb_list = [CelestialBody(radius=random.randint(30, 500), M=1.0e5, n=n_steps) for i in range(N)]
          
     def visualize(self, i=0):
+        """
+        Shows a snapshot of the current state of the system at time t_i.
+        Particles are scaled by size.
+        """
+
         fig = plt.figure()
         ax = fig.add_subplot(111, projection='3d')
         for body in self.cb_list:
             ax.scatter(body.r[i,0], body.r[i,1], body.r[i,2], s=body.R)
         plt.show()
-        
+        plt.clf()
+    
+    def animate(self):
+        """
+        Animates the time-evolution of the system.
+        """
+        for i in range(self.n_steps):
+            self.visualize(i)
+            for body in self.cb_list:
+                body.step(1, i, self.cb_list)
+
     def initialize(self, radius=20):
         for body in self.cb_list:
             body.set_initial_conditions(r0=Cluster.pick_point_in_sphere(radius), v0=[0, 0, 0], a0=[0, 0, 0])

### 2015-11-19 projects/project_5/python_prototype/N_body_problem.py

@@ -4,7 +4,7 @@ Solves the N-body problem using RungeKutta or Velocity-Verlet
 from Cluster import Cluster
 if __name__ == "__main__":
 
-    test = Cluster(N=4)
+    test = Cluster(N=4, n_steps=100)
     test.initialize()
-    test.visualize()
+    test.animate()
 

### 2015-11-27 projects/project_5/astro/python_prototype/CelestialBody.py
stopped working on astro project, moved to MD

@@ -1,29 +1,26 @@
 import numpy as np
-
 class CelestialBody:
     """
     Represents one celestial particle. With it is associated a size, a position
     vector, a velocity vector and an acceleration vector. 
     """
-    G = 1
-    def __init__(self, radius=1, mass=10.0, n=10):
+    G = 6.6740831E-11
+    def __init__(self, radius=20.0, M=1.0e5, n=10):
         """
-        Mass is given in multiples of solar masses. Distance is given in light
-        years and the gravitational constant is set to 1. radius given in km
+        Initializes the celestial particle with vectors of given length, a mass
+        (given in number of suns) and a radius.
         """
         self.r = np.zeros([n+1, 3])
         self.v = np.zeros([n+1, 3])
         self.a = np.zeros([n+1, 3])
-        
-        self.mass = mass
-        self.radius = radius
+        self.M = M
+        self.R = radius
 
-    def set_initial_conditions(self, U0):
+    def set_initial_conditions(self, r0, v0, a0):
         """
-        Initializes the position, velocity and acceleration of the celstial
+        Sets the initial position, velocity and acceleration of the celestial
         body.
         """
-        r0, v0, a0 = U0 
         try:
             self.r[0, :] = r0
             self.v[0, :] = v0
@@ -32,20 +29,21 @@ class CelestialBody:
             print 'Cannot assign to index - is the Celestial-Body initialized with a proper n?'
         except ValueError:
             print 'Invalid arguments - expects numpy arrays of length 3'
-
-    def __call__(self, celestial_body_list, i):
+    
+    def __call__ (self, celestial_body_list, i):
         """
         Returns the total forces exerted by all the other celestial bodies on
         this body at a timestep i.
         """
         total_force = [0, 0, 0]
-        collision_correction = 1.0
         for body in celestial_body_list:
             if body == self:
                 continue
             else: 
-                total_force += CelestialBody.G * self.mass * body.mass * (self - (body, i))/float(np.linalg.norm(self - (body, i)) + collision_correction)**3
-        return total_force
+                total_force += CelestialBody.G * self.M * body.M * (self - (body, i))/float(np.linalg.norm(self - (body, i)))**3
+        print np.linalg.norm(total_force)
+        return -total_force
+
 
     def __sub__(self, (other, i)):
         """
@@ -57,6 +55,7 @@ class CelestialBody:
         """
[…]

### 2015-11-27 projects/project_5/astro/python_prototype/Cluster.py

@@ -13,9 +13,9 @@ class Cluster:
     def __init__(self, N=2, n_steps=6):
         self.N = N
         self.n_steps = n_steps
-        self.cb_list = [CelestialBody(radius=random.randint(30, 500), M=1.0e5, n=n_steps) for i in range(N)]
+        self.cb_list = [CelestialBody(radius=random.randint(30, 500), M=float(random.randint(9, 11)), n=n_steps) for i in range(N)]
          
-    def visualize(self, i=0):
+    def visualize(self, i):
         """
         Shows a snapshot of the current state of the system at time t_i.
         Particles are scaled by size.
@@ -26,7 +26,6 @@ class Cluster:
         for body in self.cb_list:
             ax.scatter(body.r[i,0], body.r[i,1], body.r[i,2], s=body.R)
         plt.show()
-        plt.clf()
     
     def animate(self):
         """
@@ -34,6 +33,7 @@ class Cluster:
         """
         for i in range(self.n_steps):
             self.visualize(i)
+            print self.cb_list[0].v[i,:]
             for body in self.cb_list:
                 body.step(1, i, self.cb_list)
 
@@ -47,9 +47,12 @@ class Cluster:
         Picks a uniformly random point in a sphere of radius R. Pretends to be
         picking in a cube, then discards anything outside the sphere radius.
         """
+        astronomical_unit = 1.4960e11
         while True:
+
             x = random.random()*R
             y = random.random()*R
             z = random.random()*R
+
             if np.sqrt(x**2 + y**2 + z**2) < R:
                 return np.array([x, y, z])

### 2015-11-27 projects/project_5/astro/python_prototype/N_body_problem.py

@@ -4,7 +4,7 @@ Solves the N-body problem using RungeKutta or Velocity-Verlet
 from Cluster import Cluster
 if __name__ == "__main__":
 
-    test = Cluster(N=4, n_steps=100)
-    test.initialize()
+    test = Cluster(N=4, n_steps=5)
+    test.initialize(radius=100)
     test.animate()
 

### 2015-11-27 projects/project_5/astro/python_prototype/revised/CelestialBody.py

@@ -1,26 +1,29 @@
 import numpy as np
+
 class CelestialBody:
     """
     Represents one celestial particle. With it is associated a size, a position
     vector, a velocity vector and an acceleration vector. 
     """
-    G = 6.6740831E-11
-    def __init__(self, radius=20.0, M=1.0e5, n=10):
+    G = 1
+    def __init__(self, radius=1, mass=10.0, n=10):
         """
-        Initializes the celestial particle with vectors of given length, a mass
-        and a radius.
+        Mass is given in multiples of solar masses. Distance is given in light
+        years and the gravitational constant is set to 1. radius given in km
         """
         self.r = np.zeros([n+1, 3])
         self.v = np.zeros([n+1, 3])
         self.a = np.zeros([n+1, 3])
-        self.M = M
-        self.R = radius
+        
+        self.mass = mass
+        self.radius = radius
 
-    def set_initial_conditions(self, r0, v0, a0):
+    def set_initial_conditions(self, U0):
         """
-        Sets the initial position, velocity and acceleration of the celestial
+        Initializes the position, velocity and acceleration of the celstial
         body.
         """
+        r0, v0, a0 = U0 
         try:
             self.r[0, :] = r0
             self.v[0, :] = v0
@@ -29,20 +32,19 @@ class CelestialBody:
             print 'Cannot assign to index - is the Celestial-Body initialized with a proper n?'
         except ValueError:
             print 'Invalid arguments - expects numpy arrays of length 3'
-    
-    def __call__ (self, celestial_body_list, i):
+
+    def __call__(self, celestial_body_list, i):
         """
         Returns the total forces exerted by all the other celestial bodies on
         this body at a timestep i.
         """
         total_force = [0, 0, 0]
+        collision_correction = 2.0
         for body in celestial_body_list:
             if body == self:
-                print 'Not calculating self'
                 continue
             else: 
-                total_force += CelestialBody.G * self.M * body.M * (self - (body, i))/float(np.linalg.norm(self - (body, i)))**3
-
+                total_force += CelestialBody.G * self.mass * body.mass * (self - (body, i))/float(np.linalg.norm(self - (body, i)) + collision_correction**2)**3
         return total_force
 
     def __sub__(self, (other, i)):
@@ -55,7 +57,6 @@ class CelestialBody:
         """
[…]

### 2015-11-27 projects/project_5/astro/python_prototype/revised/Cluster.py

@@ -4,6 +4,7 @@ import numpy as np
 import os
 import shutil, glob
 import random
+from scipy.misc import imread
 
 from CelestialBody import CelestialBody
 class Cluster:
@@ -15,7 +16,7 @@ class Cluster:
     def __init__(self, N=2, n_steps=1000):
         self.N = N
         self.n_steps = n_steps
-        self.cb_list = [CelestialBody(radius=random.randint(30, 70), mass=float(random.randint(9, 11)), n=n_steps) for i in range(N)]
+        self.cb_list = [CelestialBody(radius=random.randint(30,70), mass=float(random.randint(9, 11)), n=n_steps) for i in range(N)]
          
     def visualize(self, i):
         """
@@ -24,11 +25,13 @@ class Cluster:
         """
         fig = plt.figure()
         ax = fig.add_subplot(111, projection='3d')
-        ax.set_xlim3d([-20, 20])
-        ax.set_ylim3d([-20, 20])
-        ax.set_zlim3d([-20, 20])
+        ax.set_xlim3d([-50, 50])
+        ax.set_ylim3d([-50, 50])
+        ax.set_zlim3d([-50, 50])
+        ax.view_init(elev=11, azim=360*float(i)/self.n_steps)
         for body in self.cb_list:
-            ax.scatter(body.r[i,0], body.r[i,1], body.r[i,2], s=body.radius)
+            ax.scatter(body.r[i,0], body.r[i,1], body.r[i,2], s=body.radius, c=body.radius**4/70**4, cmap='autumn')
+        plt.axis('off')
         plt.savefig('tmp_%04d.png' % i)
         plt.close()
     
@@ -41,11 +44,14 @@ class Cluster:
             shutil.rmtree(subdir)
         os.mkdir(subdir) # create directory
         os.chdir(subdir) # enter new directory 
-
+        
+        dt = 0.5
         for i in range(self.n_steps):
+            if i % 10 == 0:
+                print i, '/', self.n_steps
             self.visualize(i)
             for body in self.cb_list:
-                body.step(0.1, i, self.cb_list)
+                body.step(dt, i, self.cb_list)
 
         cmd = 'convert -delay 4 tmp_*.png movie.gif'
         os.system(cmd)

### 2015-11-27 projects/project_5/astro/src/two_body/main.cpp

@@ -1,0 +1,7 @@
+#include <armadillo>
+int main(int argc, char *argv[])
+{
+    n =  
+    r1 
+    return 0;
+}

### 2015-11-27 projects/project_5/molecular_dynamics/article/article.tex
working on MD

\documentclass{article}
\usepackage[]{enumerate} 
\usepackage[]{graphicx} 
\usepackage[]{subcaption} 
\usepackage[]{amsthm, amsmath, amssymb} 
\usepackage[]{hyperref} 
\graphicspath{{../pictures/}}

\renewcommand{\vec}[1]{\mathbf{#1}}

\title{%
    \textsc{Atomic modeling of argon}\\
    \footnotesize{\textsc{Project 5 --- FYS3150}}
}
\author{%
    Ivar Haugal{\o}kken Stangeby\\
    \footnotesize{\textsc{Candidate no. 30}}
}

\begin{document}
\maketitle    
\tableofcontents
\section{Introduction}
\label{sec:introduction}
    Molecular dynamics (referred to as MD), deals with the time evolution of a
    system of particles, typically atoms or molecules. Being first explored in
    the late 1950's, it has had a wide range of applications, in fields as
    materials science, biochemistry and biophysics. It is based on the simple
    idea of considering a finite number of molecules and letting them interact
    for a fixed amount of time, which gives us an idea of how the system
    evolves.

    In this project we examine what goes into creating a working MD-program
    that can compute various physical quantities and then proceed by
    considering some specific systems where we look at, amongst other, the
    kinetic energy, the initial temperature required in order to reach a
    desired final temperature, and finding the diffusion constant for the
    system in order to compute the melting temperature.

\section{Building a working MD-program}
\label{sec:building_a_working_md_program}

    Initially, we are presented with the skeleton of a MD-program, and we are
    tasked with filling in the blanks. In this section, we walk through each of
    the steps that went into the implementation of the program, from skeleton
    to working program. The process can be summarized as

    \begin{enumerate}[1)]
        \item Introduce periodic boundary conditions.
        \item Implement a function for setting the net-momentum for the system to zero.
        \item Make the atoms conform to a face-centered cubic lattice configuration.
        \item Introduce the force model to the system, given by the Lennard-Jones potential.
        \item Implement the Velocity-Verlet algorithm in order to get a better
            conservation of energy.
        \item Write functions for sampling and storing physical quantities
    \end{enumerate}

    Running the program in its initial state yields a random configuration of
    the atoms. Since there is no force acting on the molecules they move in
    straight lines determined by their initial velocity vector. Their motion is
    unbounded, due to the periodic boundary condition not being implemented
    yet. This can be seen in figure \ref{fig:first_iteration}.
    \begin{figure}
        \centering \includegraphics[width=0.4\linewidth]{first_iteration.png}
        \caption[Initial state of the program]{First iteration of the program --- no boundary conditions and
            no force acting on the atoms. The atoms move with a constant speed
            in the direction of their initial velocity and the initial
        positions are uniformly distributed.}
        \label{fig:first_iteration}
    \end{figure}

\subsection{Periodic boundary conditions}
\label{sub:periodic_boundary_conditions}

    We want our system to conform to periodic boundary conditions. The reasons
    for doing this is that to a certain extent the system of finite size more
    closely resembles that of an infinite size. It is important that the
    bounding box which determines the boundary of the system is of sufficient
    size, so we do not get artifacts that arise from the fact that an atom can
    interact with itself.

    There are two approaches to the implementation of periodic boundary
    condition. The most basic, and the one we will employ here is to have an
    atom exit through one face of the bounding box, and reappear through the
    opposite face.

    Another approach is to not restrict the coordinates of the atom using the
    bounding box, and instead use the \emph{periodic image} of the atom when
    calculating interactions with ''nearby'' particles.

    We implement the function \texttt{applyPeriodicBoundaryConditions} in the
    \texttt{System}-class.  Our method is simply iterating over each atom, and
    in the cases where we find an atom outside the bounding box, we
    add/subtract the size of the box. If we let the vector $\vec{r}_i$, with
    components $r_q$ for $q = x, y, z$, denote the position of atom $i$ at any
    given time, we simply let the position be governed by the following
    relation:
    \begin{equation}
        \notag
        r_q = \begin{cases}
            r_q + L_q, & r_q < 0; \\
            r_q - L_q, & r_q > L_q; \\
            r_q, & \text{else.}
        \end{cases}
    \end{equation}
    Here $L_q$ is assumed to be the dimensions of the system in direction $q =
    x, y, z$.

    During simulations we need to be a bit cautious about what we define as the
    \emph{distance between} two atoms $i$ and $j$. Typically, we see this as
    the distance through the system, but due to our periodic boundary
    conditions there might be a shorter path through one of the borders of the
    system. This is formally stated as the \emph{minimum image
    criterion}\cite{min_image}.

\subsection{Zero out momentum}
\label{sub:zero_out_momentum}

    The Maxwell-Boltzmann distribution is good for velocities because it fairly
    accurately represents the way velocities are distributed in an ideal gas.
    The distribution was defined initially by Maxwell, for the specific purpose
    of describing particle speeds inside a stationary container with no
    interaction.  The distribution is directly dependent on the temperature of
    the system as well as the mass of each particle. We want to initialize each
    atom with a random velocity given by the Maxwell-Boltzmann distribution.

    Mathematically, the distribution function reads
    \begin{equation}
        \notag
        P(v_i)\, dv_i = \left( \frac{m_i}{2\pi k_BT} \right)^{1/2} e
    \end{equation}

    The problem that arises when using the Maxwell-Boltzmann distribution is
    that in its initial state the system has a non-zero net momentum. This
    causes the system to drift. This is why we want to implement the function
    \texttt{removeMomentum} in the \texttt{System}-class. Its purpose is to
    compute the total momentum of the system, and then remove a small portion
    of momentum from each atom in order to have a net momentum of zero.

    We compute the velocity of the center of mass of the system, and then
    subtract this velocity from each atom.  Let $m_i$ and $v_i$ denote the mass
    and velocity of atom $i$ respectively and let $M$ and $V$ denote the total
    mass of the system and the velocity of the center of mass. Then
    \begin{align*}
        \notag
        V = \sum^{n}_{i=1} m_i v_i, \quad M = \sum^{n}_{i=1} m_i
    \end{align*}
    where $n$ is the total number of atoms in the system. We then need to
    subtract the value $V_{\mathrm{sub}} = V / M$ from each $v_i$.

\subsection{Face-centered cubic lattice}
\label{sub:face_centered_cubic_lattice}

    We now, instead of having random initial positions, want the atoms to form
    a crystal structure. In a face-centered cubic lattice (FCC), each unit cell
    has, in addition to the eight corner lattice points, one lattice point for
    each face.

    A \emph{unit cell} is a group of atoms which can be stacked on top and next
    to each other to form a grid structure. The \emph{lattice constant} defines
    the size of the unit cell, and is denoted $b$. In this project we are
    assuming cubic grids, so we let $N$ denote the number of unit cells in each
    dimension.

    We now define a local coordinate system for each unit cell as
    \begin{align*}
        \vec{r}_1 &= 0\hat{\vec{i}} + 0\hat{\vec{j}} + 0\hat{\vec{k}}, &
        \vec{r}_2 &= \frac{b}{2}\hat{\vec{i}} + \frac{b}{2}\hat{\vec{j}} +
        0\hat{\vec{k}},\\ \vec{r}_3 &= 0\hat{\vec{i}} +
        \frac{b}{2}\hat{\vec{j}} + \frac{b}{2}\hat{\vec{k}}, & \vec{r}_4 &=
        \frac{b}{2}\hat{\vec{i}} + 0\hat{\vec{j}} + \frac{b}{2}\hat{\vec{k}}.
    \end{align*}
    The origin for an arbitrary unit cell is then given (globally) as
    \begin{equation}
        \notag \vec{R}_{i, j, k} = i\hat{\vec{u}}_1 + j\hat{\vec{u}}_2 +
        k\hat{\vec{u}}_3
    \end{equation}
    We implement the function \texttt{createFFCLattice} in the
    \texttt{System}-class which takes $N$ and $b$ as arguments. The initial
    configuration for an example system is displayed in figure
    \ref{fig:ffc_iteration}.
    \begin{figure}
        \centering \includegraphics[width=0.4\linewidth]{ffc_iteration}
        \caption[Crystallic structure in argon]{Initial configuration of a
            FFC-based lattice. We have here used a lattice constant $b =
            5.26${\AA} and the number of unit cells in each dimension is $N =
            5$. This specific system has a density of $\rho = M / V = 4N^3m /
            b^3N^3$ where $m$ is the mass of each atom, $b$ and $N$ given as
            above. This evaluates to approximately $0.1445 \,\mathrm{a.m.u} /
        \mathrm{\AA}^3$.}
        \label{fig:ffc_iteration}
    \end{figure}

\subsection{Lennard-Jones potential}
\label{sub:lennard_jones_potential}

    The Lennard-Jones potential approximates the interaction between a pair of
    atoms. In its most common form it reads
    \begin{equation}
        \notag U(r_{ij}) = 4\varepsilon \left[ \left( \frac{\sigma}{r_{ij}}
        \right)^{12} - \left( \frac{\sigma}{r_{ij}} \right)^6 \right],
    \end{equation}
    where $r_{ij}$ is the distance between atom $i$ and atom $j$ and
    $\varepsilon$ is the depth of the potential well. The quantity $\sigma$ is
    the distance for which the potential is zero.  With an expression for the
    potential we can sum over all distinct pairs of atoms and acquire an
    expression for the total potential energy $V$:
    \begin{equation}
        \notag V = \sum^{}_{i>j}U(r_{ij}).
    \end{equation}
    The force exerted on atom $i$ by atom $j$ is then given as the negative
    gradient of the potential between the two:
    \begin{equation}
        \notag \vec{F}(r_{ij}) = -\nabla U(r_{ij}).
    \end{equation}
    The three force-components are given by:
    \begin{align*}
        F_q(r_{ij}) = -\frac{\partial U}{\partial r_{ij}} \frac{\partial
        r_ij}{\partial q_{ij}} \quad (q = x, y, z)
    \end{align*}
    Analytically, this evaluates to
    \begin{align*}
        F_q(r_{i, j}) = 24\varepsilon \left[ \left(
        \frac{2\sigma^{12}}{r_{ij}^{13}}\right) - \left(
        \frac{\sigma^6}{r_{ij}^7}\right)\right] \quad (q = x, y, z).
    \end{align*}
    The brute force way of computing this is summing over each distinct pair of
    atoms, and computing the mutual force between them. This however,
    constitutes a fairly heavy computational load that quickly scales with the
    size of the system. For a total of $n$ atoms in the system, we have a time
    complexity of $\mathcal{O}(n^2)$. Luckily more efficient methods are
    available, most notably is the Verlet list method.

    Argon has a melting point of $\approx 84$K and a boiling point of $\approx
    87$K. If we run the simulations, once for a temperature of $40$K, once for
    $85$K and finally once for $120K$ we should see indications of three
    different phases. The results shows that the model somewhat accurately
    manages to capture these phase transitions. Snapshots taken at the same
    point in time for all three cases are presented in figure
    \ref{fig:potential_iteration}.
    \begin{figure}
        \begin{tabular}[c]{ccc}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{solid_40K.png}
                \caption{$T = 40$K}
            \end{subfigure}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{liquid_85K.png}
                \caption{$T = 85$K}
            \end{subfigure}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{gas_120K.png}
                \caption{$T = 120$K}
            \end{subfigure}
        \end{tabular}
        \caption[Argon phases]{Snapshots taken of the system at frame 500/999 for
            temperatures of $40$, $85$ and $120$ kelvin. It is a bit difficult
            to conclude from the pictures, one really needs to see the system
            in motion, but the system do behave more or less like a solid, a
            liquid and a gas for respective temperatures. Any artifacts can be
            contributed to numerical instability. It is possible to see a
            certain change in the entropy of the system across the three
            images. These made use of the Euler-Cromer integration method.}
        \label{fig:potential_iteration}
    \end{figure}

    \subsection{Velocity Verlet, a symplectic integrator}
    \label{sub:velocity_verlet_a_symplectic_integrator}

    In order to improve upon the fairly poor energy conservation of the
    Euler-Cromer method, we chose to implement the Velocity Verlet integrator.
    The algorithm is dependent on us having computed the forces for the first
    step and consists of three steps per time-step:
    \begin{align*}
        \vec{v}(t + \Delta t/2) &= \vec{v}(t) + \frac{\vec{F}(t)}{m}\frac{\Delta t}{2},\\
        \vec{r}(t + \Delta t) &= \vec{r}(t) + \vec{v}(t + \Delta t/2)\Delta t, \\
        \vec{v}(t + \Delta t) &= \vec{v}(t + \Delta t/2) + \frac{\vec{F}(t+\Delta t)}{m} \frac{\Delta t}{2}.
    \end{align*}
    We implement this algorithm in the \texttt{integrate} method in the
    \texttt{VelocityVerlet}-class. The method takes a system of atoms and a
    time step $\Delta t$ as arguments. We also introduce the private auxiliary
    variable \texttt{m\_first\_step}, which helps us takes care of the initial
    calculation of the forces.

    Of interest now is how our new method of integration compares with the old
    one. One way of looking at this is by examining the fluctuations in the
    total energy as functions of the time step $\Delta t$. This will be
    examined in upcoming sections.

    \subsection{Sampling quantities}
    \label{sub:sampling_quantities}

    We now have pretty much a working program, however we still need to sample
    the physical properties of the system at each time step. For this we turn
    to the \texttt{StatisticsSampler}-class where we will implement the methods
    for sampling the density, kinetic and potential energy and temperature of
    the system. The methods are named accordingly.

    We first start by considering the kinetic and potential energy of the
    system. We already have computed the potential energy, in the
    \texttt{LennardJones}-class, so we are now interested in computing the
    kinetic energy. The kinetic energy of an arbitrary atom $i$ is defined as
    $E_{k}^i = m_iv_i^2/2$, and so the total kinetic energy $E_k$ of the system
    is simply
    \begin{equation}
        \notag
        E_k = \sum^{n}_{i=1} E_k^i = \sum^{n}_{i=1} \frac{1}{2}m_iv_i^2,
    \end{equation}
    where $m_i$ and $v_i$ is the mass and velocity atom $i$ and $n$ is again
    the total number of atoms in the system.

    We are also interested in the instantaneous temperature of the system, and
    the equipartition theorem \cite{equipar} can help us with this. The theorem
    relates the average energy of a system to its temperaturkke and gives us the
    relation $\langle E_k \rangle = (3/2)k_bT$ which can be solved for
    temperature:
    \begin{equation}
        \notag
        T = \frac{2}{3} \frac{E_k}{nk_b}.
    \end{equation}
    Notice the removal of the brackets --- we are only interested in the
    \emph{instantaneous} temperature at any point in time. We also note that
    since our units are scaled, Boltzmann's constant is equal to one.
    The density of the system is the total mass of the system divided by the volume.

\listoffigures

\begin{thebibliography}{9}

\bibitem{min_image}
    \url{https://en.wikipedia.org/wiki/Periodic_boundary_conditions#Practical_implementation:_continuity_and_the_minimum_image_convention}
\bibitem{equipar}
    \url{https://en.wikipedia.org/wiki/Equipartition\_theorem}
\end{thebibliography}

\section{Simulations}
\label{sec:simulations}

\end{document}

### 2015-11-27 projects/project_5/molecular_dynamics/python_src/system.py

@@ -1,0 +1,37 @@
+import sys
+import matplotlib.pyplot as plt
+import numpy as np
+
+try:
+    filename = sys.argv[1]
+except:
+    pass
+
+
+kinetic_energy = []
+potential_energy = []
+temperature = []
+density = []
+
+with open(filename, 'r') as infile:
+    for line in infile:
+        line = [float(value) for value in line.split()]
+        temperature.append(line[0])
+        kinetic_energy.append(line[1])
+        potential_energy.append(line[2])
+        density.append(line[3])
+
+kinetic_energy = np.array(kinetic_energy)
+potential_energy = np.array(potential_energy)
+temperature = np.array(temperature)
+density = np.array(density)
+initial_temp = 300
+plt.plot(initial_temp/temperature, label='temperature')
+plt.legend()
+plt.show()
+plt.plot(kinetic_energy, label='kinetic')
+plt.legend()
+plt.show()
+plt.plot(potential_energy, label='potential')
+plt.legend()
+plt.show()

### 2015-11-27 projects/project_5/molecular_dynamics/article/article.tex
fixed the methods

@@
 \usepackage[]{hyperref} 
+\urlstyle{sf}
 \graphicspath{{../pictures/}}
@@
 
+    We need to chose the parameters $\sigma$ and $\varepsilon$ so that they are
+    suited for simulation of argon. It turns out that $\sigma = 3.405
+    \mathrm{\AA}$ and $\varepsilon = 119.8k_B\mathrm{K}$ with $k_B = 1$ in
+    MD-units.
+
     Argon has a melting point of $\approx 84$K and a boiling point of $\approx
@@
             in motion, but the system do behave more or less like a solid, a
-            liquid and a gas for respective temperatures. Any artifacts can be
-            contributed to numerical instability. It is possible to see a
+            liquid and a gas for respective temperatures. It is possible to see a
             certain change in the entropy of the system across the three
-            images. These made use of the Euler-Cromer integration method.}
+            images. These made use of the Velocity-Verlet integration method.}
         \label{fig:potential_iteration}
@@
     Notice the removal of the brackets --- we are only interested in the
-    \emph{instantaneous} temperature at any point in time. We also note that
+    \emph{instantaneous} temperature at any point in time. We also recall that
     since our units are scaled, Boltzmann's constant is equal to one.
-    The density of the system is the total mass of the system divided by the volume.
+
+\section{Simulations}
+\label{sec:simulations}
+
+In this section we consider various systems and simulate them using the program
+devised in the previous section.
+
+\subsection{Energy conservation of integration method}
+\label{sub:energy_conservation_of_integration_method}
+
+We initially decided to use the Velocity-Verlet integration method as opposed
+to Euler-Cromer for its good energy conservation. We now simulate a set of
+similar systems with varying temperature and look at how the fluctuations in
+total energy scales with the timestep $\Delta T$. In order to do this we can
+consider the standard-devation of the total energy.
+
+\subsection{Melting temperature of system}
+\label{sub:melting_temperature_of_system}
 
@@
 
-\section{Simulations}
-\label{sec:simulations}
-
 \end{document}

### 2015-11-27 projects/project_5/molecular_dynamics/python_src/system.py

@@ -25,8 +25,8 @@ kinetic_energy = np.array(kinetic_energy)
 potential_energy = np.array(potential_energy)
 temperature = np.array(temperature)
 density = np.array(density)
-initial_temp = 300
-plt.plot(initial_temp/temperature, label='temperature')
+initial_temp = 80
+plt.plot(temperature, label='temperature')
 plt.legend()
 plt.show()
 plt.plot(kinetic_energy, label='kinetic')

### 2015-11-27 projects/project_5/molecular_dynamics/python_src/system_class.py
working on python class

@@ -1,0 +1,35 @@
+import os
+class System:
+
+    """ This class represents the setup of a system with the following paramenters:
+    
+        T: initial temperature
+        N_each_dimension: number of unit cells in each dimension
+        b: lattice constant  
+    """
+
+    def __init__(self, T=300, N_each_dimension=5, b=5.26, ID = 1):
+        self.T = T
+        self.N_each_dimension = N_each_dimension
+        self.b = b
+        self.ID = ID
+        self.dir_name = '../data/%d' % ID
+        self.file_name = '%d_%04f_%04f.dat' % (N_each_dimension, T, b) + self.dir_name
+         
+    def run_simulations(self):
+        build_path = '/Users/ivar/Documents/University/build-molecular-dynamics-fys3150-Desktop_Qt_5_5_1_clang_64bit-Release/'
+        os.system(build_path + 'molecular-dynamics-fys3150 %d %f %f' % (self.N_each_dimension, self.T, self.b))
+        
+        try:
+            os.stat(self.dir_name)
+        except:
+            os.mkdir(self.dir_name)
+        
+        os.system('mv movie.xyz %s' % self.dir_name)
+        os.system('mv samples.dat %s/%s' % (self.dir_name, self.file_name))
+
+    def read_data(self):
+       pass    
+if __name__ == "__main__":
+    test = System(T=10, N_each_dimension=6, b=5.26)
+    test.run_simulations()

### 2015-11-27 projects/project_5/molecular_dynamics/python_src/system_class.py
Skeleton of a working framework done

@@ -8,13 +8,13 @@ class System:
         b: lattice constant  
     """
 
-    def __init__(self, T=300, N_each_dimension=5, b=5.26, ID = 1):
+    def __init__(self, T=300, N_each_dimension=5, b=5.26, ID = 1, simulation_description='example'):
         self.T = T
         self.N_each_dimension = N_each_dimension
         self.b = b
         self.ID = ID
-        self.dir_name = '../data/%d' % ID
-        self.file_name = '%d_%04f_%04f.dat' % (N_each_dimension, T, b) + self.dir_name
+        self.dir_name = '../data/%s/%d' % (simulation_description, ID)
+        self.file_name = self.dir_name + '/%d_%04f_%04f.dat' % (N_each_dimension, T, b)
          
     def run_simulations(self):
         build_path = '/Users/ivar/Documents/University/build-molecular-dynamics-fys3150-Desktop_Qt_5_5_1_clang_64bit-Release/'
@@ -23,13 +23,32 @@ class System:
         try:
             os.stat(self.dir_name)
         except:
-            os.mkdir(self.dir_name)
+            os.makedirs(self.dir_name)
         
         os.system('mv movie.xyz %s' % self.dir_name)
-        os.system('mv samples.dat %s/%s' % (self.dir_name, self.file_name))
+        os.system('mv samples.dat %s' % (self.file_name))
 
     def read_data(self):
-       pass    
+        temperature = []
+        potential = []
+        kinetic = []
+        total = []
+        
+        with open(self.file_name, 'r') as data_file:
+            for line in data_file:
+                line = [float(value) for value in line.split()]
+                temperature.append(line[0])
+                potential.append(line[1])
+                kinetic.append(line[2])
+                total.append(line[2] + line[1])
+
 if __name__ == "__main__":
-    test = System(T=10, N_each_dimension=6, b=5.26)
-    test.run_simulations()
+    """
+    Simulate a set of systems for various initial temperatures (task d)
+    """
+    simulations = []
+    for ID, T in enumerate(range(40, 200, 40)):
+        simulations.append(System(T, 5, 5.26, ID, 'varying_initial_temperatures'))
+    
+    for sim in simulations:
+        sim.read_data()

### 2015-11-28 projects/project_5/molecular_dynamics/article/article.tex
getting plots

@@
-\documentclass{article}
+\documentclass[a4paper]{article}
 \usepackage[]{enumerate} 
@@
 \usepackage[]{amsthm, amsmath, amssymb} 
+\usepackage[]{fancyhdr}
 \usepackage[]{hyperref} 
@@
 \maketitle
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\linewidth]{ch.pdf}
+\end{figure}
+
+\begin{abstract}
+\end{abstract}
 \tableofcontents
@@
     the equipartition theorem \cite{equipar} can help us with this. The theorem
-    relates the average energy of a system to its temperaturkke and gives us the
+    relates the average energy of a system to its temperature and gives us the
     relation $\langle E_k \rangle = (3/2)k_bT$ which can be solved for
@@
 
-We initially decided to use the Velocity-Verlet integration method as opposed
-to Euler-Cromer for its good energy conservation. We now simulate a set of
-similar systems with varying temperature and look at how the fluctuations in
-total energy scales with the timestep $\Delta T$. In order to do this we can
-consider the standard-devation of the total energy.
+We initially decided to use the Velocity-Verlet integration method instead of
+Euler-Cromer because of its good energy conservation. It is therefore of
+interest to examine exactly what kind of impact this change of integration
+method has for our simulations.  We now simulate a set of similar systems with
+varying temperature and look at how the fluctuations in total energy scales
+with the timestep $\Delta T$. In order to do this we can consider the
+standard-devation of the total energy.
 

### 2015-11-28 projects/project_5/molecular_dynamics/python_src/system.py

@@ -1,37 +1,0 @@
-import sys
-import matplotlib.pyplot as plt
-import numpy as np
-
-try:
-    filename = sys.argv[1]
-except:
-    pass
-
-
-kinetic_energy = []
-potential_energy = []
-temperature = []
-density = []
-
-with open(filename, 'r') as infile:
-    for line in infile:
-        line = [float(value) for value in line.split()]
-        temperature.append(line[0])
-        kinetic_energy.append(line[1])
-        potential_energy.append(line[2])
-        density.append(line[3])
-
-kinetic_energy = np.array(kinetic_energy)
-potential_energy = np.array(potential_energy)
-temperature = np.array(temperature)
-density = np.array(density)
-initial_temp = 80
-plt.plot(temperature, label='temperature')
-plt.legend()
-plt.show()
-plt.plot(kinetic_energy, label='kinetic')
-plt.legend()
-plt.show()
-plt.plot(potential_energy, label='potential')
-plt.legend()
-plt.show()

### 2015-11-28 projects/project_5/molecular_dynamics/python_src/system_class.py

@@ -1,4 +1,5 @@
 import os
+import matplotlib.pyplot as plt
 class System:
 
     """ This class represents the setup of a system with the following paramenters:
@@ -29,18 +30,18 @@ class System:
         os.system('mv samples.dat %s' % (self.file_name))
 
     def read_data(self):
-        temperature = []
-        potential = []
-        kinetic = []
-        total = []
+        self.temperature = []
+        self.potential = []
+        self.kinetic = []
+        self.total = []
         
         with open(self.file_name, 'r') as data_file:
             for line in data_file:
                 line = [float(value) for value in line.split()]
-                temperature.append(line[0])
-                potential.append(line[1])
-                kinetic.append(line[2])
-                total.append(line[2] + line[1])
+                self.temperature.append(line[0])
+                self.potential.append(line[1])
+                self.kinetic.append(line[2])
+                self.total.append(line[2] + line[1])
 
 if __name__ == "__main__":
     """
@@ -52,3 +53,23 @@ if __name__ == "__main__":
     
     for sim in simulations:
         sim.read_data()
+    
+    plt.title('Energy conservation')
+    for sim in simulations:
+        plt.subplot(311)
+        plt.plot(sim.kinetic, label='$T = %0.2f$' % sim.T)
+        plt.xlabel='$t$'
+        plt.ylabel='$E_k [eV]$'
+        plt.legend(loc='best')
+        plt.subplot(312)
+        plt.plot(sim.potential, label='$T = %0.2f$' % sim.T)
+        plt.xlabel='$t$'
+        plt.ylabel='$U [eV]$'
+        plt.legend(loc='best')
+        plt.subplot(313)
+        plt.plot(sim.total, label='$T = %0.2f$' % sim.T)
+        plt.xlabel='$t$'
+        plt.ylabel='$E_{\\mathrm{tot}} [eV]$'
+        plt.legend(loc='best')
+    plt.grid('on')
+    plt.show()

### 2015-11-28 projects/project_5/molecular_dynamics/article/article.tex
working on energy fluctuations

@@
             conservation of energy.
-        \item Write functions for sampling and storing physical quantities
+        \item Write functions for sampling and storing physical quantities.
+        \item Miscellaneous changes.
     \end{enumerate}
@@
     yet. This can be seen in figure \ref{fig:first_iteration}.
+
     \begin{figure}
@@
 
+    \subsection{Miscellaneous changes}
+    \label{sub:miscellaneous_changes}
+
+    In order to ease the implementation of the python framework used to run the
+    simulations, we need to make some auxiliary changes to the
+    \texttt{main}-method of the program. First of all, we are interested in
+    comparing the two integration methods the program currently supports, so we
+    add a fourth possible command line argument for setting the integration
+    method. The program defaults to Velocity-Verlet.
+
+    We also need a way of changing the time step $\Delta t$ used in the
+    simulations. This is also implemented as a command line argument, and we
+    use a default time step of $\Delta t = 1.0 \times 10^{-15}$ seconds.
+
     \section{Simulations}
@@
 varying temperature and look at how the fluctuations in total energy scales
-with the timestep $\Delta T$. In order to do this we can consider the
-standard-devation of the total energy.
+with the time step $\Delta T$. In order to do this we can consider the
+standard-deviation of the total energy.
 
+We are going to run the simulations for a starting temperature of 300 kelvin,
+and vary the time step from $1.0 \times 10^{-16}$ to $1.0 \times 10^{-13}$
+seconds.
 \subsection{Melting temperature of system}

### 2015-11-28 projects/project_5/molecular_dynamics/python_src/system_class.py

@@ -17,9 +17,9 @@ class System:
         self.dir_name = '../data/%s/%d' % (simulation_description, ID)
         self.file_name = self.dir_name + '/%d_%04f_%04f.dat' % (N_each_dimension, T, b)
          
-    def run_simulations(self):
+    def run_simulations(self, integration_method=2):
         build_path = '/Users/ivar/Documents/University/build-molecular-dynamics-fys3150-Desktop_Qt_5_5_1_clang_64bit-Release/'
-        os.system(build_path + 'molecular-dynamics-fys3150 %d %f %f' % (self.N_each_dimension, self.T, self.b))
+        os.system(build_path + 'molecular-dynamics-fys3150 %d %f %f %d' % (self.N_each_dimension, self.T, self.b, integration_method))
         
         try:
             os.stat(self.dir_name)
@@ -47,15 +47,20 @@ if __name__ == "__main__":
     """
     Simulate a set of systems for various initial temperatures (task d)
     """
-    simulations = []
-    for ID, T in enumerate(range(40, 200, 40)):
-        simulations.append(System(T, 5, 5.26, ID, 'varying_initial_temperatures'))
+    simulations_euler = []
+    simulations_verlet = []
+    for ID, T in enumerate(range(40, 200, 10)):
+        simulations_euler.append(System(T, 5, 5.26, ID, 'varying_initial_temperatures_euler'))
+        simulations_verlet.append(System(T, 5, 5.26, ID, 'varying_initial_temperatures_verlet'))
     
-    for sim in simulations:
-        sim.read_data()
+    for sim_e, sim_v in zip(simulations_euler, simulations_verlet):
+        sim_e.run_simulations(integration_method=1)
+        sim_v.run_simulations(integration_method=2)
+        sim_e.read_data()
+        sim_v.read_data()
     
     plt.title('Energy conservation')
-    for sim in simulations:
+    for sim in simulations_euler:
         plt.subplot(311)
         plt.plot(sim.kinetic, label='$T = %0.2f$' % sim.T)
         plt.xlabel='$t$'

### 2015-11-28 projects/project_5/molecular_dynamics/article/article.tex
energy fluctuation comparison done

@@
 We are going to run the simulations for a starting temperature of 300 kelvin,
-and vary the time step from $1.0 \times 10^{-16}$ to $1.0 \times 10^{-13}$
-seconds.
+and vary the time step from $1.0 \times 10^{-15}$ to $1.0 \times 10^{-14}$
+seconds with 40 intermediate values. The results, as shown in figure
+\ref{fig:energy_fluctuations}, tells us exactly how much better Velocity Verlet
+is when it comes to energy conservation.
+
+\begin{figure}[h]
+    \centering \includegraphics[width=0.8\linewidth]{energy_fluctuations.pdf}
+    \caption[Energy fluctuations]{The standard deviation of the total energy as
+        a function of the time-step $\Delta t$ with the energy being per atom.
+        This clearly shows the benefits of using Velocity-Verlet over
+    Euler-Cromer. While Euler-Cromer gives rise to fluctuations that scale
+somewhat linearly in the time step $\Delta t$, the fluctuations attributed to
+Velocity-Verlet is negligible in comparison.}
+    \label{fig:energy_fluctuations}
+\end{figure}
+
+\subsection{Initial to final temperature}
+\label{sub:initial_to_final_temperature}
+
+We are now interested in determining what initial temperature $T_i$ is needed
+in order to have a final temperature of $T$.  One way of examining this is by
+looking at the ratio $T/T_i$. 
 \subsection{Melting temperature of system}

### 2015-11-28 projects/project_5/molecular_dynamics/python_src/system_class.py

@@ -1,5 +1,6 @@
 import os
 import matplotlib.pyplot as plt
+import numpy as np
 class System:
 
     """ This class represents the setup of a system with the following paramenters:
@@ -9,17 +10,18 @@ class System:
         b: lattice constant  
     """
 
-    def __init__(self, T=300, N_each_dimension=5, b=5.26, ID = 1, simulation_description='example'):
+    def __init__(self, T=300, dt=1.0e-15, N_each_dimension=5, b=5.26, ID = 1, simulation_description='example'):
         self.T = T
         self.N_each_dimension = N_each_dimension
         self.b = b
         self.ID = ID
+        self.dt = dt
         self.dir_name = '../data/%s/%d' % (simulation_description, ID)
         self.file_name = self.dir_name + '/%d_%04f_%04f.dat' % (N_each_dimension, T, b)
          
     def run_simulations(self, integration_method=2):
         build_path = '/Users/ivar/Documents/University/build-molecular-dynamics-fys3150-Desktop_Qt_5_5_1_clang_64bit-Release/'
-        os.system(build_path + 'molecular-dynamics-fys3150 %d %f %f %d' % (self.N_each_dimension, self.T, self.b, integration_method))
+        os.system(build_path + 'molecular-dynamics-fys3150 %d %f %f %d %.10g' % (self.N_each_dimension, self.T, self.b, integration_method, self.dt))
         
         try:
             os.stat(self.dir_name)
@@ -45,36 +47,42 @@ class System:
 
 if __name__ == "__main__":
     """
-    Simulate a set of systems for various initial temperatures (task d)
+    TASK D
     """
-    simulations_euler = []
-    simulations_verlet = []
-    for ID, T in enumerate(range(40, 200, 10)):
-        simulations_euler.append(System(T, 5, 5.26, ID, 'varying_initial_temperatures_euler'))
-        simulations_verlet.append(System(T, 5, 5.26, ID, 'varying_initial_temperatures_verlet'))
+    simulations = []
+    initial_temp = 300 
+    b = 5.26
+    N = 5
+    description = 'energy_fluctuations_'
     
-    for sim_e, sim_v in zip(simulations_euler, simulations_verlet):
-        sim_e.run_simulations(integration_method=1)
-        sim_v.run_simulations(integration_method=2)
[…]

### 2015-11-28 projects/project_5/molecular_dynamics/article/article.tex
on diffusion now

@@
 in order to have a final temperature of $T$.  One way of examining this is by
-looking at the ratio $T/T_i$. 
+looking at the ratio $T/T_i$. As we can see in \ref{fig:temperature_ratio} the
+ratio stabilizes around $T/T_i = 0.5$. Solving this for $T_i$ tells us that if
+we are interested in a final temperature of $T$, then we should chose our
+initial temperature to be
+\begin{equation}
+    \notag
+    T_i = 2T.
+\end{equation}
+
+As we see from the plot, the temperature drastically drops just after the
+initial state. This is because $T \propto E_k$, i.e., the temperature is
+proportional to the kinetic energy, and since the potential has not yet kicked
+in, the kinetic energy is at its maximum. While what we have just observed is
+an empirical result, we can alternatively look at the virial theorem
+\cite{virial}, which relates the time average of the kinetic energy to the time
+average of the potential energy. This gives us a description of this
+stabilization independent of the temperature.
+
+\begin{figure}[h]
+    \centering \includegraphics[width=0.8\linewidth]{temperature_ratio.pdf}
+    \caption[Temperature ratio]{Plotting the ratio between initial and
+        instantaneous temperature for varying $T_i$. Here the initial
+        temperatures ranges from 20 to 500 kelvin, with 48 intermediate steps.
+    As we can see the ratio stabilizes around 0.5. The graph of an arbitrary
+temperature has been highlighted in black.}
+    \label{fig:temperature_ratio}
+\end{figure}
+
 \subsection{Melting temperature of system}
@@
     \url{https://en.wikipedia.org/wiki/Equipartition\_theorem}
+\bibitem{virial}
+    \url{https://en.wikipedia.org/wiki/Virial_theorem#Time_averaging}
 \end{thebibliography}

### 2015-11-28 projects/project_5/molecular_dynamics/python_src/simulations.py

@@ -1,0 +1,84 @@
+from system_class import System
+import os
+import matplotlib.pyplot as plt
+import numpy as np
+
+def taskE(run_simulations=False):
+    """
+    Performs the simulations needed to compare the two
+    integration methods Velocity Verlet and Euler-Cromer.
+    """
+    simulations = []
+    initial_temp = 300 
+    b = 5.26
+    N = 5
+    description = 'energy_fluctuations_'
+    
+    dt_values = 10**np.linspace(-15, -14, 40)
+    for i, dt in enumerate(dt_values):
+        
+        euler = System(T = initial_temp, dt=dt, N_each_dimension=5, b=5.26, ID=i, simulation_description=description+'euler')
+        verlet = System(T = initial_temp, dt=dt, N_each_dimension=5, b=5.26, ID=i, simulation_description=description+'verlet')
+        
+        simulations.append([euler, verlet])
+    
+    # run simulations 
+
+    if run_simulations:
+        for sim in simulations:
+            sim[0].run_simulations(1) # euler
+            sim[1].run_simulations(2) # verlet
+    
+    # extract data
+    euler_sd = []
+    verlet_sd = []
+    for sim in simulations:
+        sim[0].read_data()
+        sim[1].read_data()
+    
+        euler_sd.append(np.std(sim[0].total))
+        verlet_sd.append(np.std(sim[1].total))
+
+    # plot data
+    plt.plot(dt_values, euler_sd, 'ro-', dt_values, verlet_sd, 'go-')
+    plt.legend(['Euler-Cromer', 'Velocity-Verlet'], loc='best')
+    plt.xlabel('$\\Delta t [s]$')
+    plt.ylabel('$\\sigma_E/N_{\\mathrm{atoms}} [eV]$')
+    plt.grid('on')
+    plt.savefig('energy_fluctuations.pdf')
+
+def taskF(run_simulations=False):
+    """
+    This method runs simulations for a system with a fixed
+    lattice constant, fixed number of unit cells, but for
+    varying initial temperature. We are interested in plotting
+    the ratio between the instantaneous temperature and the
+    initial temperature.
+    """
+
+    simulations = []
+    
+    T_values = range(20, 500, 10) 
+    b = 5.26
+    N = 5
+    description = 'temperature_ratios' 
+    for i, T_i in enumerate(T_values):
[…]

### 2015-11-28 projects/project_5/molecular_dynamics/python_src/system_class.py

@@ -44,45 +44,7 @@ class System:
                 self.potential.append(line[1])
                 self.kinetic.append(line[2])
                 self.total.append(line[2] + line[1])
-
-if __name__ == "__main__":
-    """
-    TASK D
-    """
-    simulations = []
-    initial_temp = 300 
-    b = 5.26
-    N = 5
-    description = 'energy_fluctuations_'
-    
-    dt_values = 10**np.linspace(-15, -14, 40)
-    for i, dt in enumerate(dt_values):
-        
-        euler = System(T = initial_temp, dt=dt, N_each_dimension=5, b=5.26, ID=i, simulation_description=description+'euler')
-        verlet = System(T = initial_temp, dt=dt, N_each_dimension=5, b=5.26, ID=i, simulation_description=description+'verlet')
-        
-        simulations.append([euler, verlet])
-    
-    # run simulations 
-    for sim in simulations:
-        print sim[0].ID
-        #sim[0].run_simulations(1) # euler
-        #sim[1].run_simulations(2) # verlet
-    
-    # extract data
-    euler_sd = []
-    verlet_sd = []
-    for sim in simulations:
-        sim[0].read_data()
-        sim[1].read_data()
-    
-        euler_sd.append(np.std(sim[0].total))
-        verlet_sd.append(np.std(sim[1].total))
-
-    # plot data
-    plt.plot(dt_values, euler_sd, 'ro-', dt_values, verlet_sd, 'go-')
-    plt.legend(['Euler-Cromer', 'Velocity-Verlet'], loc='best')
-    plt.xlabel('$\\Delta t [s]$')
-    plt.ylabel('$\\sigma_E/N_{\\mathrm{atoms}} [eV]$')
-    plt.grid('on')
-    plt.savefig('energy_fluctuations.pdf')
+        self.temperature = np.array(self.temperature)
+        self.potential = np.array(self.potential)
+        self.kinetic = np.array(self.kinetic)
+        self.total = np.array(self.total)

### 2015-11-28 projects/project_5/molecular_dynamics/article/article.tex
getting started on diffusion

@@
 
+    In order to compute the \emph{mean square displacement} in section
+    \ref{sub:melting_temperature_of_system} we need to keep track of the
+    initial position of each atom. We therefore introduce a variable
+    \texttt{m\_initialPosition} in the \texttt{Atom-class} which is set during
+    initialization.  When computing the diffusion constant $D$, we need to make
+    sure that $r_i^2(t)$ is the true displacement, and not the displacement
+    relative to the nearest periodic image. This will be implemented in the
+    \texttt{applyBoundaryConditions}-method.
+
 \section{Simulations}
@@
 
+In order to measure the melting temperature of the system, we can use the Einstein relation which states the following:
+\begin{equation}
+    \notag
+    \langle r^2(t) \rangle = 6Dt.
+\end{equation}
+$D$ is here the diffusion constant, $t$ is time and the \emph{mean square displacement} is calculated as
+\begin{equation}
+    \notag
+    r_i^2(t) = \left| \vec{r}_i(t) - \vec{r}_i(0) \right|^2.
+\end{equation}
 \listoffigures

### 2015-11-29 projects/project_5/molecular_dynamics/article/article.tex
fixed some typos and wonky sentences

@@
 \begin{abstract}
+    In this project we explore the world of molecular dynamics. We are first
+    presented with the skeleton of a working MD-program, and we are asked to
+    implement the functionality we need as we go along. We look at the time
+    evolution of a system of argon, with an initial face-centered cubic lattice
+    configuration. The atoms are under the influence of the Lennard-Jones
+    potential. We use the Velocity-Verlet as our time integrator of choice, and
+    we compare this to the Euler-Cromer method in terms of energy conservation
+    properties. We also compute the melting temperature of argon, by means of
+    the Einstein relation and diffusion constant.
 \end{abstract}
 \tableofcontents
+\listoffigures
+
 \section{Introduction}
@@
 
+    The construction of the program will be covered in section
+    \ref{sec:building_a_working_md_program}, where some of the motivation
+    behind each addition to the program will be discussed. In section
+    \ref{sec:simulations} we employ our newly implemented program and look at
+    some specific systems.
+
 \section{Building a working MD-program}
@@
     the steps that went into the implementation of the program, from skeleton
-    to working program. The process can be summarized as
+    to working program. The process can be summarized as following:
 
@@
     We want our system to conform to periodic boundary conditions. The reasons
-    for doing this is that to a certain extent the system of finite size more
-    closely resembles that of an infinite size. It is important that the
+    for doing this is that a finite system with periodic boundary condition
+    more closely resembles that of an infinite size. It is important that the
     bounding box which determines the boundary of the system is of sufficient
@@
         \notag
-        P(v_i)\, dv_i = \left( \frac{m_i}{2\pi k_BT} \right)^{1/2} e
+        P(v_i)\, dv_i = \left( \frac{m_i}{2\pi k_BT} \right)^{1/2} e^{-\frac{mv_i^2}{2k_BT}}dv_i
     \end{equation}
@@
     time step $\Delta t$ as arguments. We also introduce the private auxiliary
-    variable \texttt{m\_first\_step}, which helps us takes care of the initial
+    variable \texttt{m\_first\_step}, which helps us take care of the initial
     calculation of the forces.
@@
     relates the average energy of a system to its temperature and gives us the
-    relation $\langle E_k \rangle = (3/2)k_bT$ which can be solved for
+    relation $\langle E_k \rangle = (3/2)k_BT$ which can be solved for
     temperature:
@@
         \notag
-        T = \frac{2}{3} \frac{E_k}{nk_b}.
+        T = \frac{2}{3} \frac{E_k}{nk_B}.
     \end{equation}
@@
 interest to examine exactly what kind of impact this change of integration
-method has for our simulations.  We now simulate a set of similar systems with
-varying temperature and look at how the fluctuations in total energy scales
-with the time step $\Delta T$. In order to do this we can consider the
-standard-deviation of the total energy.
+method has for our simulations.  We now simulate a set of similar systems and
+look at how the fluctuations in total energy scales with the time step $\Delta
+T$. In order to do this we can consider the standard-deviation of the total
+energy.
 
@@
 in order to have a final temperature of $T$.  One way of examining this is by
-looking at the ratio $T/T_i$. As we can see in \ref{fig:temperature_ratio} the
+looking at the ratio $T/T_i$. As we can see in figure \ref{fig:temperature_ratio} the
 ratio stabilizes around $T/T_i = 0.5$. Solving this for $T_i$ tells us that if
@@
 \begin{figure}[h]
-    \centering \includegraphics[width=0.8\linewidth]{temperature_ratio.pdf}
+    \centering \includegraphics[width=\linewidth]{temperature_ratio.pdf}
     \caption[Temperature ratio]{Plotting the ratio between initial and
@@
     \notag
-    r_i^2(t) = \left| \vec{r}_i(t) - \vec{r}_i(0) \right|^2.
+    r_i^2(t) = \left| \vec{r}_i(t) - \vec{r}_i(0) \right|^2
 \end{equation}
-\listoffigures
+for atom $i$.
+
+\clearpage
 
 \begin{thebibliography}{9}
-
 \bibitem{min_image}
@@
 \bibitem{virial}
-    \url{https://en.wikipedia.org/wiki/Virial_theorem#Time_averaging}
+    \url{https://en.wikipedia.org/wiki/Virial_theorem}
 \end{thebibliography}
-
 \end{document}

### 2015-11-30 projects/project_5/molecular_dynamics/article/article.tex
fixed a reference
fixed src indentation
blabla

@@
     total energy as functions of the time step $\Delta t$. This will be
-    examined in upcoming sections.
+    examined in section \ref{sub:energy_conservation_of_integration_method}.
 
@@
     relative to the nearest periodic image. This will be implemented in the
-    \texttt{applyBoundaryConditions}-method.
+    \texttt{applyBoundaryConditions}-method and is as simple as in addition to
+    moving the position of the atom, also move the initial position of the atom
+    by the same amount, because then the relative distance between instantaneous
+    position and initial position is equal to the true displacement.
 
@@
 
-In order to measure the melting temperature of the system, we can use the Einstein relation which states the following:
+    In order to measure the melting temperature of the system, we can use the
+    Einstein relation which states the following:
     \begin{equation}
@@
     \end{equation}
-$D$ is here the diffusion constant, $t$ is time and the \emph{mean square displacement} is calculated as
+    $D$ is here the diffusion constant, $t$ is time and the \emph{mean square
+    displacement} is calculated as
     \begin{equation}
@@
     \end{equation}
-for atom $i$.
+    for atom $i$. Solving this for $D$ gives yields
+    \begin{equation}
+        \notag
+        D = \frac{\langle r^2(t) \rangle}{6t}.
+    \end{equation}
+    We can now make set of simulations for varying initial temperature $T$.
 

### 2015-11-30 projects/project_5/molecular_dynamics/python_src/simulations.py

@@ -81,4 +81,32 @@ def taskF(run_simulations=False):
     plt.ylabel('$T / T_i$')
     plt.savefig('temperature_ratio.pdf')
 
-taskF(False)
+def taskG(run_simulations=False):
+    """
+    This method runs a set of simulations for varying initial temperature T
+    and looks at the diffusion constant. This is in order to measure the melting
+    point of the system.
+    """
+
+    simulations = []
+
+    T_values = range(400, 2000, 100)
+    b = 5.26
+    N = 10
+    description = 'diffusion_constant'
+    for i, T_i in enumerate(T_values):
+        simulations.append(System(T = T_i, b=5.26, N_each_dimension=N, ID=i, simulation_description=description))
+
+    if run_simulations:
+        for sim in simulations:
+            sim.run_simulations()
+
+    for sim in simulations:
+        sim.read_data()
+        plt.plot(sim.diffusion)
+    plt.xlabel('$t$')
+    plt.ylabel('$D = \\langle r^2(t) \\rangle / 6t$')
+    plt.grid('on')
+    plt.savefig('diffusion_constant.pdf')
+
+taskG(True)

### 2015-11-30 projects/project_5/molecular_dynamics/python_src/system_class.py

@@ -10,7 +10,7 @@ class System:
         b: lattice constant  
     """
 
-    def __init__(self, T=300, dt=1.0e-15, N_each_dimension=5, b=5.26, ID = 1, simulation_description='example'):
+    def __init__(self, T=300, dt=5.0e-15, N_each_dimension=5, b=5.26, ID = 1, simulation_description='example'):
         self.T = T
         self.N_each_dimension = N_each_dimension
         self.b = b
@@ -36,6 +36,7 @@ class System:
         self.potential = []
         self.kinetic = []
         self.total = []
+        self.diffusion = []
         
         with open(self.file_name, 'r') as data_file:
             for line in data_file:
@@ -44,7 +45,9 @@ class System:
                 self.potential.append(line[1])
                 self.kinetic.append(line[2])
                 self.total.append(line[2] + line[1])
+                self.diffusion.append(line[4])
         self.temperature = np.array(self.temperature)
         self.potential = np.array(self.potential)
         self.kinetic = np.array(self.kinetic)
         self.total = np.array(self.total)
+        self.diffusion = np.array(self.diffusion)

### 2015-12-03 projects/project_5/molecular_dynamics/python_src/simulations.py
diffusion constant done

@@ -83,16 +83,16 @@ def taskF(run_simulations=False):
 
 def taskG(run_simulations=False):
     """
-    This method runs a set of simulations for varying initial temperature T
-    and looks at the diffusion constant. This is in order to measure the melting
-    point of the system.
+    This method runs a set of simulations for varying initial temperature T and
+    looks at the diffusion constant. This is in order to measure the melting
+    point of the system. We pick the final D of the simulation.
     """
 
     simulations = []
 
-    T_values = range(400, 2000, 100)
+    T_values = range(400, 800, 5)
     b = 5.26
-    N = 10
+    N = 6
     description = 'diffusion_constant'
     for i, T_i in enumerate(T_values):
         simulations.append(System(T = T_i, b=5.26, N_each_dimension=N, ID=i, simulation_description=description))
@@ -100,13 +100,31 @@ def taskG(run_simulations=False):
     if run_simulations:
         for sim in simulations:
             sim.run_simulations()
-
+    
+    diffusion = []
     for sim in simulations:
         sim.read_data()
-        plt.plot(sim.diffusion)
-    plt.xlabel('$t$')
-    plt.ylabel('$D = \\langle r^2(t) \\rangle / 6t$')
+        diffusion.append(sim.diffusion[-1])
+
+    diffusion = np.array(diffusion)
+    T_values = np.array(T_values)
+
+    T_before_melt = np.nonzero(T_values < 620)
+    T_after_melt = np.nonzero(T_values >= 615)
+    
+    print T_values[T_before_melt][-1]
+    A = np.vstack([T_values[T_before_melt], np.ones(len(T_values[T_before_melt]))]).T
+    m, c = np.linalg.lstsq(A, diffusion[T_before_melt])[0]
+    plt.plot(T_values[T_before_melt], m*T_values[T_before_melt] + c)
+
+    A = np.vstack([T_values[T_after_melt], np.ones(len(T_values[T_after_melt]))]).T
+    m, c = np.linalg.lstsq(A, diffusion[T_after_melt])[0]
+    plt.plot(T_values[T_after_melt], m*T_values[T_after_melt] + c, )
+
+    plt.plot(T_values, diffusion, 'bo', alpha=0.5)
+    plt.xlabel('$t \\quad [K]$')
+    plt.ylabel('$D = \\langle r^2(t) \\rangle / 6t \\quad [m^2/s]$')
     plt.grid('on')
     plt.savefig('diffusion_constant.pdf')
 
-taskG(True)
+taskG(False)

### 2015-12-03 projects/project_5/molecular_dynamics/article/article.tex
diffusion constant done
cell lists explained

@@
     complexity of $\mathcal{O}(n^2)$. Luckily more efficient methods are
-    available, most notably is the Verlet list method.
+    available, most notably is the Cell list method.
+
+    \subsubsection{Cell lists}
+    \label{ssub:cell_lists}
+
+    The main idea behind cell lists is that at some fixed distance, the force
+    induced by the potential between two atoms is negligible and there is
+    therefore no need for computing it. If we for each atom $i$ keep track of
+    all atoms within this fixed distance, which we will call $r_\mathrm{cut}$,
+    at any given time step, we can greatly reduce the number of computations
+    needed.
+
+    We divide the domain of our system into \emph{cells}, each with an edge
+    length of at least the cut-of radius of the potential. This ensures that if
+    we only compute the mutual forces between atoms in the same or adjacent
+    cells, then the atoms are never a distance farther than $r_\mathrm{cut}$
+    apart. For each integration step we check the validity of the list of
+    neighbors for each atom by iterating through each atom in the system, and
+    checking whether it has moved a distance larger than $r_\mathrm{cut}$ since
+    the last time the list of neighbors was built.
+
+    We additionally need to rescale our potential so that it takes into
+    consideration the cut-off radius. We define the scaled potential $U' =
+    U(r_{ij}) - U(r_\mathrm{cut})$ when $r_{ij} < r_\mathrm{cut}$ and 0
+    otherwise.
+
+    There is a quite subtle consideration which is often overlooked, and that
+    is the fact that in order to make use of cell-lists, we need a sufficient
+    number of unit cells in each dimensions. Otherwise, there is not enough
+    room for our cells.  In the program, if the number of unit cells is less
+    than or equal to 5, we employ our old potential. If it is larger, we use
+    the new scaled potential with cell lists.
+
+    By implementing cell-lists we reduce the computation of the forces from
+    $\mathcal{O}(N^2)$ to $\mathcal{O}(N)$.
+
+    \subsubsection{Choice of parameters}
+    \label{ssub:choice_of_parameters}
 
@@
     \end{equation}
-    We can now make set of simulations for varying initial temperature $T$.
+    We can now make set of simulations for varying initial temperature $T_i$
+    and store the final value for the diffusion constant for each final
+    temperature $T$. The results are shown in figure
+    \ref{fig:diffusion_constant} and tells us that the system melts around a
+    temperature of $T = 0.5T_i$, which here amounts to a temperature of about
+    $304K$.
 
+    \begin{figure}[h]
+        \centering
+        \includegraphics[width=0.8\linewidth]{diffusion_constant.pdf}
+        \caption[Diffusion constant]{The diffusion constant as a function of temperature. We here
+                see that $D$ abruptly increases at the temperature of $T = 304K$
+                which indicates that the crystal is melting.  The two lines are the
+                least square solutions to the noisy data before and after,
+                respectively. The diffusion constant is given in units of area per second.}
+        \label{fig:diffusion_constant}
+    \end{figure}
 \clearpage

### 2015-12-03 projects/project_5/molecular_dynamics/python_src/simulations.py

@@ -90,12 +90,12 @@ def taskG(run_simulations=False):
 
     simulations = []
 
-    T_values = range(400, 800, 5)
+    T_values = range(400, 1000, 5)
     b = 5.26
     N = 6
     description = 'diffusion_constant'
     for i, T_i in enumerate(T_values):
-        simulations.append(System(T = T_i, b=5.26, N_each_dimension=N, ID=i, simulation_description=description))
+        simulations.append(System(T = T_i, b=b, N_each_dimension=N, ID=i, simulation_description=description))
 
     if run_simulations:
         for sim in simulations:
@@ -109,20 +109,20 @@ def taskG(run_simulations=False):
     diffusion = np.array(diffusion)
     T_values = np.array(T_values)
 
-    T_before_melt = np.nonzero(T_values < 620)
-    T_after_melt = np.nonzero(T_values >= 615)
+    T_before_melt = np.nonzero(T_values < 600)
+    T_after_melt = np.nonzero(T_values >= 600)
     
     print T_values[T_before_melt][-1]
-    A = np.vstack([T_values[T_before_melt], np.ones(len(T_values[T_before_melt]))]).T
+    A = np.vstack([T_values[T_before_melt]/2.0, np.ones(len(T_values[T_before_melt]))]).T
     m, c = np.linalg.lstsq(A, diffusion[T_before_melt])[0]
-    plt.plot(T_values[T_before_melt], m*T_values[T_before_melt] + c)
+    plt.plot(T_values[T_before_melt]/2.0, m*T_values[T_before_melt]/2.0 + c, label='$D$ before melting')
 
-    A = np.vstack([T_values[T_after_melt], np.ones(len(T_values[T_after_melt]))]).T
+    A = np.vstack([T_values[T_after_melt]/2.0, np.ones(len(T_values[T_after_melt]))]).T
     m, c = np.linalg.lstsq(A, diffusion[T_after_melt])[0]
-    plt.plot(T_values[T_after_melt], m*T_values[T_after_melt] + c, )
-
-    plt.plot(T_values, diffusion, 'bo', alpha=0.5)
-    plt.xlabel('$t \\quad [K]$')
+    plt.plot(T_values[T_after_melt]/2.0, m*T_values[T_after_melt]/2.0 + c, label='$D$ after melting')
+    plt.plot(T_values/2.0, diffusion, 'bo', alpha=0.5)
+    plt.legend(loc='best')
[…]

### 2015-12-04 projects/project_5/molecular_dynamics/article/article.tex
fixed units on temperature ratio

@@
 \usepackage[]{fancyhdr}
+\usepackage[]{setspace} 
 \usepackage[]{hyperref} 
@@
 \end{abstract}
+\doublespacing
 \tableofcontents
 \listoffigures      
+\singlespacing
 
@@
 
+    As shown in figure \ref{fig:maxwell_boltzmann}, increasing the temperature
+    yields a more even velocity distribution. 
     The problem that arises when using the Maxwell-Boltzmann distribution is
@@
 
+    \begin{figure}[h]
+        \centering \includegraphics[width=0.7\linewidth]{maxwell_boltzmann.pdf}
+        \caption[The Maxwell-Boltzmann distribution]{The Maxwell-Boltzmann
+            distribution for a varying scale parameter $a = k_BT/m$.  As we can
+            see, increasing the temperature increases the likelihood of an atom
+        being assigned an arbitrary velocity, where as for low temperatures the
+    distribution is biased towards one specific velocity.}
+        \label{fig:maxwell_boltzmann}
+    \end{figure}
     We compute the velocity of the center of mass of the system, and then
@@
     checking whether it has moved a distance larger than $r_\mathrm{cut}$ since
-    the last time the list of neighbors was built.
+    the last time the list of neighbors was built. It is interesting to note
+    that we have to rebuild the neighbor lists more often for higher
+    temperature due to the velocities being proportional to the temperature.
 
@@
     Argon has a melting point of $\approx 84$K and a boiling point of $\approx
-    87$K. If we run the simulations, once for a temperature of $40$K, once for
-    $85$K and finally once for $120K$ we should see indications of three
-    different phases. The results shows that the model somewhat accurately
-    manages to capture these phase transitions. Snapshots taken at the same
-    point in time for all three cases are presented in figure
+    87$K. If we run the simulations, once for a temperature of $80$K, once for
+    $170$K and finally once for $240K$ we should see indications of three
+    different phases (see section \ref{sub:initial_to_final_temperature} for
+    the choices of these temperatures). The results shows that the model does
+    not manage to capture these phase transitions. This is probably attributed
+    to the Lennard-Jones potential. If we run the simulations for initial
+    temperatures of $400K, 700K$ and $900K$ we should see indications of phase
+    change. We will get back to this in section
+    \ref{sub:melting_temperature_of_system}. Snapshots taken at the same point
+    in time for all three cases are presented in figure
     \ref{fig:potential_iteration}.
@@
                 \includegraphics[height=100pt, width=\linewidth]{solid_40K.png}
-                \caption{$T = 40$K}
+                \caption{$T = 400$K}
             \end{subfigure}
@@
                 \includegraphics[height=100pt, width=\linewidth]{liquid_85K.png}
-                \caption{$T = 85$K}
+                \caption{$T = 700$K}
             \end{subfigure}
@@
                 \includegraphics[height=100pt, width=\linewidth]{gas_120K.png}
-                \caption{$T = 120$K}
+                \caption{$T = 900$K}
             \end{subfigure}
         \end{tabular}
-        \caption[Argon phases]{Snapshots taken of the system at frame 500/999 for
-            temperatures of $40$, $85$ and $120$ kelvin. It is a bit difficult
-            to conclude from the pictures, one really needs to see the system
-            in motion, but the system do behave more or less like a solid, a
-            liquid and a gas for respective temperatures. It is possible to see a
-            certain change in the entropy of the system across the three
-            images. These made use of the Velocity-Verlet integration method.}
+        \caption[Argon phases]{Snapshots taken of the system at frame 500/999
+            for temperatures of $400$, $700$ and $900$ kelvin. It is a bit
+            difficult to conclude from the pictures, one really needs to see
+            the system in motion, but the system do behave more or less like a
+            solid, a liquid and a gas for respective temperatures. It is
+        possible to see a certain change in the entropy of the system across
+    the three images. These made use of the Velocity-Verlet integration
+method.}
         \label{fig:potential_iteration}
@@
         \caption[Temperature ratio]{Plotting the ratio between initial and
-            instantaneous temperature for varying $T_i$. Here the initial
-            temperatures ranges from 20 to 500 kelvin, with 48 intermediate steps.
-        As we can see the ratio stabilizes around 0.5. The graph of an arbitrary
-    temperature has been highlighted in black.}
+            instantaneous temperature for varying $T_i$. Each individual graph
+            corresponds to one value of $T_i$ and the ratio is plotted against
+            time given in units of tenths of a picosecond. Here the initial
+            temperatures ranges from 20 to 500 kelvin, with 48 intermediate
+            steps.  As we can see the ratio stabilizes around $1/2$. The graph of
+            an arbitrary temperature has been highlighted in black.}
         \label{fig:temperature_ratio}
@@
         \includegraphics[width=0.8\linewidth]{diffusion_constant.pdf}
-        \caption[Diffusion constant]{The diffusion constant as a function of temperature. We here
-                see that $D$ abruptly increases at the temperature of $T = 304K$
-                which indicates that the crystal is melting.  The two lines are the
-                least square solutions to the noisy data before and after,
-                respectively. The diffusion constant is given in units of area per second.}
+        \caption[Diffusion constant]{The diffusion constant as a function of
+            temperature. We here see that $D$ abruptly increases at the
+            temperature of $T = 304K$ which indicates that the crystal is
+        melting. The two lines are the linear graphs best fitted to the noisy
+    data.  The diffusion constant is given in units of area per second.}
         \label{fig:diffusion_constant}

### 2015-12-04 projects/project_5/molecular_dynamics/python_src/maxwell_boltzmann.py

@@ -1,0 +1,17 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+def maxwell_boltzmann(a, x):
+    return np.sqrt(2 / np.pi) * x**2 * np.exp(-x**2 / (2.0*a**2)) / a**3
+
+x_values = np.linspace(0, 20, 1000)
+a_values = range(1, 5)
+
+for a in a_values:
+    m_values = maxwell_boltzmann(a, x_values)
+    plt.plot(x_values, m_values, label='$ a = %d $' % a, alpha=0.7, linewidth=1.5)
+plt.legend(loc='best')
+plt.xlabel('$x$')
+plt.ylabel('PDF')
+plt.grid('on')
+plt.savefig('maxwell_boltzmann.pdf')

### 2015-12-04 projects/project_5/molecular_dynamics/python_src/simulations.py

@@ -60,7 +60,7 @@ def taskF(run_simulations=False):
     
     T_values = range(20, 500, 10) 
     b = 5.26
-    N = 5
+    N = 6
     description = 'temperature_ratios' 
     for i, T_i in enumerate(T_values):
         simulations.append([System(T=T_i, b=5.26, N_each_dimension=5, ID=i, simulation_description=description), T_i])
@@ -72,12 +72,12 @@ def taskF(run_simulations=False):
     for i, sim in enumerate(simulations):
         sim[0].read_data()
         if i == int(len(simulations) / 2):
-            plt.plot(sim[0].temperature / sim[1], 'black', alpha=1.0)
+            plt.plot(sim[0].time[:500], sim[0].temperature[:500] / sim[1], 'black', alpha=1.0)
             continue;
-        plt.plot(sim[0].temperature / sim[1], alpha=0.3)
+        plt.plot(sim[0].time[:500], sim[0].temperature[:500] / sim[1], alpha=0.3)
 
     plt.grid('on')
-    plt.xlabel('$t$')
+    plt.xlabel('$t \\quad [ps/10]$')
     plt.ylabel('$T / T_i$')
     plt.savefig('temperature_ratio.pdf')
 
@@ -127,4 +127,5 @@ def taskG(run_simulations=False):
     plt.grid('on')
     plt.savefig('diffusion_constant.pdf')
 
-taskG(False)
+
+taskF(False)

### 2015-12-04 projects/project_5/molecular_dynamics/python_src/system_class.py

@@ -37,6 +37,7 @@ class System:
         self.kinetic = []
         self.total = []
         self.diffusion = []
+        self.time = []
         
         with open(self.file_name, 'r') as data_file:
             for line in data_file:
@@ -46,8 +47,11 @@ class System:
                 self.kinetic.append(line[2])
                 self.total.append(line[2] + line[1])
                 self.diffusion.append(line[4])
+                self.time.append(line[5])
+
         self.temperature = np.array(self.temperature)
         self.potential = np.array(self.potential)
         self.kinetic = np.array(self.kinetic)
         self.total = np.array(self.total)
         self.diffusion = np.array(self.diffusion)
+        self.time = np.array(self.time)

### 2015-12-07 projects/project_5/molecular_dynamics/article/article.tex
Writing introduction

@@
     the late 1950's, it has had a wide range of applications, in fields as
-    materials science, biochemistry and biophysics. It is based on the simple
-    idea of considering a finite number of molecules and letting them interact
-    for a fixed amount of time, which gives us an idea of how the system
-    evolves.
+    materials science, biochemistry and biophysics. Some of the more pioneering
+    works in molecular dynamics are often found in the study of chemical
+    physics and more specifically proteins.  In 1975, Levitt and Warshel
+    \cite{folding} simulated protein folding and in 1976 Cooper
+    \cite{thermoprotein} presented results on thermodynamic fluctuations in
+    protein folding with methods comparable to todays standards.
+
+    In order to verify the molecular dynamics model one can for instance
+    compare with physical experiments. For molecular systems one particular
+    method is of special interest, namely nuclear magnetic resonance
+    spectroscopy.
+
+    Molecular Dynamics can take on many forms. Notable in the addition of what
+    we examine in this project, is the Monte-Carlo and Brownian dynamics
+    variants of MD.
+
+    It is based on the simple idea of considering a finite number of molecules
+    and letting them interact for a fixed amount of time, which gives us an
+    idea of how the system evolves.
 
@@
 
+\section{Conclusion}
+\label{sec:conclusion}
+
 \begin{thebibliography}{9}
+\bibitem{folding}
+    Levitt, M., \& Warshel, A., `Computer simulation of protein folding`, Nature, Vol. 253, No. 5494. Available from: \url{http://csb.stanford.edu/levitt/Levitt_NAT75_folding.pdf}
+\bibitem{thermoprotein}
+    Cooper, A.
+`Thermodynamic fluctuations in protein molecules`, Proceedings of the National Academy of Sciences of the United States of America, Vol.
+73, No.
+8.
+Available from: \url{http://www.pnas.org/content/73/8/2740.full.pdf}
 \bibitem{min_image}
     \url{https://en.wikipedia.org/wiki/Periodic_boundary_conditions#Practical_implementation:_continuity_and_the_minimum_image_convention}
-\bibitem{equipar}
-    \url{https://en.wikipedia.org/wiki/Equipartition\_theorem}
-\bibitem{virial}
-    \url{https://en.wikipedia.org/wiki/Virial_theorem}
+\bibitem{equipar} \url{https://en.wikipedia.org/wiki/Equipartition\_theorem}
+\bibitem{virial} \url{https://en.wikipedia.org/wiki/Virial_theorem}
 \end{thebibliography}

### 2015-12-07 projects/project_5/molecular_dynamics/article/article.tex
introduction finished

@@
 \label{sec:introduction}
+
     Molecular dynamics (referred to as MD), deals with the time evolution of a
@@
     \cite{thermoprotein} presented results on thermodynamic fluctuations in
-    protein folding with methods comparable to todays standards.
+    protein folding with methods comparable to todays standards. More related
+    to our project are the results of A. Rahman in 1964 \cite{liquid_argon}
+    where he simulates liquid argon and find his simulations to more or less
+    agree with experiments.
 
@@
     method is of special interest, namely nuclear magnetic resonance
-    spectroscopy.
+    spectroscopy.  Molecular Dynamics can take on many forms. Notable in the
+    addition to what we examine in this project, is the Monte-Carlo and
+    Brownian dynamics variants of MD.
 
-    Molecular Dynamics can take on many forms. Notable in the addition of what
-    we examine in this project, is the Monte-Carlo and Brownian dynamics
-    variants of MD.
-
-    It is based on the simple idea of considering a finite number of molecules
-    and letting them interact for a fixed amount of time, which gives us an
-    idea of how the system evolves.
-
-    In this project we examine what goes into creating a working MD-program
-    that can compute various physical quantities and then proceed by
-    considering some specific systems where we look at, amongst other, the
-    kinetic energy, the initial temperature required in order to reach a
-    desired final temperature, and finding the diffusion constant for the
-    system in order to compute the melting temperature.
+    Molecular dynamics is based on the simple idea of considering a finite
+    number of molecules and letting them interact for a fixed amount of time,
+    which gives us an idea of how the system evolves.  In this project we
+    examine what goes into creating a working MD-program that can compute
+    various physical quantities and then proceed by considering some specific
+    systems where we look at, amongst other, the kinetic energy, the initial
+    temperature required in order to reach a desired final temperature, and
+    finding the diffusion constant for the system in order to compute the
+    melting temperature.
 
@@
 Available from: \url{http://www.pnas.org/content/73/8/2740.full.pdf}
+\bibitem{liquid_argon}
+    Rahman, A., `Correlations in the Motion of Atoms in Liquid Argon`, Physical Review, Volume 136, No. 2A. Available from: \url{http://journals.aps.org/pr/pdf/10.1103/PhysRev.136.A405}.
 \bibitem{min_image}
-    \url{https://en.wikipedia.org/wiki/Periodic_boundary_conditions#Practical_implementation:_continuity_and_the_minimum_image_convention}
-\bibitem{equipar} \url{https://en.wikipedia.org/wiki/Equipartition\_theorem}
-\bibitem{virial} \url{https://en.wikipedia.org/wiki/Virial_theorem}
+    `Periodic boundary conditions` (wiki article), December 7, 2015.
+Available from: \url{https://en.wikipedia.org/wiki/Periodic_boundary_conditions#Practical_implementation:_continuity_and_the_minimum_image_convention} [November 13, 2015]
+\bibitem{equipar} 
+    `Equipartition theorem` (wiki article), December 7, 2015.
+    Available from: \url{https://en.wikipedia.org/wiki/Equipartition\_theorem} [November 21, 2015]
+\bibitem{virial} 
+    `Virial theorem` (wiki article), December 7, 2015, \url{https://en.wikipedia.org/wiki/Virial_theorem} [October 7, 2015]
 \end{thebibliography}

### 2015-12-07 projects/project_5/molecular_dynamics/article/article.tex
done?

@@
 \end{abstract}
+\clearpage
 \doublespacing
@@
 \singlespacing
-
+\clearpage
 \section{Introduction}
@@
 
+In this project we have mainly been concerned with the properties of argon. We
+have most notably looked at how a molecular dynamics simulation of argon using
+the Lennard-Jones potential behaves.  The results tells us that the
+experimental values for argon are not properly reflected in the simulations,
+however the system does more or less behave as it should, albeit for a larger
+temperature. When considering the force model we made some optimizations to the
+potential, and only looked at the interactions between atoms sufficiently close
+to each other. This lead to a significant increase in simulation speed, from a
+time complexity of $\mathcal{O}(N^2)$ to one of $\mathcal{O}(N)$.
+
+We also chose to implement the Velocity Verlet algorithm as our time
+integrator, in preference over the more basic Euler-Cromer method. This lead to
+a more stable energy conservation over varying time steps. The improvement was
+from approximately linear to a constant.
+
+We also considered the melting temperature of argon, and measured it to be
+somewhere close to $300K$, which does not really coincide with the
+experimentally observed melting temperature of approximately $84K$. Despite
+this disparity the system at least behaves consistent.
+
 \begin{thebibliography}{9}

### 2015-12-07 projects/project_5/molecular_dynamics/article/article.tex
cleaning up

@@
 \usepackage[]{graphicx} 
+\usepackage[]{xcolor} 
 \usepackage[]{subcaption} 
@@
 \renewcommand{\vec}[1]{\mathbf{#1}}
+\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
+\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{github}}}
 
@@
     properties. We also compute the melting temperature of argon, by means of
-    the Einstein relation and diffusion constant.
+    the Einstein relation and diffusion constant. All visualizations of the
+    system has been performed using the program \texttt{VMD}.\\
+    \noindent
+    As always, relevant code can be found on my GitHub page:\\
+
+    \centering{\href{https://github.com/qTipTip/FYS3150}{\githublogo \,
+    \color{gray!50}\textit{github.com/qTipTip}}}
 \end{abstract}
@@
     \mathrm{\AA}$ and $\varepsilon = 119.8k_B\mathrm{K}$ with $k_B = 1$ in
-    MD-units.
+    MD-units are appropriate values for argon.
 
@@
 experimentally observed melting temperature of approximately $84K$. Despite
-this disparity the system at least behaves consistent.
+this disparity the system at least behaves consistently.
 

### 2015-12-07 projects/project_5/molecular_dynamics/src/atom.cpp
added c++ source

@@ -1,0 +1,22 @@
+#include "atom.h"
+#include "math/random.h"
+#include <cmath>
+
+Atom::Atom(double mass) :
+    m_mass(mass)
+{
+    
+}
+
+void Atom::resetForce()
+{
+    force.setToZero();
+}
+
+void Atom::resetVelocityMaxwellian(double temperature)
+{
+    // Resetting the velocity according to a Maxwell-Boltzmann distribution (see http://en.wikipedia.org/wiki/Maxwell%E2%80%93Boltzmann_distribution )
+    double boltzmannConstant = 1.0; // In these units, the boltzmann constant equals 1
+    double standardDeviation = sqrt(boltzmannConstant*temperature/m_mass);
+    velocity.randomGaussian(0, standardDeviation);
+}

### 2015-12-07 projects/project_5/molecular_dynamics/src/atom.h

@@ -1,0 +1,25 @@
+#ifndef ATOM_H
+#define ATOM_H
+#include "math/vec3.h"
+#include <vector>
+using std::vector;
+class Atom
+{
+private:
+    float m_mass;
+public:
+    vector<Atom*> neighbors;
+    vec3 position;
+    vec3 initial_position;
+    vec3 positionLastNeighborlistBuild;
+    vec3 velocity;
+    vec3 force;
+
+    Atom(double mass);
+    void resetForce();
+    void resetVelocityMaxwellian(double temperature);
+
+    double mass() { return m_mass; }
+    void setMass(double mass) { m_mass = mass; }
+};
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/celllist.cpp

@@ -1,0 +1,37 @@
+#include "celllist.h"
+#include "system.h"
+#include "atom.h"
+#include <iostream>
+using namespace std;
+
+CellList::CellList()
+{
+
+}
+
+void CellList::build(System &system)
+{
+    numberOfCellsPerDimension = system.systemSize().x() / cutoffDistance;
+    cutoffDistance = system.systemSize().x() / numberOfCellsPerDimension;
+
+    // Resize so our cell vectors have the correct size
+    m_cells.resize(numberOfCellsPerDimension);
+    for(int i=0; i<numberOfCellsPerDimension; i++) {
+        m_cells[i].resize(numberOfCellsPerDimension);
+        for(int j=0; j<numberOfCellsPerDimension; j++) {
+            m_cells[i][j].resize(numberOfCellsPerDimension);
+            for(int k=0; k<numberOfCellsPerDimension; k++) {
+                m_cells[i][j][k].clear();
+            }
+        }
+    }
+
+    // Compute the cell dimensions
+    double cellSize = system.systemSize().x()/numberOfCellsPerDimension;
+    for(Atom *atom : system.atoms()) {
+        int i = atom->position.x()/cellSize;
+        int j = atom->position.y()/cellSize;
+        int k = atom->position.z()/cellSize;
+        m_cells.at(i).at(j).at(k).push_back(atom);
+    }
+}

### 2015-12-07 projects/project_5/molecular_dynamics/src/celllist.h

@@ -1,0 +1,21 @@
+#ifndef CELLLIST_H
+#define CELLLIST_H
+#include <vector>
+using std::vector;
+class Atom; class System;
+
+class CellList
+{
+private:
+    // This array will then have indices [cx][cy][cz][atomIndex], where the three first
+    // are the 3-dimensional coordinates to the cell.
+    vector<vector<vector<vector<Atom *> > > > m_cells;
+public:
+    double cutoffDistance = 2.5*3.405; // rCut + rShell
+    int numberOfCellsPerDimension = -1;
+    CellList();
+    void build(System &system);
+    vector<Atom*> & cell(int i, int j, int k) { return m_cells[i][j][k]; }
+};
+
+#endif // CELLLIST_H

### 2015-12-07 projects/project_5/molecular_dynamics/src/integrators/eulercromer.cpp

@@ -1,0 +1,13 @@
+#include "eulercromer.h"
+#include "../system.h"
+
+void EulerCromer::integrate(System *system, double dt)
+{
+    system->calculateForces();
+    for(Atom *atom : system->atoms()) {
+        atom->velocity += atom->force*dt / atom->mass();
+        atom->position += atom->velocity*dt;
+    }
+
+    system->applyPeriodicBoundaryConditions();
+}

### 2015-12-07 projects/project_5/molecular_dynamics/src/integrators/eulercromer.h

@@ -1,0 +1,14 @@
+#ifndef EULERCROMER_H
+#define EULERCROMER_H
+#include "integrators/integrator.h"
+
+class System;
+class EulerCromer : public Integrator
+{
+public:
+    EulerCromer() {}
+    ~EulerCromer() {}
+    virtual void integrate(System* system, double timestep);
+};
+
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/integrators/integrator.h

@@ -1,0 +1,13 @@
+#ifndef INTEGRATOR_H
+#define INTEGRATOR_H
+
+class System;
+class Integrator
+{
+public:
+    Integrator();
+    virtual ~Integrator() { }
+    virtual void integrate(System* system, double timestep) = 0;
+};
+
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/integrators/velocityverlet.cpp

@@ -1,0 +1,43 @@
+#include "velocityverlet.h"
+#include "../system.h"
+#include "../atom.h"
+#define FAST
+VelocityVerlet::VelocityVerlet() : m_first_step(true) // sets default value to true at construction
+{
+}
+
+void VelocityVerlet::integrate(System *system, double dt)
+{
+    // we first calculate the current forces
+    if(m_first_step) {
+        system->calculateForces();
+        m_first_step = false;
+    }
+    double dt_half = dt / 2;
+    // first we iterate over each atom in the system, and update the positions
+    double oneOverMass = 1.0/system->atoms()[0]->mass();
+    for(Atom *atom : system->atoms()) {
+        atom->velocity[0] += atom->force[0]*dt_half*oneOverMass;
+        atom->velocity[1] += atom->force[1]*dt_half*oneOverMass;
+        atom->velocity[2] += atom->force[2]*dt_half*oneOverMass;
+
+        atom->position[0] += atom->velocity[0] * dt;
+        atom->position[1] += atom->velocity[1] * dt;
+        atom->position[2] += atom->velocity[2] * dt;
+    }
+
+    // i spent a few good hours wondering why my simulations werent working. This line was missing
+    system->applyPeriodicBoundaryConditions();
+
+    // we then recompute the forces on all the atoms in preparation for computing the velocities
+    system->calculateForces();
+
+    // then we compute the velocities
+    for(Atom *atom : system->atoms()) {
+        atom->velocity[0] += atom->force[0]*dt_half*oneOverMass;
+        atom->velocity[1] += atom->force[1]*dt_half*oneOverMass;
+        atom->velocity[2] += atom->force[2]*dt_half*oneOverMass;
+    }
+
+
+}

### 2015-12-07 projects/project_5/molecular_dynamics/src/integrators/velocityverlet.h

@@ -1,0 +1,13 @@
+#ifndef VELOCITYVERLET_H
+#define VELOCITYVERLET_H
+#include "integrators/integrator.h"
+
+class VelocityVerlet : public Integrator
+{
+public:
+    VelocityVerlet();
+    virtual void integrate(System *system, double dt) override;
+private:
+    bool m_first_step;
+};
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/io.cpp

@@ -1,0 +1,42 @@
+#include "io.h"
+#include "system.h"
+#include "atom.h"
+#include "unitconverter.h"
+#include <cstdlib>
+using std::endl; using std::cout;
+
+IO::IO()
+{
+
+}
+
+IO::~IO() {
+    close();
+}
+
+void IO::open(const char *filename) {
+    if(file.is_open()) {
+        std::cout << "<IO.cpp> Error, tried to open file " << filename << ", but some file is already open." << endl;
+        exit(1);
+    }
+
+    file.open(filename);
+}
+
+void IO::close() {
+    if(file.is_open()) {
+        file.close();
+    }
+}
+
+// This saves the current state to a file following the xyz-standard (see http://en.wikipedia.org/wiki/XYZ_file_format )
+void IO::saveState(System *system)
+{
+    if(file.is_open()) {
+        file << system->atoms().size() << endl;
+        file << "The is an optional comment line that can be empty." << endl;
+        for(Atom *atom : system->atoms()) {
+            file << "Ar " << UnitConverter::lengthToAngstroms(atom->position.x()) << " " << UnitConverter::lengthToAngstroms(atom->position.y()) << " " << UnitConverter::lengthToAngstroms(atom->position.z()) << endl;
+        }
+    }
+}

### 2015-12-07 projects/project_5/molecular_dynamics/src/io.h

@@ -1,0 +1,20 @@
+#ifndef IO_H
+#define IO_H
+#include <fstream>
+class System;
+using std::ofstream;
+
+class IO
+{
+private:
+    ofstream file;
+public:
+    IO();
+    ~IO();
+
+    void saveState(System *system);
+    void open(const char *filename);
+    void close();
+
+};
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/main.cpp

@@ -1,0 +1,81 @@
+#include "math/random.h"
+#include "potentials/lennardjones.h"
+#include "integrators/eulercromer.h"
+#include "integrators/velocityverlet.h"
+#include "system.h"
+#include "statisticssampler.h"
+#include "atom.h"
+#include "io.h"
+#include "unitconverter.h"
+#include <iostream>
+#include <QElapsedTimer>
+
+using namespace std;
+
+int main(int numberOfArguments, char **argumentList)
+{
+    int numberOfUnitCells = 10;
+    double initialTemperature = UnitConverter::temperatureFromSI(2000); // measured in Kelvin
+    double latticeConstant = UnitConverter::lengthFromAngstroms(5.26); // measured in angstroms
+    double dt = UnitConverter::timeFromSI(1.0e-15); // measured in seconds
+
+    // If a first argument is provided, it is the number of unit cells
+    if(numberOfArguments > 1) numberOfUnitCells = atoi(argumentList[1]);
+    // If a second argument is provided, it is the initial temperature (measured in kelvin)
+    if(numberOfArguments > 2) initialTemperature = UnitConverter::temperatureFromSI(atof(argumentList[2]));
+    // If a third argument is provided, it is the lattice constant determining the density (measured in angstroms)
+    if(numberOfArguments > 3) latticeConstant = UnitConverter::lengthFromAngstroms(atof(argumentList[3]));
+    // time step in seconds from the command line
+    if(numberOfArguments > 5) dt = UnitConverter::timeFromSI(atof(argumentList[5]));
+
+    cout << "One unit of length is " << UnitConverter::lengthToSI(1.0) << " meters" << endl;
+    cout << "One unit of velocity is " << UnitConverter::velocityToSI(1.0) << " meters/second" << endl;
+    cout << "One unit of time is " << UnitConverter::timeToSI(1.0) << " seconds" << endl;
+    cout << "One unit of mass is " << UnitConverter::massToSI(1.0) << " kg" << endl;
+    cout << "One unit of temperature is " << UnitConverter::temperatureToSI(1.0) << " K" << endl;
+
+    System system;
+    system.createFCCLattice(numberOfUnitCells, latticeConstant, initialTemperature);
+    system.setPotential(new LennardJones(3.405, 1.0)); // You must insert correct parameters here
+    system.potential()->setRCut(2.5*3.405);
+    system.potential()->setRShell(0.3*3.405);
+
[…]

### 2015-12-07 projects/project_5/molecular_dynamics/src/math/random.cpp

@@ -1,0 +1,40 @@
+#include "random.h"
+#include <cmath>
+
+long     Random::iy = 0;
+long     Random::iv[NTAB];
+long     Random::seed = -1;
+void Random::setSeed(long seed) {
+    Random::seed = seed;
+}
+
+double Random::nextGaussian(double mean, double standardDeviation) {
+    double standardNormalRandomNumber = sqrt( -2.0*log(1.0 - nextDouble()) ) * cos( 6.283185307 * nextDouble() );
+    return standardDeviation*standardNormalRandomNumber + mean;
+}
+
+double Random::nextDouble()
+{
+    int             j;
+    long            k;
+    double          temp;
+    if (Random::seed <= 0 || !iy) {
+        if (-(Random::seed) < 1) Random::seed=1;
+        else Random::seed = -(Random::seed);
+        for(j = NTAB + 7; j >= 0; j--) {
+            k     = (Random::seed)/IQ;
+            Random::seed = IA*(Random::seed - k*IQ) - IR*k;
+            if(Random::seed < 0) Random::seed += IM;
+            if(j < NTAB) iv[j] = Random::seed;
+        }
+        iy = iv[0];
+    }
+    k     = (Random::seed)/IQ;
+    Random::seed = IA*(Random::seed - k*IQ) - IR*k;
+    if(Random::seed < 0) Random::seed += IM;
+    j     = iy/NDIV;
+    iy    = iv[j];
+    iv[j] = Random::seed;
+    if((temp=AM*iy) > RNMX) return RNMX;
+    else return temp;
+}

### 2015-12-07 projects/project_5/molecular_dynamics/src/math/random.h

@@ -1,0 +1,38 @@
+/*
+ This is based on the ran1-generator from lib.cpp (see http://www.uio.no/studier/emner/matnat/fys/FYS3150/h14/index.html )
+     ** The function
+     **           ran1()
+     ** is an "Minimal" random number generator of Park and Miller
+     ** (see Numerical recipe page 280) with Bays-Durham shuffle and
+     ** added safeguards. Call with idum a negative integer to initialize;
+     ** thereafter, do not alter idum between sucessive deviates in a
+     ** sequence. RNMX should approximate the largest floating point value
+     ** that is less than 1.
+     ** The function returns a uniform deviate between 0.0 and 1.0
+     ** (exclusive of end-point values).
+*/
+
+#ifndef RANDOM_H
+#define RANDOM_H
+
+#define IA 16807
+#define IM 2147483647
+#define AM (1.0/IM)
+#define IQ 127773
+#define IR 2836
+#define NTAB 32
+#define NDIV (1+(IM-1)/NTAB)
+#define EPS 1.2e-7
+#define RNMX (1.0-EPS)
+
+class Random {
+public:
+    static long iy;
+    static long iv[NTAB];
+    static long seed;
+    static double nextDouble();
+    static double nextGaussian(double mean, double standardDeviation);
+    static void setSeed(long seed);
+};
+
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/math/vec3.h

@@ -1,0 +1,107 @@
+#ifndef VEC3_H
+#define VEC3_H
+#include <string>
+#include <vector>
+#include <iostream>
+using std::string;
+
+class vec3
+{
+public:
+    double components[3];
+
+    vec3();
+    vec3(vec3 const&copy);
+    vec3(double x, double y, double z);
+
+    void print();
+    void print(string name);
+    vec3 cross(vec3 aVector);
+    double lengthSquared();
+    double length();
+    double x() const { return components[0]; }
+    double y() const { return components[1]; }
+    double z() const { return components[2]; }
+    void setX(double x) { components[0] = x; }
+    void setY(double y) { components[1] = y; }
+    void setZ(double z) { components[2] = z; }
+    void set(double x, double y, double z);
+    void setToZero() { components[0] = 0; components[1] = 0; components[2] = 0; }
+    double &operator()(int index) { return components[index]; } // Allows access like myVector(0)
+    double &operator[](int index) { return components[index]; } // Allows access like myVector[0]
+    void randomGaussian(double mean, double standardDeviation);
+    vec3 &operator+=(double rhs); // Componentwise addition with scalar
+    vec3 &operator+=(vec3 rhs);   // Componentwise addition with vector
+    vec3 &operator*=(double rhs); // Componentwise multiplication with scalar
+    vec3 &operator*=(vec3 rhs);   // Componentwise multiplicationwith vector
+    vec3 &operator-=(double rhs); // Componentwise subtraction with scalar
+    vec3 &operator-=(vec3 rhs);   // Componentwise subtraction with vector
+    vec3 &operator/=(double rhs); // Componentwise division with scalar
+    vec3 &operator/=(vec3 rhs);   // Componentwise division with vector
+    friend std::ostream& operator<<(std::ostream& os, const vec3& myVector); // Allows cout << myVector << endl;
+};
+
+inline vec3 operator+(vec3 lhs, double rhs) {
+    lhs += rhs;
+    return lhs;
+}
+
+inline vec3 operator+(double lhs, vec3 rhs) {
+    rhs += lhs;
+    return rhs;
+}
+
[…]

### 2015-12-07 projects/project_5/molecular_dynamics/src/potentials/lennardjones.cpp

@@ -1,0 +1,128 @@
+#include "lennardjones.h"
+#include <cmath>
+LennardJones::LennardJones(double sigma, double epsilon) :
+    m_sigma(sigma),
+    m_epsilon(epsilon)
+{
+
+}
+
+void LennardJones::calculateForces(System *system) {
+    if(system->systemSize().x() < 3*(m_rCut+m_rShell)) {
+        calculateForcesN2(system);
+    } else {
+        calculateForcesNeighborlist(system);
+    }
+}
+
+void LennardJones::calculateForcesNeighborlist(System *system)
+{
+    m_potentialEnergy = 0; // Remember to compute this in the loop
+    double size_x = system->systemSize().x();
+    double size_y = system->systemSize().y();
+    double size_z = system->systemSize().z();
+    double sigma_six = m_sigma*m_sigma*m_sigma*m_sigma*m_sigma*m_sigma;
+    double potentialEnergyAtRcut = 4 * m_epsilon *pow(m_rCut,-6)*sigma_six*(sigma_six*pow(m_rCut,-6)  - 1.0);
+
+    vector<Atom*> &atoms = system->atoms();
+    // we sum over all distinct pairs of atoms
+    for(Atom *atom_i : atoms) {
+        for(Atom *atom_j : atom_i->neighbors) {
+            double xij = atom_i->position[0] - atom_j->position[0];
+            double yij = atom_i->position[1] - atom_j->position[1];
+            double zij = atom_i->position[2] - atom_j->position[2];
+
+            //since our system is periodic, we have to account for the two atoms being closer through the boundaries
+            //than through the body of the lattice.
+
+            if(xij > 0.5*size_x) xij -= size_x;
+            else if(xij < -0.5*size_x) xij += size_x;
+
+            if(yij > 0.5*size_y) yij -= size_y;
+            else if(yij < -0.5*size_y) yij += size_y;
+
+            if(zij > 0.5*size_z) zij -= size_z;
+            else if(zij < -0.5*size_z) zij += size_z;
+
+            // to make the program statements less daunting we precompute some values used in the expression for F
+            double dr2 = xij*xij + yij*yij + zij*zij;
+            if(dr2 >= m_rCut*m_rCut) continue;
+
+            double oneOverDr2 = 1.0/dr2;
+            double oneOverDr6 = oneOverDr2*oneOverDr2*oneOverDr2;
+
[…]

### 2015-12-07 projects/project_5/molecular_dynamics/src/potentials/lennardjones.h

@@ -1,0 +1,16 @@
+#ifndef LENNARDJONES_H
+#define LENNARDJONES_H
+#include "potential.h"
+
+class LennardJones : public Potential
+{
+private:
+    double m_sigma = 1.0;
+    double m_epsilon = 1.0;
+public:
+    LennardJones(double sigma, double epsilon);
+    virtual void calculateForces(System *system);
+    void calculateForcesN2(System *system);
+    void calculateForcesNeighborlist(System *system);
+};
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/potentials/potential.cpp

@@ -1,0 +1,91 @@
+#include "potential.h"
+
+double Potential::rCut() const
+{
+    return m_rCut;
+}
+
+void Potential::setRCut(double rCut)
+{
+    m_rCut = rCut;
+}
+
+double Potential::rShell() const
+{
+    return m_rShell;
+}
+
+void Potential::setRShell(double rShell)
+{
+    m_rShell = rShell;
+}
+
+Potential::Potential()
+{
+    
+}
+
+double Potential::potentialEnergy()
+{
+    return m_potentialEnergy;
+}
+
+void Potential::setPotentialEnergy(double potentialEnergy)
+{
+    m_potentialEnergy = potentialEnergy;
+}
+
+void Potential::buildNeighborList(System &system) {
+    system.cellList().cutoffDistance = m_rCut + m_rShell;
+    system.cellList().build(system);
+    for(Atom *atom : system.atoms()) {
+        atom->neighbors.clear();
+        atom->positionLastNeighborlistBuild = atom->position;
+    }
+
+    CellList &cellList = system.cellList();
+    vec3 systemSize = system.systemSize();
+    double radiusSquared = (m_rCut + m_rShell)*(m_rCut + m_rShell);
+
+    for(int cx = 0; cx<cellList.numberOfCellsPerDimension; cx++) {
+        for(int cy = 0; cy<cellList.numberOfCellsPerDimension; cy++) {
+            for(int cz = 0; cz<cellList.numberOfCellsPerDimension; cz++) {
+                vector<Atom*> &cell1 = cellList.cell(cx, cy, cz);
+
+                for(int dx=-1; dx<=1; dx++) {
+                    for(int dy=-1; dy<=1; dy++) {
+                        for(int dz=-1; dz<=1; dz++) {
+                            int i = (cx + dx + cellList.numberOfCellsPerDimension) % cellList.numberOfCellsPerDimension;
+                            int j = (cy + dy + cellList.numberOfCellsPerDimension) % cellList.numberOfCellsPerDimension;
+                            int k = (cz + dz + cellList.numberOfCellsPerDimension) % cellList.numberOfCellsPerDimension;
+
+                            vector<Atom*> &cell2 = cellList.cell(i, j, k);
+                            for(Atom *atom1 : cell1) {
+                                for(Atom *atom2 : cell2) {
+                                    if(atom1 <= atom2) continue;
+                                    double dx = atom1->position[0] - atom2->position[0];
[…]

### 2015-12-07 projects/project_5/molecular_dynamics/src/potentials/potential.h

@@ -1,0 +1,26 @@
+#ifndef POTENTIAL_H
+#define POTENTIAL_H
+
+#include <string>
+#include <vector>
+#include "../system.h"
+
+class Potential
+{
+protected:
+    double m_potentialEnergy = 0;
+    double m_rCut = -1;
+    double m_rShell = -1;
+public:
+    Potential();
+    virtual ~Potential() {}
+    virtual void calculateForces(System *system) = 0;
+    double potentialEnergy();
+    void setPotentialEnergy(double potentialEnergy);
+    double rCut() const;
+    void setRCut(double rCut);
+    double rShell() const;
+    void setRShell(double rShell);
+    void buildNeighborList(System &system);
+};
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/statisticssampler.cpp

@@ -1,0 +1,91 @@
+#include "system.h"
+#include "statisticssampler.h"
+#include "potentials/potential.h"
+#include "fstream"
+#include "unitconverter.h"
+StatisticsSampler::StatisticsSampler()
+{
+
+}
+
+void StatisticsSampler::saveToFile(System &system)
+{
+    // Save the statistical properties for each timestep for plotting etc.
+
+    using namespace std;
+    ofstream outfile;
+    outfile.open("samples.dat", ios::app);
+    outfile << UnitConverter::temperatureToSI(m_temperature) << " "
+            << UnitConverter::energyToEv(m_kineticEnergy/system.atoms().size()) << " "
+            << UnitConverter::energyToEv(m_potentialEnergy/system.atoms().size()) << " "
+            << UnitConverter::energyToEv(totalEnergy()) << " "
+            << m_diffusionConstant << " "
+            << system.time() << endl;
+}
+
+void StatisticsSampler::sample(System &system)
+{
+    // Here you should measure different kinds of statistical properties and save it to a file.
+    sampleKineticEnergy(system);
+    samplePotentialEnergy(system);
+    sampleTemperature(system);
+    sampleDensity(system);
+    sampleMeanSquareDisplacement(system);
+    sampleDiffusionConstant(system);
+    saveToFile(system);
+}
+
+void StatisticsSampler::sampleKineticEnergy(System &system)
+{
+    m_kineticEnergy = 0; // Remember to reset the value from the previous timestep
+
+    // for each atom we compute its mass times velocity squared, half it and add it to the counter
+    for(Atom *atom : system.atoms()) {
+        m_kineticEnergy += 0.5*atom->mass()*atom->velocity.lengthSquared();
+    }
+}
+
+void StatisticsSampler::samplePotentialEnergy(System &system)
+{
+    // the potential is calculated in the Potential-class so
+    // we simply extract it from there
+    m_potentialEnergy = system.potential()->potentialEnergy();
+}
+
+void StatisticsSampler::sampleTemperature(System &system)
+{
+    //kb = 1
+    int n_atoms = system.atoms().size();
+    m_temperature = 2.0/3.0 * m_kineticEnergy / (n_atoms);
+}
+
+
+
+void StatisticsSampler::sampleDensity(System &system)
+{
+
+}
+
+void StatisticsSampler::sampleMeanSquareDisplacement(System &system)
+{
+    // initialize to zero
+    m_meanSquareDisplacement = 0.0;
+
[…]

### 2015-12-07 projects/project_5/molecular_dynamics/src/statisticssampler.h

@@ -1,0 +1,30 @@
+#ifndef STATISTICSSAMPLER_H
+#define STATISTICSSAMPLER_H
+
+class System;
+class StatisticsSampler
+{
+private:
+    double m_kineticEnergy = 0;
+    double m_potentialEnergy = 0;
+    double m_temperature = 0;
+    double m_density = 0;
+    double m_meanSquareDisplacement = 0;
+    double m_diffusionConstant = 0;
+public:
+    StatisticsSampler();
+    void saveToFile(System &system);
+    void sample(System &system);
+    void sampleKineticEnergy(System &system);
+    void samplePotentialEnergy(System &system);
+    void sampleTemperature(System &system);
+    void sampleDensity(System &system);
+    void sampleMeanSquareDisplacement(System &system);
+    void sampleDiffusionConstant(System &system);
+    double kineticEnergy() { return m_kineticEnergy; }
+    double potentialEnergy() { return m_potentialEnergy; }
+    double totalEnergy() { return m_kineticEnergy+m_potentialEnergy; }
+    double temperature() { return m_temperature; }
+    double density() { return m_density; }
+};
+#endif

### 2015-12-07 projects/project_5/molecular_dynamics/src/system.cpp

@@ -1,0 +1,152 @@
+#include "system.h"
+#include "integrators/integrator.h"
+#include "potentials/potential.h"
+#include "statisticssampler.h"
+#include "unitconverter.h"
+#include "math/random.h"
+#include <cmath>
+
+System::System()
+{
+
+}
+
+System::~System()
+{
+    delete m_potential;
+    delete m_integrator;
+    for(Atom *atom : m_atoms) {
+        delete atom;
+    }
+    m_atoms.clear();
+}
+
+void System::applyPeriodicBoundaryConditions() {
+    // Read here: http://en.wikipedia.org/wiki/Periodic_boundary_conditions#Practical_implementation:_continuity_and_the_minimum_image_convention
+    // This method also takes into consideration the computation of the diffusion constant, hence
+    // we in addition to moving the position, also move the initial position
+    // to keep track of the true position of the atom
+    for(Atom *atom : m_atoms) {
+
+        if(atom->position[0] > m_systemSize[0]) {
+            atom->position[0] -= m_systemSize[0];
+            atom->initial_position[0] -= m_systemSize[0];
+            atom->positionLastNeighborlistBuild[0] -= m_systemSize[0];
+        }
+        else if(atom->position[0] < 0) {
+            atom->position[0] += m_systemSize[0];
+            atom->initial_position[0] += m_systemSize[0];
+            atom->positionLastNeighborlistBuild[0] += m_systemSize[0];
+        }
+
+        if(atom->position[1] > m_systemSize[1]) {
+            atom->position[1] -= m_systemSize[1];
+            atom->initial_position[1] -= m_systemSize[1];
+            atom->positionLastNeighborlistBuild[1] -= m_systemSize[1];
+        }
+        else if(atom->position[1] < 0) {
+            atom->position[1] += m_systemSize[1];
+            atom->initial_position[1] += m_systemSize[1];
+            atom->positionLastNeighborlistBuild[1] += m_systemSize[1];
+        }
+        if(atom->position[2] > m_systemSize[2]) {
+            atom->position[2] -= m_systemSize[2];
+            atom->initial_position[2] -= m_systemSize[2];
+            atom->positionLastNeighborlistBuild[2] -= m_systemSize[2];
+        }
+        else if(atom->position[2] < 0) {
+            atom->position[2] += m_systemSize[2];
+            atom->initial_position[2] += m_systemSize[2];
[…]

### 2015-12-07 projects/project_5/molecular_dynamics/src/system.h

@@ -1,0 +1,48 @@
+#ifndef SYSTEM_H
+#define SYSTEM_H
+#include "atom.h"
+#include "math/vec3.h"
+#include <vector>
+#include "celllist.h"
+class Potential; class Integrator;
+using std::vector;
+
+class System
+{
+private:
+    CellList m_cellList;
+    vec3 m_systemSize;
+    vector<Atom*> m_atoms;
+    Potential* m_potential = nullptr;
+    Integrator* m_integrator = nullptr;
+    double m_time = 0;
+    int m_steps = 0;
+    bool m_firstStep = true;
+
+    void validateNeighborList();
+public:
+    System();
+    ~System();
+    void resetForcesOnAllAtoms();
+    void createFCCLattice(int numberOfUnitCellsEachDimension, double latticeConstant, double temperature);
+    void applyPeriodicBoundaryConditions();
+    void removeTotalMomentum();
+    void calculateForces();
+    void step(double dt);
+
+    // Setters and getters
+    vector<Atom *>& atoms() { return m_atoms; } // Returns a reference to the std::vector of atom pointers
+    vec3 systemSize() { return m_systemSize; }
+    void setSystemSize(vec3 systemSize) { m_systemSize = systemSize; }
+    Potential *potential() { return m_potential; }
+    void setPotential(Potential *potential) { m_potential = potential; }
+    double time() { return m_time; }
+    void setTime(double time) { m_time = time; }
+    Integrator *integrator() { return m_integrator; }
+    void setIntegrator(Integrator *integrator) { m_integrator = integrator; }
+    int steps() { return m_steps; }
+    void setSteps(int steps) { m_steps = steps; }
+    double volume() { return m_systemSize[0]*m_systemSize[1]*m_systemSize[2]; }
+    CellList &cellList() { return m_cellList; }
+};
+#endif

### 2015-12-09 projects/project_5/molecular_dynamics/article/article.tex
added flowchart

@@
 \usepackage[]{setspace} 
+\usepackage[]{tikz} 
+\usetikzlibrary{shapes, arrows}
 \usepackage[]{hyperref} 
@@
 \tableofcontents
+\singlespacing
 \listoffigures      
-\singlespacing
 \clearpage
@@
     atom with a random velocity given by the Maxwell-Boltzmann distribution.
-
     Mathematically, the distribution function reads
@@
 
+    \subsection{Program Flow}
+    \label{sub:program_flow}
+
+    A flow-chart describing the execution of the program can be seen in
+    figure \ref{fig:flowchart}. I only a few of the more vital processes are
+    described in detail.
+    \begin{figure}[h]
+        \centering \includegraphics[width=0.8\linewidth]{flow_chart.pdf}
+        \caption[Flow chart]{A flow-chart of the program execution. The method
+            \texttt{system.step} is described in more detail because it is what
+            constitutes most of the program and is therefore included as a
+        separate subprocess. The chart quite crude, and is meant to give an
+    overarching idea of how the program executes.}
+        \label{fig:flowchart}
+    \end{figure}
 \section{Simulations}
@@
     \begin{figure}[h]
-        \centering \includegraphics[width=\linewidth]{temperature_ratio.pdf}
+        \centering \includegraphics[width=0.8\linewidth]{temperature_ratio.pdf}
         \caption[Temperature ratio]{Plotting the ratio between initial and
@@
     \begin{figure}[h]
-        \centering
-        \includegraphics[width=0.8\linewidth]{diffusion_constant.pdf}
+        \centering \includegraphics[width=0.8\linewidth]{diffusion_constant.pdf}
         \caption[Diffusion constant]{The diffusion constant as a function of

### 2015-12-09 projects/project_5/molecular_dynamics/article/article.tex
did references properly

@@
 \begin{thebibliography}{9}
-\bibitem{folding}
-    Levitt, M., \& Warshel, A., `Computer simulation of protein folding`, Nature, Vol. 253, No. 5494. Available from: \url{http://csb.stanford.edu/levitt/Levitt_NAT75_folding.pdf}
 \bibitem{thermoprotein}
     Cooper, A.
-`Thermodynamic fluctuations in protein molecules`, Proceedings of the National Academy of Sciences of the United States of America, Vol.
-73, No.
-8.
-Available from: \url{http://www.pnas.org/content/73/8/2740.full.pdf}
+(1976).
+Thermodynamic fluctuations in protein molecules.
+Proceedings of the National Academy of Sciences, 73, 2740-2741.
+Retrieved December 9, 2015, from \url{http://www.pnas.org/content/73/8/2740.full.pdf}
+\bibitem{equipar} 
+    Equipartition Theorem. (2015, November 21). Retrieved December 7, 2015, from \url{https://en.wikipedia.org/wiki/Equipartition_theorem}
+\bibitem{folding}
+    Levitt, M., \& Warshel, A. (1975). Computer simulation of protein folding. Nature, 253, 694-698. Retrieved December 7, 2015, from \url{http://csb.stanford.edu/levitt/Levitt_NAT75_folding.pdf}
+\bibitem{min_image}
+    Periodic boundary conditions. (2015, November 13). Retrieved December 7, 2015, from \url{https://en.wikipedia.org/wiki/Periodic_boundary_conditions}
 \bibitem{liquid_argon}
-    Rahman, A., `Correlations in the Motion of Atoms in Liquid Argon`, Physical Review, Volume 136, No. 2A. Available from: \url{http://journals.aps.org/pr/pdf/10.1103/PhysRev.136.A405}.
-\bibitem{min_image}
-    `Periodic boundary conditions` (wiki article), December 7, 2015.
-Available from: \url{https://en.wikipedia.org/wiki/Periodic_boundary_conditions#Practical_implementation:_continuity_and_the_minimum_image_convention} [November 13, 2015]
-\bibitem{equipar} 
-    `Equipartition theorem` (wiki article), December 7, 2015.
-    Available from: \url{https://en.wikipedia.org/wiki/Equipartition\_theorem} [November 21, 2015]
+    Rahman, A.
+(1964).
+Correlations in the Motion of Atoms in Liquid Argon.
+Phys.
+Rev.
+Physical Review, 126(2A).
+Retrieved December 7, 2015, from \url{http://journals.aps.org/pr/pdf/10.1103/PhysRev.136.A405}
 \bibitem{virial} 
-    `Virial theorem` (wiki article), December 7, 2015, \url{https://en.wikipedia.org/wiki/Virial_theorem} [October 7, 2015]
+    Virial Theorem. (2015, October 7). Retrieved December 7, 2015, from \url{https://en.wikipedia.org/wiki/Virial_theorem}
 \end{thebibliography}

### 2015-12-11 projects/project_5/molecular_dynamics/article/article.tex
added potential strength plot

@@
 
+    \subsubsection{Derivation}
+    \label{ssub:derivation}
+
+    The Velocity Verlet algorithm can be derived using Taylor expansion and
+    this derivation follows closely that of \cite{velo-ver}. We start with
+    $\vec{r}(t + \Delta t)$ and write this as a Taylor expansion:
+    \begin{align*}
+        \vec{r}(t + \Delta t) &= \vec{r}(t) + \Delta t \dot{\vec{r}}(t) +
+        \frac{\Delta t^2}{2}\ddot{\vec{r}}(t) + \mathcal{O}(\Delta t^3).
+    \intertext{We can rewrite this using that $\dot{\vec{v}}(t) = \vec{F}(\vec{r}(t)) / m$
+    and that $\ddot{\vec{r}} = \vec{F}(\vec{r}(t)) / m$.}
+        \vec{r}(t + \Delta t) &= \vec{r}(t) + \Delta t \vec{v}(t) + \frac{\Delta t^2}{2} \frac{\vec{F}(\vec{r}(t))}{m}+ \mathcal{O}(\Delta t^3). \\ 
+    \intertext{We now look at $\vec{v}(t + \Delta t)$ and its Taylor expansion:}
+        \vec{v}(t + \Delta t) &= \vec{t} + \Delta t \dot{\vec{v}}(t) \frac{\Delta t^2}{2}\ddot{\vec{v}}(t) + \mathcal{O}(\Delta t^3). \\
+    \intertext{In order to simplify this, we need an expression for $\ddot{\vec{v}}$. So we look at the Taylor expansion for $\dot{\vec{v}}(t + \Delta t)$:}
+        \dot{\vec{v}}(t + \Delta t) &= \dot{\vec{v}}(t) + \Delta t
+        \ddot{\vec{v}}(t) + \mathcal{O}(\Delta t^2). \\
+    \intertext{This equation can be solved for $\Delta t^2 \ddot{\vec{v}} / 2$ and when plugged into the equation above yields}
+        \vec{v}(t + \Delta t) &= \vec{v}(t) + \Delta t  \dot{\vec{v}}(t) + \frac{\Delta t}{2}\left( \dot{\vec{v}}(t + \Delta t) - \dot{\vec{v}}(t)\right) + \mathcal{O}(\Delta t^3).\\
+    \intertext{One final rewrite, using the standard equations of motion gives}
+    \vec{v}(t + \Delta t) &= \vec{v}(t) + \frac{\Delta t}{2m} \left( \vec{F}(\vec{r}(t + \Delta t)) + \vec{F}(\vec{r}(t))\right) + \mathcal{O}(\Delta t)^3.
+    \end{align*}
     \subsection{Sampling quantities}
@@
 Retrieved December 7, 2015, from \url{http://journals.aps.org/pr/pdf/10.1103/PhysRev.136.A405}
+\bibitem{velo-ver}
+    J'onsson, H. (n.d.). Classical Dynamics. Retrieved December 11, 2015, from
+    \url{https://notendur.hi.is//~hj/EE4/Verlet.pdf}
 \bibitem{virial} 

### 2015-12-11 projects/project_5/molecular_dynamics/python_src/lennard_jones.py

@@ -1,0 +1,18 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+sigma = 3.405 # angstrom
+epsil = 1 # MD units
+def lennard_jones(rij):
+
+
+    return 4 * epsil * ( (sigma / rij) ** 12 - (sigma / rij) ** 6)
+
+rij_values = np.linspace(3, 10, 10000)
+pot_values = lennard_jones(rij_values)
+
+plt.plot(rij_values / sigma, pot_values / epsil)
+plt.grid('on')
+plt.xlabel('$r_{ij} / \\sigma$')
+plt.ylabel('$U_{ij} / \\epsilon$')
+plt.savefig('lennard_jones.pdf')
